### 2024-05-09 README.Rmd
Running from the snare until fear is a habit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# traveltime

![GitHub License](https://img.shields.io/github/license/geryan/traveltime)
[![Lifecycle:](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
![GitHub commit activity](https://img.shields.io/github/commit-activity/w/geryan/traveltime)
![GitHub Downloads (all assets, all releases)](https://img.shields.io/github/downloads/geryan/traveltime/total)
![GitHub last commit](https://img.shields.io/github/last-commit/geryan/traveltime)
![GitHub R package version](https://img.shields.io/github/r-package/v/geryan/traveltime)
![GitHub commits since latest release](https://img.shields.io/github/commits-since/geryan/traveltime/latest)

## Installation

You can install the development version of traveltime from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("geryan/traveltime")
```

## Example

```{r example}
library(traveltime)
library(terra)
library(sdmtools)

friction_surface <- get_friction_surface(
    surface = "motor2020",
    file_name = temptif(),
    extent = matrix(
      data = c("111", "0", "112", 1),
      nrow = 2,
      ncol = 2,
      dimnames = list(
        c("x", "y"),
        c("min", "max")
      )
    )
  )

plot(friction_surface)
```

```{r}
from_here <- tibble::tibble(
  x = c(111.2, 111.9),
  y = c(0.2, 0.35)
)

travel_time <- calculate_travel_time(
  friction_surface = friction_surface,
  points = from_here,
  file_name = temptif()
)

plot(travel_time)
points(from_here, pch = 19, add = TRUE)
```

### 2024-05-09 R/calculate_travel_time.R
Hurry on and on and on.

@@ -5,7 +5,7 @@
 #'
 #' @param friction_surface
 #' @param points
-#' @param travel_time_filename
+#' @param file_name
 #' @param overwrite_raster
 #'
 #' @return
@@ -15,19 +15,19 @@
 calculate_travel_time <- function(
     friction_surface,
     points,
-    travel_time_filename = "outputs/travel_time.tif",
+    file_name = "outputs/travel_time.tif",
     overwrite_raster = FALSE
 ){
 
-  if(file.exists(travel_time_filename) & !overwrite_raster){
+  if(file.exists(file_name) & !overwrite_raster){
 
     warning(sprintf(
       "%s exists\nUsing existing file\nto re-generate, change overwrite_raster to TRUE %s",
-      travel_time_filename,
-      travel_time_filename
+      file_name,
+      file_name
     ))
 
-    return(terra::rast(travel_time_filename))
+    return(terra::rast(file_name))
 
   }
 
@@ -47,10 +47,10 @@ calculate_travel_time <- function(
 
   raster::writeRaster(
     travel_time,
-    travel_time_filename,
+    file_name,
     overwrite = overwrite_raster
   )
 
-  terra::rast(travel_time_filename)
+  terra::rast(file_name)
 
 }

### 2024-05-09 R/get_friction_surface.R
Hurry on and on, hurry on and on

@@ -61,10 +61,16 @@ get_friction_surface <- function(
     surface_name <- "Explorer__2020_walking_only_friction_surface"
   }
 
-  malariaAtlas::getRaster(
+  fs <- malariaAtlas::getRaster(
     dataset_id = surface_name,
     extent = extent
-  ) |>
-    sdmtools::writereadrast(filename = file_name)
+  )
+
+  names(fs) <- "friction_surface"
+
+  sdmtools::writereadrast(
+    x = fs,
+    filename = file_name
+  )
 
 }

### 2024-05-10 tests/testthat/test-calculate_travel_time.R
Run to the end of the world til you find a place

@@ -1,0 +1,40 @@
+testthat::test_that("Is SpatRaster", {
+library(traveltime)
+
+  ext <- matrix(
+    data = c("111", "0", "112", 1),
+    nrow = 2,
+    ncol = 2,
+    dimnames = list(
+      c("x", "y"),
+      c("min", "max")
+    )
+  )
+
+  wfs <- get_friction_surface(
+    surface = "walk2020",
+    extent = ext
+  )
+
+  testthat::expect_s4_class(wfs, "SpatRaster")
+
+  from_here <- data.frame(
+    x = c(111.2, 111.9),
+    y = c(0.2, 0.35)
+  )
+
+  tt1 <- calculate_travel_time(
+    friction_surface = wfs,
+    points = from_here
+  )
+
+  testthat::expect_s4_class(tt1, "SpatRaster")
+
+  tt2 <- calculate_travel_time(
+    friction_surface = wfs,
+    points = from_here,
+    file_name = tempfile(fileext = ".tif")
+  )
+
+  testthat::expect_s4_class(tt2, "SpatRaster")
+})

### 2024-05-10 tests/testthat/test-get_friction_surface.R

@@ -1,4 +1,5 @@
-test_that("Are SpatRasters", {
+testthat::test_that("Are SpatRasters", {
+  library(traveltime)
   ext <- matrix(
     data = c("111", "0", "112", 1),
     nrow = 2,
@@ -11,17 +12,17 @@ test_that("Are SpatRasters", {
 
   mfs <- get_friction_surface(
     surface = "motor2020",
-    file_name = tempfile(fileext = ".tif"),
-    extent = ext
+    extent = ext,
+    file_name = tempfile(fileext = ".tif")
   )
 
-  expect_s4_class(mfs, "SpatRaster")
+  testthat::expect_s4_class(mfs, "SpatRaster")
 
   wfs <- get_friction_surface(
     surface = "walk2020",
-    file_name = tempfile(fileext = ".tif"),
     extent = ext
   )
 
-  expect_s4_class(wfs, "SpatRaster")
+  testthat::expect_s4_class(wfs, "SpatRaster")
+
 })

### 2024-05-10 README.Rmd
Where they never never never

@@
 ![GitHub last commit](https://img.shields.io/github/last-commit/geryan/traveltime)
-![GitHub R package version](https://img.shields.io/github/r-package/v/geryan/traveltime)
 ![GitHub commits since latest release](https://img.shields.io/github/commits-since/geryan/traveltime/latest)
+
+Implements methods from Weiss et al. 2018, 2020 to calculate travel time from
+given locations over a friction surface.
+
+Citations:
+D. J. Weiss, A. Nelson, C. A. Vargas-Ruiz, K. Gligoric, S., Bavadekar, E.
+Gabrilovich, A. Bertozzi-Villa, J. Rozier, H. S. Gibson, T., Shekel, C. Kamath,
+A. Lieber, K. Schulman, Y. Shao, V. Qarkaxhija, A. K. Nandi, S. H. Keddie, S.
+Rumisha, P. Amratia, R. Arambepola, E. G. Chestnutt, J. J. Millar, T. L. Symons,
+E. Cameron, K. E. Battle, S. Bhatt, and P. W. Gething. Global maps of travel
+time to healthcare facilities. (2020) Nature Medicine.
+https://doi.org/10.1038/s41591-020-1059-1
+
+D. J. Weiss, A. Nelson, H.S. Gibson, W. Temperley, S. Peedell, A. Lieber, M.
+Hancher, E. Poyart, S. Belchior, N. Fullman, B. Mappin, U. Dalrymple, J. Rozier,
+T.C.D. Lucas, R.E. Howes, L.S. Tusting, S.Y. Kang, E. Cameron, D. Bisanzio, K.E.
+Battle, S. Bhatt, and P.W. Gething. A global map of travel time to cities to
+assess inequalities in accessibility in 2015. (2018). Nature.
+doi:10.1038/nature25181.
 
@@
 library(terra)
-library(sdmtools)
 
@@
     surface = "motor2020",
-    file_name = temptif(),
     extent = matrix(
@@
   friction_surface = friction_surface,
-  points = from_here,
-  file_name = temptif()
+  points = from_here
 )

### 2024-05-10 R/calculate_travel_time.R
No never no they′ll never lock you in

@@ -1,41 +1,94 @@
 #' @title Calculate travel time
-#' @description
-#' Implements Weiss et al method to calculate travel time given a
+#' @description Calculate the travel time from a set of points over a friction
+#' surface.
 #'
 #'
-#' @param friction_surface
-#' @param points
-#' @param file_name
-#' @param overwrite_raster
+#' @param friction_surface A `SpatRaster` friction surface layer
+#' @param points A two-column `data.frame` or `tibble` with longitude (x) in the
+#'   first column and latitude (y) in the second in the came coordinate
+#'   reference system
+#' @param file_name `character`. Output file name with extension suitable for
+#'   `terra::writeRaster`
+#' @param overwrite_raster `logical`. If `TRUE` `file_name` is overwritten.
 #'
-#' @return
+#' @details Implements methods from Weiss et al. 2018, 2020 to calculate travel
+#' time from given locations over a friction surface.
+#'
+#' Over large areas this function can require significant RAM and will be slow.
+#'
+#' Citations:
+#' D. J. Weiss, A. Nelson, C. A. Vargas-Ruiz, K. Gligoric, S., Bavadekar, E.
+#' Gabrilovich, A. Bertozzi-Villa, J. Rozier, H. S. Gibson, T., Shekel, C.
+#' Kamath, A. Lieber, K. Schulman, Y. Shao, V. Qarkaxhija, A. K. Nandi, S. H.
+#' Keddie, S. Rumisha, P. Amratia, R. Arambepola, E. G. Chestnutt, J. J. Millar,
+#' T. L. Symons, E. Cameron, K. E. Battle, S. Bhatt, and P. W. Gething. Global
+#' maps of travel time to healthcare facilities. (2020) Nature Medicine.
+#' https://doi.org/10.1038/s41591-020-1059-1
+#'
+#' D. J. Weiss, A. Nelson, H.S. Gibson, W. Temperley, S. Peedell, A. Lieber, M.
+#' Hancher, E. Poyart, S. Belchior, N. Fullman, B. Mappin, U. Dalrymple, J.
+#' Rozier, T.C.D. Lucas, R.E. Howes, L.S. Tusting, S.Y. Kang, E. Cameron, D.
+#' Bisanzio, K.E. Battle, S. Bhatt, and P.W. Gething. A global map of travel
+#' time to cities to assess inequalities in accessibility in 2015. (2018).
+#' Nature. doi:10.1038/nature25181.
+#'
[…]

### 2024-05-10 R/get_friction_surface.R
Never, no never, no never let them win

@@ -2,8 +2,8 @@
 #' @description Wrapper function to download friction surfaces via
 #' `malariaAtlas::getRaster`
 #'
-#' @param surface `"motor2020"` or `"walk2020`
-#' @param file_name Character of file name to write raster to disc
+#' @param surface `"motor2020"` or `"walk2020`.
+#' @param file_name `character`. File name for output layer.
 #' @param overwrite Overwrite `file_name` if exists
 #' @param extent Extent to pass to `malariaAtlas::getRaster`: 2x2 matrix
 #'   specifying the spatial extent within which raster data is desired, as
@@ -11,9 +11,32 @@
 #'   the maximum values; rows 1 & 2 represent the x & y dimensions respectively
 #'   (matrix(c("xmin", "ymin","xmax", "ymax"), nrow = 2, ncol = 2, dimnames =
 #'   list(c("x", "y"), c("min", "max")))) (use either shp OR extent; if neither
-#'   is specified global raster is returned).
+#'   is specified global raster is returned). `NULL` extent downloads (large)
+#'   global layer
 #'
-#' @return
+#' @details
+#'  Convenience wrapper to `malariaAtlas::getRaster` to access motorised and
+#'  walking travel friction layers per Weiss et al. 2020, that adds safety to
+#'  check existing files before download. Surfaces can be downloaded directly
+#'  from:
+#'  https://malariaatlas.org/project-resources/accessibility-to-healthcare/
+#'
+#'  `surface = "motor2020"` downloads
+#'  `"Explorer__2020_motorized_friction_surface"`.
+#'
+#'  `surface = "walk2020"` downloads
+#'  `"Explorer__2020_walking_only_friction_surface"`.
+#'
+#' D. J. Weiss, A. Nelson, C. A. Vargas-Ruiz, K. Gligoric, S., Bavadekar, E.
+#' Gabrilovich, A. Bertozzi-Villa, J. Rozier, H. S. Gibson, T., Shekel, C.
+#' Kamath, A. Lieber, K. Schulman, Y. Shao, V. Qarkaxhija, A. K. Nandi, S. H.
+#' Keddie, S. Rumisha, P. Amratia, R. Arambepola, E. G. Chestnutt, J. J. Millar,
+#' T. L. Symons, E. Cameron, K. E. Battle, S. Bhatt, and P. W. Gething. Global
+#' maps of travel time to healthcare facilities. (2020) Nature Medicine.
[…]

### 2024-05-10 README.Rmd
Run man, run while you can

@@
   )
+friction_surface
+```
 
+```{r}
 plot(friction_surface)
@@
 )
+from_here
+```
 
+```{r}
 travel_time <- calculate_travel_time(
@@
 )
+travel_time
+```
 
+```{r}
 plot(travel_time)

### 2024-05-10 README.Rmd
Run man, run man, run.

@@
 
-You can install the development version of traveltime from [GitHub](https://github.com/) with:
+You can install the development version of traveltime from
+[GitHub](https://github.com/) with:
 
@@
 
-## Example
+## Let's calculate some travel times
 
+First download a friction surface — here using the motorised travel time from
+Weiss *et al.* 2020.
 ```{r example}
@@
 
+Let's have a look at that
 ```{r}
@@
 
+Prepare points we would like to calculate travel time from
 ```{r}
@@
 
+And calculate the travel time
 ```{r}
@@
 
+Et voila!
 ```{r}

### 2024-05-10 README.Rmd
Run man, run man, run.

@@
 Citations:
-D. J. Weiss, A. Nelson, C. A. Vargas-Ruiz, K. Gligoric, S., Bavadekar, E.
+
+*D. J. Weiss, A. Nelson, C. A. Vargas-Ruiz, K. Gligoric, S., Bavadekar, E.
 Gabrilovich, A. Bertozzi-Villa, J. Rozier, H. S. Gibson, T., Shekel, C. Kamath,
@@
 Rumisha, P. Amratia, R. Arambepola, E. G. Chestnutt, J. J. Millar, T. L. Symons,
-E. Cameron, K. E. Battle, S. Bhatt, and P. W. Gething. Global maps of travel
-time to healthcare facilities. (2020) Nature Medicine.
+E. Cameron, K. E. Battle, S. Bhatt, and P. W. Gething.* **Global maps of travel
+time to healthcare facilities.** (2020) Nature Medicine.
 https://doi.org/10.1038/s41591-020-1059-1
 
-D. J. Weiss, A. Nelson, H.S. Gibson, W. Temperley, S. Peedell, A. Lieber, M.
+*D. J. Weiss, A. Nelson, H.S. Gibson, W. Temperley, S. Peedell, A. Lieber, M.
 Hancher, E. Poyart, S. Belchior, N. Fullman, B. Mappin, U. Dalrymple, J. Rozier,
 T.C.D. Lucas, R.E. Howes, L.S. Tusting, S.Y. Kang, E. Cameron, D. Bisanzio, K.E.
-Battle, S. Bhatt, and P.W. Gething. A global map of travel time to cities to
-assess inequalities in accessibility in 2015. (2018). Nature.
+Battle, S. Bhatt, and P.W. Gething.* **A global map of travel time to cities to
+assess inequalities in accessibility in 2015.** (2018). Nature.
 doi:10.1038/nature25181.

### 2024-05-10 R/ext_from_spatraster.R
I will love you from now until the end of time
resolves #3

@@ -1,0 +1,31 @@
+#' @title Extent from `SpatRaster`
+#' @description Formats spatial extent for use in `get_friction_surface`.
+#'
+#' @param r `SpatRaster`
+#'
+#' @return 2x2 `matrix`
+#' @export
+#'
+#' @examples
+#'
+#' library(terra)
+#' r <- terra::rast(
+#'     extent = terra::ext(c(111, 112, 0, 1))
+#'   )
+#'
+#' ext_from_spatraster(r)
+ext_from_spatraster <- function(r){
+
+  x <- terra::ext(r)
+
+  matrix(
+    data = c(x[1], x[3], x[2], x[4]),
+    nrow = 2,
+    ncol = 2,
+    dimnames = list(
+      c("x", "y"),
+      c("min", "max")
+    )
+  )
+
+}

### 2024-05-10 tests/testthat/test-ext_from_spatraster.R

@@ -1,0 +1,12 @@
+testthat::test_that("extent is extent", {
+
+  r <- terra::rast(
+     extent = terra::ext(c(111, 112, 0, 1))
+   )
+
+  x <- ext_from_spatraster(r)
+
+  testthat::expect_is(x, "matrix")
+
+  testthat::expect_equal(dim(x), c(2, 2))
+})

### 2024-05-10 README.Rmd
Let me guide you through every word and every ride

@@
   fig.path = "man/figures/README-",
-  out.width = "100%"
+  out.width = "100%",
+  warning = NA
 )
@@
 Weiss *et al.* 2020.
-```{r example}
+```{r }
 library(traveltime)
@@
 ```
+
+## Let's go to Singapore
+
+Here it is:
+```{r}
+library(sdmtools)
+sin <- sdmtools::make_africa_mask(
+  type = "vector",
+  countries = "SGP"
+)
+plot(sin)
+```
+Get the extent of that `SpatVector` in the format we need it.
+```{r}
+sin_ext <- ext_from_spatraster(sin)
+sin_ext
+```
+
+We're going to see how long it takes to walk home from Changi Airport. So we'll 
+download the walking-only friction surface this time.
+```{r }
+library(traveltime)
+library(terra)
+
+friction_singapore <- get_friction_surface(
+    surface = "walk2020",
+    extent = sin_ext
+  )|> 
+  mask(sin)
+
+friction_singapore
+```
+
+And where is Changi Airport?
+```{r}
+changi_airport <- tibble::tibble(
+  x = c(103.984),
+  y = c(1.355)
+)
+changi_airport
+```
+Let's look at those.
+```{r}
+plot(friction_singapore)
+plot(sin, add = TRUE)
+points(changi_airport, pch = 19, add = TRUE)
+```
+
+And calculate the travel time
+```{r}
+travel_time_sin <- calculate_travel_time(
+  friction_surface = friction_singapore,
+  points = changi_airport
+)
+travel_time_sin
+```
+
+Et voi*lah*!
+```{r}
+plot(travel_time_sin)
+points(changi_airport, pch = 19, add = TRUE)
+plot(sin, add = TRUE)
+```

### 2024-05-10 R/ext_from_terra.R
Burns for you like the moon and the sun

@@ -1,7 +1,7 @@
-#' @title Extent from `SpatRaster`
+#' @title Extent from `SpatRaster` or `SpatVector`
 #' @description Formats spatial extent for use in `get_friction_surface`.
 #'
-#' @param r `SpatRaster`
+#' @param r `terra::SpatRaster` or `terra::SpatVector`
 #'
 #' @return 2x2 `matrix`
 #' @export
@@ -13,8 +13,8 @@
 #'     extent = terra::ext(c(111, 112, 0, 1))
 #'   )
 #'
-#' ext_from_spatraster(r)
-ext_from_spatraster <- function(r){
+#' ext_from_terra(r)
+ext_from_terra <- function(r){
 
   x <- terra::ext(r)
 

### 2024-05-10 tests/testthat/test-ext_from_terra.R

@@ -4,7 +4,7 @@ testthat::test_that("extent is extent", {
      extent = terra::ext(c(111, 112, 0, 1))
    )
 
-  x <- ext_from_spatraster(r)
+  x <- ext_from_terra(r)
 
   testthat::expect_is(x, "matrix")
 

### 2024-05-15 README.Rmd
Even endless
Burns for you like the moon and the sun
add GH actions R CMD Check + code coverage

@@
 
+<!-- badges: start -->
+[![R-CMD-check](https://github.com/idem-lab/traveltime/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/idem-lab/traveltime/actions/workflows/R-CMD-check.yaml)
+[![Codecov test coverage](https://codecov.io/gh/idem-lab/traveltime/branch/main/graph/badge.svg)](https://app.codecov.io/gh/idem-lab/traveltime?branch=main)
 ![GitHub License](https://img.shields.io/github/license/geryan/traveltime)
@@
 ![GitHub commits since latest release](https://img.shields.io/github/commits-since/geryan/traveltime/latest)
+<!-- badges: end -->
 
@@
 # install.packages("remotes")
-remotes::install_github("geryan/traveltime")
+remotes::install_github("idem-lab/traveltime")
 ```
@@
 ```{r}
-sin_ext <- ext_from_spatraster(sin)
+sin_ext <- ext_from_terra(sin)
 sin_ext

### 2024-05-16 R/get_friction_surface.R
remove `sdmtools` import from pkg, still in readme maybe fixes #4?

@@ -93,10 +93,13 @@ get_friction_surface <- function(
   names(fs) <- "friction_surface"
 
   if(!is.null(file_name)){
-    sdmtools::writereadrast(
+    terra::writeRaster(
       x = fs,
       filename = file_name
     )
+
+    fs <- terra::rast(file_name)
+
   } else{
     fs
   }

### 2024-05-21 R/ext_from_terra.R
use `ext_vect_to_matrix` under bonnet

@@ -18,14 +18,6 @@ ext_from_terra <- function(r){
 
   x <- terra::ext(r)
 
-  matrix(
-    data = c(x[1], x[3], x[2], x[4]),
-    nrow = 2,
-    ncol = 2,
-    dimnames = list(
-      c("x", "y"),
-      c("min", "max")
-    )
-  )
+  ext_vect_to_matrix(x)
 
 }

### 2024-05-21 R/get_friction_surface.R
simplify `extent` input by allowing `vector` `SpatExtent` or `SpatRaster` extents as well as matrix resolve #9

@@ -1,31 +1,25 @@
 #' @title Get friction surface
 #' @description Wrapper function to download friction surfaces via
-#' `malariaAtlas::getRaster`
+#'   `malariaAtlas::getRaster`
 #'
 #' @param surface `"motor2020"` or `"walk2020`.
 #' @param file_name `character`. File name for output layer.
 #' @param overwrite Overwrite `file_name` if exists
-#' @param extent Extent to pass to `malariaAtlas::getRaster`: 2x2 matrix
-#'   specifying the spatial extent within which raster data is desired, as
-#'   returned by sf::st_bbox() - the first column has the minimum, the second
-#'   the maximum values; rows 1 & 2 represent the x & y dimensions respectively
-#'   (matrix(c("xmin", "ymin","xmax", "ymax"), nrow = 2, ncol = 2, dimnames =
-#'   list(c("x", "y"), c("min", "max")))) (use either shp OR extent; if neither
-#'   is specified global raster is returned). `NULL` extent downloads (large)
-#'   global layer
+#' @param extent Spatial extent as either numeric vector specifying `c(xmin,
+#'   xmax, ymin, ymax)`, `SpatExtent`, `SpatRaster` (from which the extent will
+#'   be taken), or 2x2 matrix (see details).
 #'
-#' @details
-#'  Convenience wrapper to `malariaAtlas::getRaster` to access motorised and
-#'  walking travel friction layers per Weiss et al. 2020, that adds safety to
-#'  check existing files before download. Surfaces can be downloaded directly
-#'  from:
-#'  https://malariaatlas.org/project-resources/accessibility-to-healthcare/
+#' @details Convenience wrapper to `malariaAtlas::getRaster` to access motorised
+#' and walking travel friction layers per Weiss et al. 2020, that adds safety to
+#' check existing files before download. Surfaces can be downloaded directly
+#' from:
+#' \url{https://malariaatlas.org/project-resources/accessibility-to-healthcare/}
 #'
-#'  `surface = "motor2020"` downloads
-#'  `"Explorer__2020_motorized_friction_surface"`.
+#' `surface = "motor2020"` downloads
+#' `"Explorer__2020_motorized_friction_surface"`.
 #'
-#'  `surface = "walk2020"` downloads
[…]

### 2024-05-21 README.Rmd
Simplify extent in README

@@
     surface = "motor2020",
-    extent = matrix(
-      data = c("111", "0", "112", 1),
-      nrow = 2,
-      ncol = 2,
-      dimnames = list(
-        c("x", "y"),
-        c("min", "max")
-      )
-    )
+    extent = c(111,112,0,1)
   )
@@
 ```
-Get the extent of that `SpatVector` in the format we need it.
-```{r}
-sin_ext <- ext_from_terra(sin)
-sin_ext
-```
 
@@
 download the walking-only friction surface this time.
+
+We can feed in our `sin` `SpatVector` directly as the `extent`, instead of specifying by hand as above. 
 ```{r }
@@
     surface = "walk2020",
-    extent = sin_ext
+    extent = sin
   )|> 

### 2024-05-21 README.Rmd
fix md issue

@@
 <!-- badges: start -->
+[![traveltime status badge](https://idem-lab.r-universe.dev/badges/traveltime)](https://idem-lab.r-universe.dev/traveltime)
 [![R-CMD-check](https://github.com/idem-lab/traveltime/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/idem-lab/traveltime/actions/workflows/R-CMD-check.yaml)

### 2024-05-28 R/get_friction_surface.R
add overwrite into writeRaster as it should be resolves #14

@@ -124,7 +124,8 @@ get_friction_surface <- function(
   if(!is.null(file_name)){
     terra::writeRaster(
       x = fs,
-      filename = file_name
+      filename = file_name,
+      overwrite = overwrite
     )
 
     fs <- terra::rast(file_name)

### 2024-05-28 R/calculate_travel_time.R
replce `file_name` with `filename`
breaks old code using this param

resolve #16

@@ -7,9 +7,9 @@
 #' @param points A two-column `data.frame` or `tibble` with longitude (x) in the
 #'   first column and latitude (y) in the second in the came coordinate
 #'   reference system
-#' @param file_name `character`. Output file name with extension suitable for
+#' @param filename `character`. Output file name with extension suitable for
 #'   `terra::writeRaster`
-#' @param overwrite_raster `logical`. If `TRUE` `file_name` is overwritten.
+#' @param overwrite_raster `logical`. If `TRUE` `filename` is overwritten.
 #'
 #' @details Implements methods from Weiss et al. 2018, 2020 to calculate travel
 #' time from given locations over a friction surface.
@@ -66,20 +66,20 @@
 calculate_travel_time <- function(
     friction_surface,
     points,
-    file_name = NULL,
+    filename = NULL,
     overwrite_raster = FALSE
 ){
 
-  if(!is.null(file_name)){
-    if(file.exists(file_name) & !overwrite_raster){
+  if(!is.null(filename)){
+    if(file.exists(filename) & !overwrite_raster){
 
       warning(sprintf(
         "%s exists\nUsing existing file\nto re-generate, change overwrite_raster to TRUE %s",
-        file_name,
-        file_name
+        filename,
+        filename
       ))
 
-      return(terra::rast(file_name))
+      return(terra::rast(filename))
 
     }
   }
@@ -99,14 +99,14 @@ calculate_travel_time <- function(
   travel_time <- gdistance::accCost(tgc, xy.matrix)
   names(travel_time) <- "travel_time"
 
-  if(!is.null(file_name)){
+  if(!is.null(filename)){
     raster::writeRaster(
       travel_time,
-      file_name,
+      filename,
       overwrite = overwrite_raster
     )
 
-    return(terra::rast(file_name))
+    return(terra::rast(filename))
   } else {
     terra::rast(travel_time)
   }

### 2024-05-28 R/get_friction_surface.R

@@ -3,8 +3,8 @@
 #'   `malariaAtlas::getRaster`
 #'
 #' @param surface `"motor2020"` or `"walk2020`.
-#' @param file_name `character`. File name for output layer.
-#' @param overwrite Overwrite `file_name` if exists
+#' @param filename `character`. File name for output layer.
+#' @param overwrite Overwrite `filename` if exists
 #' @param extent Spatial extent as either numeric vector specifying `c(xmin,
 #'   xmax, ymin, ymax)`, `SpatExtent`, `SpatVector` or `SpatRaster` (from which
 #'   the extent will be taken), or 2x2 `matrix` (see details).
@@ -64,22 +64,22 @@
 #'
 get_friction_surface <- function(
   surface = c("motor2020", "walk2020"),
-  file_name = NULL,
+  filename = NULL,
   overwrite = FALSE,
   extent = NULL
 ){
   surface <- match.arg(surface)
 
-  if(!is.null(file_name)){
-    if(!overwrite & file.exists(file_name)){
+  if(!is.null(filename)){
+    if(!overwrite & file.exists(filename)){
 
       warning(sprintf(
         "%s exists\n Returning %s. To replace, change overwrite_raster to TRUE",
-        file_name,
-        file_name
+        filename,
+        filename
       ))
 
-      return(terra::rast(file_name))
+      return(terra::rast(filename))
     }
   }
 
@@ -121,14 +121,14 @@ get_friction_surface <- function(
 
   names(fs) <- "friction_surface"
 
-  if(!is.null(file_name)){
+  if(!is.null(filename)){
     terra::writeRaster(
       x = fs,
-      filename = file_name,
+      filename = filename,
       overwrite = overwrite
     )
 
-    fs <- terra::rast(file_name)
+    fs <- terra::rast(filename)
 
   } else{
     fs

### 2024-05-28 tests/testthat/test-calculate_travel_time.R

@@ -32,7 +32,7 @@ test_that("Is SpatRaster", {
   tt2 <- calculate_travel_time(
     friction_surface = wfs,
     points = from_here,
-    file_name = tempfile(fileext = ".tif")
+    filename = tempfile(fileext = ".tif")
   )
 
   expect_s4_class(tt2, "SpatRaster")

### 2024-05-28 tests/testthat/test-get_friction_surface.R

@@ -12,7 +12,7 @@ test_that("Are SpatRasters", {
   mfs <- get_friction_surface(
     surface = "motor2020",
     extent = ext,
-    file_name = tempfile(fileext = ".tif")
+    filename = tempfile(fileext = ".tif")
   )
 
   expect_s4_class(mfs, "SpatRaster")

### 2024-05-28 R/calculate_travel_time.R
potentially breaking old code: replace `overwrite_raster` with `overwrite` resolve #17

@@ -9,7 +9,7 @@
 #'   reference system
 #' @param filename `character`. Output file name with extension suitable for
 #'   `terra::writeRaster`
-#' @param overwrite_raster `logical`. If `TRUE` `filename` is overwritten.
+#' @param overwrite `logical`. If `TRUE` `filename` is overwritten.
 #'
 #' @details Implements methods from Weiss et al. 2018, 2020 to calculate travel
 #' time from given locations over a friction surface.
@@ -67,14 +67,14 @@ calculate_travel_time <- function(
     friction_surface,
     points,
     filename = NULL,
-    overwrite_raster = FALSE
+    overwrite = FALSE
 ){
 
   if(!is.null(filename)){
-    if(file.exists(filename) & !overwrite_raster){
+    if(file.exists(filename) & !overwrite){
 
       warning(sprintf(
-        "%s exists\nUsing existing file\nto re-generate, change overwrite_raster to TRUE %s",
+        "%s exists\nUsing existing file\nto re-generate, change overwrite to TRUE %s",
         filename,
         filename
       ))
@@ -103,7 +103,7 @@ calculate_travel_time <- function(
     raster::writeRaster(
       travel_time,
       filename,
-      overwrite = overwrite_raster
+      overwrite = overwrite
     )
 
     return(terra::rast(filename))

### 2024-05-28 R/get_friction_surface.R

@@ -74,7 +74,7 @@ get_friction_surface <- function(
     if(!overwrite & file.exists(filename)){
 
       warning(sprintf(
-        "%s exists\n Returning %s. To replace, change overwrite_raster to TRUE",
+        "%s exists\n Returning %s. To replace, change overwrite to TRUE",
         filename,
         filename
       ))

### 2024-05-28 README.Rmd
suggest install via r-universe
add nicer plot and resolve #21

@@
 
-You can install the development version of traveltime from
-[GitHub](https://github.com/) with:
+You can install `traveltime` with:
 
 ``` r
-# install.packages("remotes")
-remotes::install_github("idem-lab/traveltime")
+install.packages("traveltime", repos = c("https://idem-lab.r-universe.dev"))
 ```
@@
 plot(travel_time)
-points(from_here, pch = 19, add = TRUE)
+points(from_here, pch = 19)
 ```
@@
 plot(sin, add = TRUE)
-points(changi_airport, pch = 19, add = TRUE)
+points(changi_airport, pch = 19)
 ```
@@
 plot(travel_time_sin)
-points(changi_airport, pch = 19, add = TRUE)
+points(changi_airport, pch = 19)
 plot(sin, add = TRUE)
 ```
+
+### I want my plots to be nicer
+```{r}
+library(tidyterra)
+library(ggplot2)
+
+ggplot() +
+  # plot the spatraster first
+  geom_spatraster(
+    data = travel_time_sin
+  ) +
+  theme_minimal() +
+  scale_fill_whitebox_c(palette = "deep") +
+  # overlay the vector outline
+  geom_spatvector(
+    data = sin,
+    colour = "khaki3",
+    linewidth = 1,
+    fill = NA
+  ) +
+  # add the points from tibble
+   geom_point(
+     data = changi_airport,
+     aes(x = x, y = y)
+   ) +
+  labs(x = NULL, y = NULL, fill = "Travel time\n(minutes)")
+
+```

### 2024-07-11 README.Rmd
Pour André

@@
     data = sin,
-    colour = "khaki3",
+    colour = "grey70",
     linewidth = 1,
@@
      data = changi_airport,
-     aes(x = x, y = y)
+     aes(x = x, y = y),
+     colour = "hotpink"
    ) +

### 2024-08-16 README.Rmd
I saw the best minds of my generation destroyed by madness

@@
 ```{r}
+# install.packages("sdmtools", repos = "https://idem-lab.r-universe.dev")
 library(sdmtools)

### 2024-12-10 data-raw/stations.R
add stations data

@@ -1,0 +1,14 @@
+## code to prepare `DATASET` dataset goes here
+
+library(terra)
+
+# Land Transport Authority. (2019). LTA MRT Station Exit (GEOJSON) (2024)
+# [Dataset]. data.gov.sg. Retrieved December 10, 2024 from
+# https://data.gov.sg/datasets/d_b39d3a0871985372d7e1637193335da5/view
+
+mrt_vect <- vect("data-raw/LTAMRTStationExitGEOJSON.geojson")
+
+
+stations <-geom(mrt_vect)[,c("x", "y")]
+
+usethis::use_data(stations, overwrite = TRUE)

### 2024-12-10 README.Rmd
update readme to use stations

--- 2024-05-28 README.Rmd
@@
 ```{r}
-library(sdmtools)
-sin <- sdmtools::make_africa_mask(
-  type = "vector",
-  countries = "SGP"
+#install.packages("geodata")
+library(geodata)
+sin <- gadm(
+  country = "Singapore",
+  level = 0,
+  path = tempdir(),
+  resolution = 2
 )
@@
 
-We're going to see how long it takes to walk home from Changi Airport. So we'll 
-download the walking-only friction surface this time.
+We're going to see how long it takes to walk home from an MRT or LRT station. So
+we'll download the walking-only friction surface this time.
 
-We can feed in our `sin` `SpatVector` directly as the `extent`, instead of specifying by hand as above. 
+We can feed in our `sin` `SpatVector` directly as the `extent`, instead of
+specifying by hand as above. 
 ```{r }
@@
 
-And where is Changi Airport?
+And where are the stations? The the `stations` data set contains the latitude
+and longitude of all LRT and MRT station exits [^1].
 ```{r}
-changi_airport <- tibble::tibble(
-  x = c(103.984),
-  y = c(1.355)
-)
-changi_airport
+head(stations)
+plot(stations)
 ```
-Let's look at those.
-```{r}
-plot(friction_singapore)
-plot(sin, add = TRUE)
-points(changi_airport, pch = 19)
-```
-
-And calculate the travel time
-```{r}
-travel_time_sin <- calculate_travel_time(
-  friction_surface = friction_singapore,
-  points = changi_airport
-)
-travel_time_sin
-```
-
-Et voi*lah*!
-```{r}
-plot(travel_time_sin)
-points(changi_airport, pch = 19)
-plot(sin, add = TRUE)
-```
-
-### I want my plots to be nicer
+Let's look at those. Below we plot the friction surface raster, with the vector
+boundary as a dashed grey line, and the station exits as grey points.
 ```{r}
@@
 ggplot() +
-  # plot the spatraster first
   geom_spatraster(
-    data = travel_time_sin
+    data = friction_singapore
   ) +
-  theme_minimal() +
-  scale_fill_whitebox_c(palette = "deep") +
-  # overlay the vector outline
   geom_spatvector(
     data = sin,
-    colour = "khaki3",
-    linewidth = 1,
-    fill = NA
+    fill = "transparent",
+    col = "grey50",
+    lty = 2
   ) +
-  # add the points from tibble
   geom_point(
-     data = changi_airport,
-     aes(x = x, y = y)
+    data = stations,
+    aes(
+      x = x,
+      y = y
+    ),
+    col = "grey60",
+    size = 0.5
   ) +
-  labs(x = NULL, y = NULL, fill = "Travel time\n(minutes)")
+  scale_fill_viridis_c(
+    option = "A",
+    na.value = "transparent",
+    direction = -1
+  ) +
+  labs(
+    fill = "Friction"
+  )
 
 ```
+
+OK, now we want to calculate the walking travel time in minutes across the 
+friction surface from the nearest station exit:
+```{r}
+travel_time_sin <- calculate_travel_time(
+  friction_surface = friction_singapore,
+  points = stations
+)
+travel_time_sin
+```
+
+Et voi*lah* --- a raster of walking time from the nearest station.
+```{r}
+contour(
+  x = travel_time_sin,
+  filled = TRUE,
+  nlevels = 20,
+  col = viridis::magma(19, direction = -1)
+)
+```
+
+[^1]: Land Transport Authority. (2019). LTA MRT Station Exit (GEOJSON) (2024)
+[Dataset]. data.gov.sg. Retrieved December 10, 2024 from
+https://data.gov.sg/datasets/d_b39d3a0871985372d7e1637193335da5/view

### 2024-12-10 R/stations.R
document the `stations` data

@@ -1,0 +1,13 @@
+#' Singapore MRT and LRT data
+#'
+#' Locations of all MRT and LRT station exits in Singapore.
+#'
+#' @format ## `stations` A data frame with 563 rows and 2 columns:
+#' \describe{
+#'   \item{x}{longitude}
+#'   \item{y}{latitude}
+#' }
+#' @source Land Transport Authority. (2019). LTA MRT Station Exit (GEOJSON)
+#'   (2024) Dataset. data.gov.sg. Retrieved December 10, 2024 from
+#'   <https://data.gov.sg/datasets/d_b39d3a0871985372d7e1637193335da5/view>
+"stations"

### 2024-12-10 data-raw/stations.R

@@ -9,6 +9,7 @@ library(terra)
 mrt_vect <- vect("data-raw/LTAMRTStationExitGEOJSON.geojson")
 
 
-stations <-geom(mrt_vect)[,c("x", "y")]
+stations <- geom(mrt_vect)[,c("x", "y")]
+
 
 usethis::use_data(stations, overwrite = TRUE)

### 2024-12-10 README.Rmd
update readme example

@@
 
-First download a friction surface — here using the motorised travel time from
-Weiss *et al.* 2020.
+First download a friction surface --— here we are using the motorised travel
+time from Weiss *et al.* 2020. We use the function
+`traveltime::get_friction_surface`, specify the surface (type) as `"motor2020"`,
+ and provide the spatial extent of interest:
 ```{r }
@@
 
-Let's have a look at that
+Let's have a look at that `SpatRaster`:
 ```{r}
@@
 
-Prepare points we would like to calculate travel time from
+Now, prepare points that we would like to calculate travel time from:
 ```{r}
@@
 
-And calculate the travel time
+And calculate the travel time from our points `from_here` over the friction 
+surface `friction_surface` using the function
+`traveltime::calculate_travel_time`:
 ```{r}
@@
 
-Et voila!
+Et voila! Here is the motorised travel time in minutes for each cell, with our
+points in pink.
 ```{r}
 plot(travel_time)
-points(from_here, pch = 19)
+points(from_here, pch = 19, col = "hotpink")
 ```
 
-## Let's go to Singapore
+## A more tangible example: Walking in Singapore
 
-Here it is:
+Let's create a map of the walking time across the island of Singapore from the nearest
+[MRT or LRT](https://en.wikipedia.org/wiki/Transport_in_Singapore#Rail_transport)
+station.
+
+To do this, we need:
+
+- a map of Singapore
+- locations of the stations
+
+Here's our basemap via `geodata`:
 ```{r}
@@
 
-We're going to see how long it takes to walk home from an MRT or LRT station. So
-we'll download the walking-only friction surface this time.
+We're going to see how long it takes to walk home from a station, so we'll
+download the walking-only friction surface this time by specifying 
+`surface = "walk2020`.
 
-We can feed in our `sin` `SpatVector` directly as the `extent`, instead of
-specifying by hand as above. 
+We can also pass in our basemap `sin`, a `SpatVector`, directly as the `extent`,
+instead of specifying by hand as above. We're also only interested in walking 
+*on land* so we mask out areas outside of `sin`, that are within the extent of 
+the raster:
+
 ```{r }
@@
 
-And where are the stations? The the `stations` data set contains the latitude
-and longitude of all LRT and MRT station exits [^1].
+The the `stations` data set in this package contains the longitude and latitude
+of all LRT and MRT station exits in Singapore[^1].
+
 ```{r}
 head(stations)
-plot(stations)
 ```
-Let's look at those. Below we plot the friction surface raster, with the vector
-boundary as a dashed grey line, and the station exits as grey points.
+
+Let's look at our data now.
+
+Below we plot the friction surface raster `friction_singapore`, with the vector
+boundary of `sin` as a dashed grey line, and `stations` as grey points:
+
 ```{r}
@@
 friction surface from the nearest station exit:
+
 ```{r}

### 2024-12-10 README.Rmd

@@
 
-Let's create a map of the walking time across the island of Singapore from the nearest
+Let's create a map of the walking time across the island of Singapore from the 
+nearest
 [MRT or LRT](https://en.wikipedia.org/wiki/Transport_in_Singapore#Rail_transport)
@@
 Here's our basemap via `geodata`:
+
 ```{r}

### 2024-12-12 R/calculate_travel_time.R
check classes and accept SpatVector for points resolve #25

@@ -1,37 +1,37 @@
 #' @title Calculate travel time
 #' @description Calculate the travel time from a set of points over a friction
-#' surface.
+#'   surface.
 #'
 #'
 #' @param friction_surface A `SpatRaster` friction surface layer. See
-#'  `?get_friction_surface`
-#' @param points A two-column `data.frame` or `tibble` with longitude (x) in the
-#'   first column and latitude (y) in the second in the same coordinate
-#'   reference system as `friction_surface`
+#'   `?get_friction_surface`
+#' @param points A two-column `matrix`, `data.frame`, or `tibble` with longitude
+#'   (x) in the first column and latitude (y) in the second, or a `SpatVector`,
+#'   in the same coordinate, reference system as `friction_surface`.
 #' @param filename `character`. Output file name with extension suitable for
 #'   `terra::writeRaster`
 #' @param overwrite `logical`. If `TRUE` `filename` is overwritten.
 #'
 #' @details Implements methods from Weiss et al. 2018, 2020 to calculate travel
-#' time from given locations over a friction surface.
+#'   time from given locations over a friction surface.
 #'
-#' Over large areas this function can require significant RAM and will be slow.
+#'   Over large areas this function can require significant RAM and will be
+#'   slow.
 #'
-#' Citations:
-#' D. J. Weiss, A. Nelson, C. A. Vargas-Ruiz, K. Gligoric, S., Bavadekar, E.
-#' Gabrilovich, A. Bertozzi-Villa, J. Rozier, H. S. Gibson, T., Shekel, C.
-#' Kamath, A. Lieber, K. Schulman, Y. Shao, V. Qarkaxhija, A. K. Nandi, S. H.
-#' Keddie, S. Rumisha, P. Amratia, R. Arambepola, E. G. Chestnutt, J. J. Millar,
-#' T. L. Symons, E. Cameron, K. E. Battle, S. Bhatt, and P. W. Gething. Global
-#' maps of travel time to healthcare facilities. (2020) Nature Medicine.
-#' https://doi.org/10.1038/s41591-020-1059-1
+#'   Citations: D. J. Weiss, A. Nelson, C. A. Vargas-Ruiz, K. Gligoric, S.,
[…]

### 2024-12-12 R/calculate_travel_time.R
simplify code - remove extraneous steps

@@ -105,10 +105,12 @@ calculate_travel_time <- function(
   tsn <- gdistance::transition(friction, function(x) 1/mean(x), 8)
   tgc <- gdistance::geoCorrection(tsn)
 
-  xy.data.frame <- data.frame()
-  xy.data.frame[1:npoints,1] <- points[,1]
-  xy.data.frame[1:npoints,2] <- points[,2]
-  xy.matrix <- as.matrix(xy.data.frame)
+  # xy.data.frame <- data.frame()
+  # xy.data.frame[1:npoints,1] <- points[,1]
+  # xy.data.frame[1:npoints,2] <- points[,2]
+  # xy.matrix <- as.matrix(xy.data.frame)
+
+  xy.matrix <- as.matrix(points[,1:2])
 
   travel_time <- gdistance::accCost(tgc, xy.matrix)
   names(travel_time) <- "travel_time"

### 2024-12-18 R/calculate_travel_time.R
test aut

@@ -75,7 +75,10 @@ calculate_travel_time <- function(
     if(file.exists(filename) & !overwrite){
 
       warning(sprintf(
-        "%s exists\nUsing existing file\nto re-generate, change overwrite to TRUE %s",
+        "%s exists\n
+        Using existing file\n
+        to re-generate,
+        change overwrite to TRUE %s",
         filename,
         filename
       ))

### 2024-12-18 tests/testthat/test-ext_from_terra.R

@@ -6,7 +6,9 @@ test_that("extent is extent", {
 
   x <- ext_from_terra(r)
 
-  expect_is(x, "matrix")
+  #expect_is(x, "matrix")
+  expect_type(x, "double")
+  #expect_s3_class(x, "matrix")
 
   expect_equal(dim(x), c(2, 2))
 })

### 2024-12-18 R/calculate_travel_time.R
small tweaks regarding cli/explaining variables

@@ -71,35 +71,40 @@ calculate_travel_time <- function(
     overwrite = FALSE
 ){
 
-  if(!is.null(filename)){
-    if(file.exists(filename) & !overwrite){
+  # wrap into a function
+  # warn_if_filename_already_used(filename, overwrite)
+  filename_used <- !is.null(filename)
+  filename_exists_and_no_overwrite <- file.exists(filename) && !overwrite
+  warn_user_not_overwrite <- filename_used && filename_exists_and_no_overwrite
+  if (warn_user_not_overwrite) {
 
-      warning(sprintf(
-        "%s exists\n
-        Using existing file\n
-        to re-generate,
-        change overwrite to TRUE %s",
-        filename,
-        filename
-      ))
+    cli::cli_warn(
+      message = c(
+        "x" = "{.path {filename}} already exists",
+        "Using existing file, {.path {filename}}",
+        "i" =  "To re-generate file, change {.arg overwrite} to {.code TRUE}"
+      )
+    )
 
-      return(terra::rast(filename))
+    return(terra::rast(filename))
 
-    }
   }
 
-  if(!"SpatRaster" %in% class(friction_surface)){
-    stop("friction_surface must be a SpatRaster class object")
+  if (!inherits(friction_surface, "SpatRaster")){
+    cli::cli_abort(
+      "{.arg friction_surface} must be a {.cls SpatRaster}."
+    )
   }
 
+  # explaining variable
   if(!any(c("matrix", "data.frame", "SpatVector") %in% class(points))){
     stop("points must be a SpatVector, data.frame, or matrix")
   }
 
+  # use inherits
   if("SpatVector" %in% class(points)){
     points <- terra::geom(points)[,c("x", "y")]
   }
-
 
   #npoints <- nrow(points)
 

### 2024-12-18 R/ext_matrix.R
various code review:
* add cli
* use S3 methods for new function, ext_matrix
* update tests to use inherits

@@ -1,0 +1,76 @@
+#' @title Convert extent to matrix
+#'
+#' @description
+#' This function converts from formats ...
+#' TODO - explain why we need to do this, which is that we need a special esoteric format
+#'
+#' @param x `numeric` length 4, consisting of `c(xmin, xmax, ymin, ymax)`
+#' dimensions of extent
+#'
+#' @return 2x2 `matrix` - explain esoteric format of matrix
+#'
+#' @examples
+#' ext_matrix(c(111,112,0, 1))
+#' # TODO
+#' # add examples of all SpatRaster, SpatVector, etc.
+#' @export
+ext_matrix <- function(x, ...){
+  UseMethod("ext_matrix")
+}
+
+#' @export
+ext_matrix.SpatExtent <- function(x, ...){
+  extent <- ext_vect_to_matrix(x)
+  extent
+}
+
+#' @export
+ext_matrix.SpatRaster <- function(x, ...){
+  extent <- ext_from_terra(x)
+  extent
+}
+
+#' @export
+ext_matrix.SpatVector <- function(x, ...){
+  extent <- ext_from_terra(x)
+  extent
+}
+
+#' @export
+ext_matrix.vector <- function(x){
+  if (length(extent) != 4){
+    cli::cli_abort(
+      message = c(
+        "{.arg extent} as numeric must be length 4",
+        "We see {.arg extent} has having length: {.val {length(extent)}}."
+        )
+    )
+  }
+  extent <- ext_vect_to_matrix(extent)
+  extent
+}
+
+#' @export
+ext_matrix.matrix <- function(x, ...){
+  is_2x2 <- identical(dim(x), c(2L,2L))
+  if(!is_2x2){
+    cli::cli_abort(
+      message = c(
+        "If {.arg x} is of class, {.cls matrix}, it must have dimensions: 2x2",
+        "However, we see that {.arg x} has dimensions: \\
+        {.val {paste0(dim(x), collapse = 'x')}}."
+        )
+    )
+  }
+}
+
+#' @export
+ext_matrix.default <- function(x, ...){
+  cli::cli_abort(
+    message = c(
+      "{.arg extent} must be of class {.cls numeric, matrix, \\
+      SpatExtent, SpatRaster}",
+      "But we see class: {.cls {class(x)}."
+      )
+  )
+}

### 2024-12-18 tests/testthat/test-ext_vect_to_matrix.R

@@ -1,8 +1,8 @@
 test_that("ext vect to matrix", {
   x <- ext_vect_to_matrix(c(111,112,0, 1))
 
-  expect_is(x, "matrix")
-
+  expect_true(inherits(x, "matrix"))
+  expect_type(x, "double")
   expect_equal(dim(x), c(2, 2))
 
 })

### 2024-12-18 R/get_friction_surface.R
change MAP friction surface pointers - Explorer server surfaces deprecated

@@ -84,9 +84,9 @@ get_friction_surface <- function(
   }
 
   if (surface == "motor2020") {
-    surface_name <- "Explorer__2020_motorized_friction_surface"
+    surface_name <- "Accessibility__202001_Global_Motorized_Friction_Surface"
   } else if (surface == "walk2020"){
-    surface_name <- "Explorer__2020_walking_only_friction_surface"
+    surface_name <- "Accessibility__202001_Global_Walking_Only_Friction_Surface"
   }
 
   if (is.null(extent)) {

### 2024-12-19 R/ext_matrix.R
correct bug in ext_matrix

@@ -46,7 +46,7 @@ ext_matrix.vector <- function(x){
         )
     )
   }
-  extent <- ext_vect_to_matrix(extent)
+  extent <- ext_vect_to_matrix(x)
   extent
 }
 
@@ -62,6 +62,8 @@ ext_matrix.matrix <- function(x, ...){
         )
     )
   }
+  extent <- x
+  extent
 }
 
 #' @export

### 2024-12-19 R/calculate_travel_time.R
fix error with `file.exists(NULL)`

@@ -74,8 +74,15 @@ calculate_travel_time <- function(
   # wrap into a function
   # warn_if_filename_already_used(filename, overwrite)
   filename_used <- !is.null(filename)
-  filename_exists_and_no_overwrite <- file.exists(filename) && !overwrite
-  warn_user_not_overwrite <- filename_used && filename_exists_and_no_overwrite
+
+  file_exists <- ifelse(
+    filename_used,
+    file.exists(filename),
+    FALSE
+  )
+
+  warn_user_not_overwrite <- filename_used && file_exists && !overwrite
+
   if (warn_user_not_overwrite) {
 
     cli::cli_warn(

### 2024-12-19 R/get_friction_surface.R
use nice cli warning

@@ -68,17 +68,28 @@ get_friction_surface <- function(
   surface <- match.arg(surface)
 
   # similar code from calculate_travel_time
-  if(!is.null(filename)){
-    if(!overwrite & file.exists(filename)){
+  filename_used <- !is.null(filename)
 
-      warning(sprintf(
-        "%s exists\n Returning %s. To replace, change overwrite to TRUE",
-        filename,
-        filename
-      ))
+  file_exists <- ifelse(
+    filename_used,
+    file.exists(filename),
+    FALSE
+  )
 
-      return(terra::rast(filename))
-    }
+  warn_user_not_overwrite <- filename_used && file_exists && !overwrite
+
+  if (warn_user_not_overwrite) {
+
+    cli::cli_warn(
+      message = c(
+        "x" = "{.path {filename}} already exists",
+        "Using existing file, {.path {filename}}",
+        "i" =  "To re-generate file, change {.arg overwrite} to {.code TRUE}"
+      )
+    )
+
+    return(terra::rast(filename))
+
   }
 
 

### 2024-12-19 R/ext_matrix.R
fix s3 issues

@@ -1,43 +1,41 @@
 #' @title Convert extent to matrix
 #'
 #' @description
-#' This function converts from formats ...
-#' TODO - explain why we need to do this, which is that we need a special esoteric format
+#' This function allows `get_friction_surface()` to accept the argument `extent` as a `vector`, 2x2 `matrix`, `SpatVector`, or `SpatRaster` and extracts and converts into the annoyingly specific format necessary to download the friction surface. See details in `?get_friction_surface`
 #'
-#' @param x `numeric` length 4, consisting of `c(xmin, xmax, ymin, ymax)`
-#' dimensions of extent
+#' @param extent `vector` of length 4, 2x2 `matrix`, `SpatVector`, or `SpatRaster`
 #'
-#' @return 2x2 `matrix` - explain esoteric format of matrix
+#' @return 2x2 `matrix` with column names "x" and "y" and row names "min"
+#'  and "max
 #'
 #' @examples
 #' ext_matrix(c(111,112,0, 1))
-#' # TODO
-#' # add examples of all SpatRaster, SpatVector, etc.
+#'
 #' @export
-ext_matrix <- function(x, ...){
+ext_matrix <- function(extent){
   UseMethod("ext_matrix")
 }
 
 #' @export
-ext_matrix.SpatExtent <- function(x, ...){
-  extent <- ext_vect_to_matrix(x)
+ext_matrix.SpatExtent <- function(extent){
+  extent <- ext_vect_to_matrix(extent)
   extent
 }
 
 #' @export
-ext_matrix.SpatRaster <- function(x, ...){
-  extent <- ext_from_terra(x)
+ext_matrix.SpatRaster <- function(extent){
+  extent <- ext_from_terra(extent)
   extent
 }
 
 #' @export
-ext_matrix.SpatVector <- function(x, ...){
-  extent <- ext_from_terra(x)
+ext_matrix.SpatVector <- function(extent){
+  extent <- ext_from_terra(extent)
   extent
 }
 
 #' @export
-ext_matrix.vector <- function(x){
+ext_matrix.double <- function(extent){
   if (length(extent) != 4){
     cli::cli_abort(
       message = c(
@@ -46,33 +44,32 @@ ext_matrix.vector <- function(x){
         )
     )
   }
-  extent <- ext_vect_to_matrix(x)
+  extent <- ext_vect_to_matrix(extent)
   extent
 }
 
 #' @export
-ext_matrix.matrix <- function(x, ...){
[…]

### 2024-12-19 README.Rmd
update codecov thing

@@
 [![R-CMD-check](https://github.com/idem-lab/traveltime/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/idem-lab/traveltime/actions/workflows/R-CMD-check.yaml)
-[![Codecov test coverage](https://codecov.io/gh/idem-lab/traveltime/branch/main/graph/badge.svg)](https://app.codecov.io/gh/idem-lab/traveltime?branch=main)
 ![GitHub License](https://img.shields.io/github/license/geryan/traveltime)
@@
 ![GitHub commits since latest release](https://img.shields.io/github/commits-since/geryan/traveltime/latest)
+[![Codecov test coverage](https://codecov.io/gh/idem-lab/traveltime/graph/badge.svg)](https://app.codecov.io/gh/idem-lab/traveltime)
 <!-- badges: end -->
@@
 time to healthcare facilities.** (2020) Nature Medicine.
-https://doi.org/10.1038/s41591-020-1059-1
+<https://doi.org/10.1038/s41591-020-1059-1>
 
@@
 assess inequalities in accessibility in 2015.** (2018). Nature.
-doi:10.1038/nature25181.
+<https://doi.org/10.1038/nature25181>
 

### 2024-12-20 R/calculate_travel_time.R
create `warn_and_return()` utility and replace complex warning checking code in main functions

@@ -71,31 +71,10 @@ calculate_travel_time <- function(
     overwrite = FALSE
 ){
 
-  # wrap into a function
-  # warn_if_filename_already_used(filename, overwrite)
-  filename_used <- !is.null(filename)
-
-  file_exists <- ifelse(
-    filename_used,
-    file.exists(filename),
-    FALSE
+  warn_and_return(
+    filename = filename,
+    overwrite = overwrite
   )
-
-  warn_user_not_overwrite <- filename_used && file_exists && !overwrite
-
-  if (warn_user_not_overwrite) {
-
-    cli::cli_warn(
-      message = c(
-        "x" = "{.path {filename}} already exists",
-        "Using existing file, {.path {filename}}",
-        "i" =  "To re-generate file, change {.arg overwrite} to {.code TRUE}"
-      )
-    )
-
-    return(terra::rast(filename))
-
-  }
 
   if (!inherits(friction_surface, "SpatRaster")){
     cli::cli_abort(

### 2024-12-20 R/get_friction_surface.R

@@ -67,31 +67,10 @@ get_friction_surface <- function(
 ){
   surface <- match.arg(surface)
 
-  # similar code from calculate_travel_time
-  filename_used <- !is.null(filename)
-
-  file_exists <- ifelse(
-    filename_used,
-    file.exists(filename),
-    FALSE
+  warn_and_return(
+    filename = filename,
+    overwrite = overwrite
   )
-
-  warn_user_not_overwrite <- filename_used && file_exists && !overwrite
-
-  if (warn_user_not_overwrite) {
-
-    cli::cli_warn(
-      message = c(
-        "x" = "{.path {filename}} already exists",
-        "Using existing file, {.path {filename}}",
-        "i" =  "To re-generate file, change {.arg overwrite} to {.code TRUE}"
-      )
-    )
-
-    return(terra::rast(filename))
-
-  }
-
 
   surface_name <- switch(
     surface,

### 2024-12-20 R/utils.R

@@ -1,0 +1,35 @@
+
+# warn_and_return checks if a file exists already and returns the raster
+# from the exsting file with a warning that no calculation is being performed
+warn_and_return <- function(
+  filename,
+  overwrite
+){
+
+  filename_used <- !is.null(filename)
+
+  file_exists <- ifelse(
+    filename_used,
+    file.exists(filename),
+    FALSE
+  )
+
+  warn_user_not_overwrite <- filename_used && file_exists && !overwrite
+
+  if (warn_user_not_overwrite) {
+
+    cli::cli_warn(
+      message = c(
+        "x" = "{.path {filename}} already exists",
+        "Using existing file, {.path {filename}}",
+        "i" =  "To re-generate file, change {.arg overwrite} to {.code TRUE}"
+      )
+    )
+
+    return(terra::rast(filename))
+
+  }
+
+  TRUE
+
+}

### 2025-01-22 paper/paper.qmd
add draft paper

---
title: "`traveltime`: an R package to calculate walking and motorised travel time from user locations"
format:
  md: default
  html: 
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
tags: 
  - R
  - geographic information systems
  - spatial analysis
authors:
  - name: Gerard E Ryan
    orcid: 0000-0000-0000-0000
    corresponding: true
    affiliation: "1, 2"
  - name: Nick Tierney
    orcid: 1234
    affiliation: "1, 3"
  - name: Nick Golding
    orcid: 1234
    affiliation: "1, 3 "
  - name: Daniel Weiss
    orcid: 1234
    affiliation: "1, 3"
affiliations:
 - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
   index: 1
 - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
   index: 2
 - name: Curtin University, Bentley, WA, Australia
   index: 3
date: In prep 2025
bibliography: paper.bib
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Summary

Understanding and mapping the time to travel among locations is useful for many activities from urban planning `[@zahavi1974traveltime]` to public health `[@weiss2020global]` and likely a myriad others we haven't thought of. Here we present a software package --- `traveltime` --- written in and for the language R `[@Rref]`. `traveltime` enables a user to supply a set of geographic point locations and an extent of interest, and calculate the motorised or walking travel time over that extent. The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest.

The work-flow requires two steps:

-   downloading a 'friction surface' for the area of interest, and then
-   calculating travel time over that surface for the points of interest.

`traveltime` provides a spatial interface using object classes from the `terra` package `[@terra]`. 
It accepts points as either `data.frame` or `SpatVector` class objects; extents as `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` objects; and return the results as a `SpatRaster` object. The package uses `gdistance` `[@gdistance2017]` to calculate the minimum least-cost-distance for each cell from the points of interest.The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by `@weiss2020global` available via the `malariaAtlas` package `[@pfeffer2018malariaatlas]`. Rather than house and manipulate these surfaces independently, `traveltime` relies on `malariaAtlas` to download the surfaces for the extent of interest.

```{r ttexample,  echo=FALSE}
#| label: "fig-tt_healthcare"
#| fig-cap: "Motorised travel time from healthcare facilities in Cambodia"
library(terra)
library(malariaAtlas)
library(tidyterra)

boundary_kh <- getShp(ISO = "KHM")

tt_kh <- getRaster(
  dataset_id = "Accessibility__202001_Global_Motorized_Travel_Time_to_Healthcare",
  shp = boundary_kh
)

ggplot() +
  geom_spatraster(
    data = tt_kh
  ) +
  scale_fill_viridis_c(
    option = "A",
    direction = -1,
    na.value = "transparent"
  ) +
  theme_minimal() +
  labs(fill = "minutes")
```

Although this article is intended to be the key reference for the `traveltime` package, any citations of it should also be citing the underlying work by `@weiss2018global; @weiss2020global`.

# Statement of need

Global maps of travel time to cities `[@weiss2018global]` and health care facilities `[@weiss2020global]` have generated significant interest and use. Collectively `@weiss2018global ; @weiss2020global` have over 1300 citations (Google Scholar, at 10th December 2024), and the city data set is available to R users through the widely-used `geodata` package `[@geodata]`. There is clear demand for these type of products.

`@weiss2020global` and colleagues made their code available as an R script to allow for reproduction and extension of their analyses (<https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt> ). To further enable extension of this work, here we have developed an R package based on that code to seamlessly calculate the travel time from any arbitrary set of locations.

Other R packages produce superficially similar though not identical functionality. At least three R packages provide interfaces to the <https://www.TravelTime.com> API `[@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]`. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location. The latter is most comparable to what the `traveltime` package achieves,

THIS NEXT SENTENCE SUX - EXPLAIN WHY ITS THE SAME AND GIVE A BETTER EXPLANATION OF THE DIFFERENCE.

however without complete spatial coverage, or distance to the nearest of a given point. Furthermore, it requires access keys, caps queries, and uses an external server for calculations, which add friction to the use experience.

With `traveltime`, we provide internally complete software to estimate motorised or walking travel time from any number of user-supplied locations, across a complete area of interest.

# Example workflow: walking in Singapore

In this example workflow, we wish to calculate the walking travel time to the nearest mass transit station across the island nation of Singapore --- specifically MRT and LRT stations --- and create a map of this.

To do this we will need two items of information: - our are of interest - in this case, a map of Singapore - the locations of MRT and LRT stations

We can download a national-level polygon of Singapore from the GADM `[@gadm]` database using the `geodata` package `[@geodata]`. Here we download only the national boundary (`level = 0`) and at a low resolution (`resolution = 2`).

```{r get basemap}
#| label: "fig-basemap"
#| fig-cap: "Base map of Singapore"
library(terra)
library(geodata)
library(tidyterra)

sin <- gadm(
  country = "Singapore",
  level = 0,
  path = tempdir(),
  resolution = 2
)

ggplot() +
  geom_spatvector(
    data = sin
  ) +
  # scale_fill_viridis_c(
  #   option = "A",
  #   direction = -1,
  #   na.value = "transparent"
  # ) +
  scale_fill_manual(na.value = "transparent") +
  theme_minimal()

```

The the `stations` data set included in the `traveltime` package is a 563 \$\*\$ 2 matrix containing the longitude (x) and latitude (y) of all LRT and MRT station exits in Singapore from `@singdata:2024`.

```{r show stations}
library(traveltime)
head(stations)
```

So now we have the two items of data we require. The next step is to download a friction surface for our area of interest.

We can pass in our basemap `sin`, a `SpatVector`, directly as the `extent`. It would also be possible pass in a `SpatRaster` with extent equal to our area of interest, or specify coordinates either as a vector, matrix, or `SpatExtent`.

We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`. Alternatively, we could use `"motor2020"` for motorised travel if that was of interest.

We're also only interested in walking *on land* so we mask out areas outside of `sin`, that are within the extent of the raster:

```{r  download friction surface}
library(traveltime)

friction_singapore <- get_friction_surface(
    surface = "walk2020",
    extent = sin
  )|> 
  mask(sin)

friction_singapore
```

Taking a look at these data, below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points. Higher values of friction indicate more time travelling across a given cell.

```{r plot data}
#| label: "fig-data"
#| fig-cap: "Friction surface raster of Singapore, showing boundary from @fig-basemap and station locations as grey points"

library(tidyterra)
library(ggplot2)

ggplot() +
  geom_spatraster(
    data = friction_singapore
  ) +
  geom_spatvector(
    data = sin,
    fill = "transparent",
    col = "grey50"
  ) +
  geom_point(
    data = stations,
    aes(
      x = x,
      y = y
    ),
    col = "grey60",
    size = 0.5
  ) +
  scale_fill_viridis_c(
    option = "A",
    na.value = "transparent",
    direction = -1
  ) +
  labs(
    fill = "Friction",
    x = element_blank(),
    y = element_blank()
  ) +
  theme_minimal()

```

With all the data collected, the function `calculate_travel_time` takes the friction surface `friction_singapore` and the locations of interest `stations` and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station.

```{r calculate result}
travel_time_sin <- calculate_travel_time(
  friction_surface = friction_singapore,
  points = stations
)
travel_time_sin
```

Here are those results as in a map:

```{r plot result}
#| label: "fig-result"
#| fig-cap: "Map of walking travel time in Singapore, in minutes from nearest MRT or LRT station."

ggplot() +
  geom_spatraster(
    data = travel_time_sin
  ) +
  scale_fill_viridis_c(
    option = "A",
    direction = -1,
    na.value = "transparent"
  ) +
  theme_minimal() +
  labs(fill = "minutes") +
  geom_spatvector(
    data = sin,
    fill = "transparent",
    col = "grey20"
  )
```

Note that the above raster include infinite (`Inf`) values. In @fig-data, a number of islands to the south are shown as cells unconnected with the mainland. These raster cells for these islands appear absent in @fig-result. Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in the plot @fig-result.

# Opportunities for future development

The `traveltime` package is immediately suitable to a range of applications where travel to custom locations of interest. But we nonetheless see opportunities to build the package utility into the future through two mechanisms: capability to distribute more friction surfaces, and additional methods to handle large spatial extents.

Firstly, `traveltime` currently has access to friction surfaces for 2020, both at 30 arc-second resolution. As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available. Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses. For instance, although the example here was chosen for it's simplicity and low computational demands, a 5 km cell size is a relatively large area to walk across, and thus actual waking times are likley to vary significantly within each cell.

At the other end of the scale, the package can require large amounts of memory for analyses over large landscapes (e.g. one analyses over Africa required \~ 72 GB RAM). Developing methods to handle large landscapes with less memory would be helpful to make such analyses accessible to those without access to larger computing resources.

# Acknowledgements

Ta luv

# References

### 2025-01-23 paper/paper.qmd
redirect yml to paper.html.md
- this keeps the author meta data (orcid, affil), whereas render direct to md with default loses it.

@@
 format:
-  md: default
+  #md: default
   html:
@@
   - name: Nick Tierney
-    orcid: 1234
+    orcid: 0000-0003-1460-8722
     affiliation: "1, 3"
   - name: Nick Golding
-    orcid: 1234
+    orcid: 0000-0001-8916-5570
     affiliation: "1, 3 "
   - name: Daniel Weiss
-    orcid: 1234
+    orcid: 0000-0000-0000-0000
     affiliation: "1, 3"
@@
 
-Ta luv
-
 # References

### 2025-01-24 paper/paper.qmd
added missing ORCIDs to paper resolve #34 resolve #35 that was easy der brain

--- 2025-01-22 paper/paper.qmd
@@
   - name: Gerard E Ryan
-    orcid: 0000-0000-0000-0000
+    orcid: 0000-0003-0183-7630
     corresponding: true
@@
   - name: Nick Tierney
-    orcid: 1234
+    orcid: 0000-0003-1460-8722
     affiliation: "1, 3"
   - name: Nick Golding
-    orcid: 1234
+    orcid: 0000-0001-8916-5570
     affiliation: "1, 3 "
   - name: Daniel Weiss
-    orcid: 1234
+    orcid: 0000-0002-6175-5648
     affiliation: "1, 3"

### 2025-01-24 R/ext_matrix.R
complete examples and tests for `ext_matrix` S3 method resolves #32

@@ -3,13 +3,27 @@
 #' @description
 #' This function allows `get_friction_surface()` to accept the argument `extent` as a `vector`, 2x2 `matrix`, `SpatVector`, or `SpatRaster` and extracts and converts into the annoyingly specific format necessary to download the friction surface. See details in `?get_friction_surface`
 #'
-#' @param extent `vector` of length 4, 2x2 `matrix`, `SpatVector`, or `SpatRaster`
+#' @param extent `vector` of length 4, 2x2 `matrix`, `SpatExtent`, `SpatVector`, or `SpatRaster`
 #'
 #' @return 2x2 `matrix` with column names "x" and "y" and row names "min"
 #'  and "max
 #'
 #' @examples
-#' ext_matrix(c(111,112,0, 1))
+#' # vector/double
+#' x <- c(111,112,0, 1)
+#' ext_matrix(x)
+#'
+#' # SpatExtent
+#' y <-  terra::ext(x)
+#' ext_matrix(y)
+#'
+#' # SpatRaster
+#' r <- terra::rast(extent = y)
+#' ext_matrix(r)
+#'
+#' # SpatVector
+#' v <- terra::vect(y)
+#' ext_matrix(v)
 #'
 #' @export
 ext_matrix <- function(extent){

### 2025-01-24 tests/testthat/test-ext_from_terra.R

@@ -10,4 +10,14 @@ test_that("extent is extent", {
   expect_true(inherits(x, "matrix"))
   expect_equal(dim(x), c(2, 2))
 
+  v <- terra::vect(
+    x = terra::ext(c(111, 112, 0, 1))
+  )
+
+  y <- ext_matrix(v)
+
+  expect_type(y, "double")
+  expect_true(inherits(y, "matrix"))
+  expect_equal(dim(y), c(2, 2))
+
 })

### 2025-01-24 R/get_friction_surface.R
tidy documentation around extents for `get_friction_surface` resolve #19

@@ -5,54 +5,53 @@
 #' @param surface `"motor2020"` or `"walk2020`.
 #' @param filename `character`. File name for output layer.
 #' @param overwrite Overwrite `filename` if exists
-#' @param extent Spatial extent as either numeric vector specifying `c(xmin,
-#'   xmax, ymin, ymax)`, `SpatExtent`, `SpatVector` or `SpatRaster` (from which
-#'   the extent will be taken), or 2x2 `matrix` (see details).
+#' @param extent Spatial extent as one of:
+#'  - a numeric vector specifying `c(xmin,
+#'   xmax, ymin, ymax)`,
+#'   - a `terra` `SpatExtent`, `SpatVector`, or `SpatRaster` from which
+#'   the extent will be taken, or
+#'   -  or 2x2 `matrix`
 #'
-#' @details Convenience wrapper to [malariaAtlas::getRaster()] to access motorised
-#' and walking travel friction layers per Weiss et al. 2020, that adds safety to
-#' check existing files before download. Surfaces can be downloaded directly
-#' from:
-#' \url{https://malariaatlas.org/project-resources/accessibility-to-healthcare/}
+#'   See details
 #'
-#' `surface = "motor2020"` downloads
-#' `"Explorer__2020_motorized_friction_surface"`.
+#' @details Convenience wrapper to [malariaAtlas::getRaster()] to access
+#'   motorised and walking travel friction layers per Weiss et al. 2020, that
+#'   adds safety to check existing files before download. Surfaces can be
+#'   downloaded directly from:
+#'   \url{https://malariaatlas.org/project-resources/accessibility-to-healthcare/}.
 #'
-#' `surface = "walk2020"` downloads
-#' `"Explorer__2020_walking_only_friction_surface"`.
+#'   Here when
 #'
-#' D. J. Weiss, A. Nelson, C. A. Vargas-Ruiz, K. Gligoric, S., Bavadekar, E.
-#' Gabrilovich, A. Bertozzi-Villa, J. Rozier, H. S. Gibson, T., Shekel, C.
-#' Kamath, A. Lieber, K. Schulman, Y. Shao, V. Qarkaxhija, A. K. Nandi, S. H.
-#' Keddie, S. Rumisha, P. Amratia, R. Arambepola, E. G. Chestnutt, J. J. Millar,
[…]

### 2025-01-24 R/get_friction_surface.R
note re warning about faulty geoserver "it's not our fault blame MAP/AWS"

@@ -20,7 +20,7 @@
 #'   downloaded directly from:
 #'   \url{https://malariaatlas.org/project-resources/accessibility-to-healthcare/}.
 #'
-#'   Here when
+#'   Here ---
 #'
 #'  - `surface = "motor2020"` will download
 #'   `"Explorer__2020_motorized_friction_surface"`, and
@@ -43,6 +43,11 @@
 #'   `matrix(c("xmin", "ymin", "xmax", "ymax"), nrow = 2, ncol = 2, dimnames =
 #'   list(c("x", "y"), c("min", "max")))`. `NULL` extent downloads (large)
 #'   global layer.
+#'
+#'   Troubleshooting: if you get a warning `Failed to connect to MAP geoserver`,
+#'   this is an issue with fetching the friction surface using
+#'   `malariaAtlas::getRaster`, and not an issue with `traveltime` itself. The
+#'   server may be down or there may be an issue with that package.
 #'
 #' @return `SpatRaster`
 #' @export

### 2025-01-24 paper/paper.qmd
several things: - headings resolving #40 - backticks attempting to address #33 - splitting chunks with better chunk options to resolve #38

@@
 
-```{r ttexample,  echo=FALSE}
-#| label: "fig-tt_healthcare"
-#| fig-cap: "Motorised travel time from healthcare facilities in Cambodia"
+```{r ttexample,  include=FALSE}
 library(terra)
@@
 )
+```
+
+```{r ttexample_image, echo=FALSE}
+#| label: "fig-tt_healthcare"
+#| fig-cap: "Existing surface of motorised travel time from healthcare facilities, exerpted to Cambodia (Weiss et al., 2020)"
 
@@
 
-Although this article is intended to be the key reference for the `traveltime` package, any citations of it should also be citing the underlying work by `@weiss2018global; @weiss2020global`.
+Reference documents for `traveltime` are available at <https://idem-lab.github.io/traveltime/>.
+Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work by `@weiss2018global; @weiss2020global`.
 
@@
 
-Global maps of travel time to cities `[@weiss2018global]` and health care facilities `[@weiss2020global]` have generated significant interest and use.
+Global maps of travel time to cities `[@weiss2018global]` and health care facilities `[@weiss2020global]` `@fig-tt_healthcare` have generated significant interest and use.
 Collectively `@weiss2018global ; @weiss2020global` have over 1300 citations (Google Scholar, at 10th December 2024), and the city data set is available to R users through the widely-used `geodata` package `[@geodata]`.
@@
 
-Other R packages produce superficially similar though not identical functionality.
-At least three R packages provide interfaces to the <https://www.TravelTime.com> API `[@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]`.
+Other R packages produce superficially similar though fundamentally different functionality.
+A gaggle of R packages provide interfaces to the <https://www.TravelTime.com> API `[@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]`.
 The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location.
-The latter is most comparable to what the `traveltime` package achieves,
-
-THIS NEXT SENTENCE SUX - EXPLAIN WHY ITS THE SAME AND GIVE A BETTER EXPLANATION OF THE DIFFERENCE.
-
-however without complete spatial coverage, or distance to the nearest of a given point.
-Furthermore, it requires access keys, caps queries, and uses an external server for calculations, which add friction to the use experience.
+The isochron polygons are most comparable to what the `traveltime` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across cells.
+TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repetitive iteration for all combinations of time and points, and additional calculations.
+Furthermore, TravelTime.com requires access keys, paid subscription beyond a short free period, caps queries, which add considerable friction to the use experience.
 
@@
 
-# Example workflow: walking in Singapore
+# Example workflow: walking home from public transport in Singapore
 
@@
 
-To do this we will need two items of information: - our are of interest - in this case, a map of Singapore - the locations of MRT and LRT stations
+## Prepare the data and download friction surface
+
+For this excercise, we need two items of data:
+
+-   our area of interest; in this case, a map of Singapore, and
+-   our points to calcute travel time from, that is, the locations of Singapore's MRT and LRT stations.
 
@@
 #| label: "fig-basemap"
-#| fig-cap: "Base map of Singapore"
+#| fig-cap: "Area of interest: Singapore"
 library(terra)
@@
   theme_minimal()
-
 ```
@@
 
-We can pass in our basemap `sin`, a `SpatVector`, directly as the `extent`.
+We can pass in our basemap `sin` `@fig-basemap`, a `SpatVector`, directly as the `extent`.
 It would also be possible pass in a `SpatRaster` with extent equal to our area of interest, or specify coordinates either as a vector, matrix, or `SpatExtent`.
@@
 
-```{r  download friction surface}
+```{r  download friction surface, echo=TRUE, output=FALSE}
 library(traveltime)
@@
   mask(sin)
+```
 
+```{r show friction surface raster }
 friction_singapore
@@
 
-Taking a look at these data, below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points.
+Taking a look at these data, below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points `@fig-data`.
 Higher values of friction indicate more time travelling across a given cell.
@@
 #| label: "fig-data"
-#| fig-cap: "Friction surface raster of Singapore, showing boundary from @fig-basemap and station locations as grey points"
+#| fig-cap: "Friction surface raster of Singapore, showing Singapore boundary in grey, and station locations as grey points"
 
@@
 
+## Calculate and plot the travel time
+
 With all the data collected, the function `calculate_travel_time` takes the friction surface `friction_singapore` and the locations of interest `stations` and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station.
@@
 
-Here are those results as in a map:
+We present those results in `@fig-result`:
 
@@
 Note that the above raster include infinite (`Inf`) values.
-In @fig-data, a number of islands to the south are shown as cells unconnected with the mainland.
-These raster cells for these islands appear absent in @fig-result.
-Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in the plot @fig-result.
+In `@fig-data`, a number of islands to the south are shown as cells unconnected with the mainland.
+These raster cells for these islands appear absent in `@fig-result`.
+Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in the plot `@fig-result`.
 

### 2025-01-24 paper/paper.qmd
fix citation format so referenes work resolve #42

@@
 ---
-title: "`traveltime`: an R package to calculate walking and motorised travel time from user locations"
+title: "`traveltime`: an R package to calculate walking and motorised travel time from user-specified locations"
 format:
@@
 
-Understanding and mapping the time to travel among locations is useful for many activities from urban planning `[@zahavi1974traveltime]` to public health `[@weiss2020global]` and likely a myriad others we haven't thought of.
-Here we present a software package --- `traveltime` --- written in and for the language R `[@Rref]`.
+Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@weiss2020global] and likely a myriad others we haven't thought of.
+Here we present a software package --- `traveltime` --- written in and for the language R [@Rref].
 `traveltime` enables a user to supply a set of geographic point locations and an extent of interest, and calculate the motorised or walking travel time over that extent.
@@
 It accepts points as either `data.frame` or `SpatVector` class objects; extents as `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` objects; and return the results as a `SpatRaster` object.
-The package uses `gdistance` `[@gdistance2017]` to calculate the minimum least-cost-distance for each cell from the points of interest.The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by `@weiss2020global` available via the `malariaAtlas` package `[@pfeffer2018malariaatlas]`.
+The package uses `gdistance` [@gdistance2017] to calculate the minimum least-cost-distance for each cell from the points of interest.
+The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by @weiss2020global available via the `malariaAtlas` package [@pfeffer2018malariaatlas].
 Rather than house and manipulate these surfaces independently, `traveltime` relies on `malariaAtlas` to download the surfaces for the extent of interest.
@@
 Reference documents for `traveltime` are available at <https://idem-lab.github.io/traveltime/>.
-Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work by `@weiss2018global; @weiss2020global`.
+Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work by @weiss2018global; @weiss2020global.
 
@@
 
-Global maps of travel time to cities `[@weiss2018global]` and health care facilities `[@weiss2020global]` `@fig-tt_healthcare` have generated significant interest and use.
-Collectively `@weiss2018global ; @weiss2020global` have over 1300 citations (Google Scholar, at 10th December 2024), and the city data set is available to R users through the widely-used `geodata` package `[@geodata]`.
+Global maps of travel time to cities [@weiss2018global] and health care facilities [@weiss2020global] `@fig-tt_healthcare` have generated significant interest and use.
+Collectively @weiss2018global ; @weiss2020global have over 1300 citations (Google Scholar, at 10th December 2024), and the city data set is available to R users through the widely-used `geodata` package [@geodata].
 There is clear demand for these type of products.
 
-`@weiss2020global` and colleagues made their code available as an R script to allow for reproduction and extension of their analyses (<https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt> ).
+@weiss2020global and colleagues made their code available as an R script to allow for reproduction and extension of their analyses (<https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt> ).
 To further enable extension of this work, here we have developed an R package based on that code to seamlessly calculate the travel time from any arbitrary set of locations.
@@
 Other R packages produce superficially similar though fundamentally different functionality.
-A gaggle of R packages provide interfaces to the <https://www.TravelTime.com> API `[@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]`.
+A gaggle of R packages provide interfaces to the <https://www.TravelTime.com> API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh].
 The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location.
@@
 
-We can download a national-level polygon of Singapore from the GADM `[@gadm]` database using the `geodata` package `[@geodata]`.
+We can download a national-level polygon of Singapore from the GADM [@gadm] database using the `geodata` package [@geodata].
 Here we download only the national boundary (`level = 0`) and at a low resolution (`resolution = 2`).
@@
 
-The the `stations` data set included in the `traveltime` package is a 563 \$\*\$ 2 matrix containing the longitude (x) and latitude (y) of all LRT and MRT station exits in Singapore from `@singdata:2024`.
+The the `stations` data set included in the `traveltime` package is a 563 \$\*\$ 2 matrix containing the longitude (x) and latitude (y) of all LRT and MRT station exits in Singapore from @singdata:2024.
 
@@
 
-We can pass in our basemap `sin` `@fig-basemap`, a `SpatVector`, directly as the `extent`.
+We can pass in our basemap `sin` @fig-basemap, a `SpatVector`, directly as the `extent`.
 It would also be possible pass in a `SpatRaster` with extent equal to our area of interest, or specify coordinates either as a vector, matrix, or `SpatExtent`.
@@
 
+GR is supported by the Gates Foundation grant for the Vector Atlas.
+
 # References

### 2025-01-24 paper/paper.qmd
fixes cross refs - resolve #43 - did I create this issue recently when it wasn't previously a problem? :shrug:

@@
 
-Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@weiss2020global] and likely a myriad others we haven't thought of.
+Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@weiss2020global] (@ref-tt_healthcare) and likely a myriad others we haven't thought of.
 Here we present a software package --- `traveltime` --- written in and for the language R [@Rref].
@@
 
-Reference documents for `traveltime` are available at <https://idem-lab.github.io/traveltime/>.
+`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and reference documents at <https://idem-lab.github.io/traveltime/>.
 Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work by @weiss2018global; @weiss2020global.
@@
 
-Global maps of travel time to cities [@weiss2018global] and health care facilities [@weiss2020global] `@fig-tt_healthcare` have generated significant interest and use.
+Global maps of travel time to cities [@weiss2018global] and health care facilities [@weiss2020global] (@fig-tt_healthcare) have generated significant interest and use.
 Collectively @weiss2018global ; @weiss2020global have over 1300 citations (Google Scholar, at 10th December 2024), and the city data set is available to R users through the widely-used `geodata` package [@geodata].
@@
 
-We can pass in our basemap `sin` @fig-basemap, a `SpatVector`, directly as the `extent`.
+We can pass in our basemap `sin` (@fig-basemap), a `SpatVector`, directly as the `extent`.
 It would also be possible pass in a `SpatRaster` with extent equal to our area of interest, or specify coordinates either as a vector, matrix, or `SpatExtent`.
@@
 
-Taking a look at these data, below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points `@fig-data`.
+Taking a look at these data, below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points @fig-data.
 Higher values of friction indicate more time travelling across a given cell.
@@
 
-We present those results in `@fig-result`:
+We present those results in @fig-result:
 
@@
 Note that the above raster include infinite (`Inf`) values.
-In `@fig-data`, a number of islands to the south are shown as cells unconnected with the mainland.
-These raster cells for these islands appear absent in `@fig-result`.
-Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in the plot `@fig-result`.
+In @fig-data, a number of islands to the south are shown as cells unconnected with the mainland.
+These raster cells for these islands appear absent in @fig-result.
+Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in the plot @fig-result.
 

### 2025-01-24 paper/paper.qmd
minor fixes: - fix faulty example extent in `get_friction_surface` - typos in `ext_matrix` - tidy chunks in paper

@@
 library(malariaAtlas)
-library(tidyterra)
 
@@
 #| fig-cap: "Existing surface of motorised travel time from healthcare facilities, exerpted to Cambodia (Weiss et al., 2020)"
+library(ggplot2)
+library(tidyterra)
 

### 2025-01-24 R/ext_matrix.R

@@ -6,11 +6,11 @@
 #' @param extent `vector` of length 4, 2x2 `matrix`, `SpatExtent`, `SpatVector`, or `SpatRaster`
 #'
 #' @return 2x2 `matrix` with column names "x" and "y" and row names "min"
-#'  and "max
+#'  and "max"
 #'
 #' @examples
 #' # vector/double
-#' x <- c(111,112,0, 1)
+#' x <- c(111, 112, 0, 1)
 #' ext_matrix(x)
 #'
 #' # SpatExtent

### 2025-01-24 R/get_friction_surface.R

@@ -56,7 +56,7 @@
 #'
 #' # for more examples of passing exten types see ?ext_matrix
 #'
-#' ext <- c(111, 0, 112, 1)
+#' ext <- c(111, 112, 0, 1)
 #'
 #'  get_friction_surface(
 #'    surface = "motor2020",

### 2025-01-28 paper/paper.qmd
edits to tidy paper

@@
 ---
-title: "`traveltime`: an R package to calculate walking and motorised travel time from user-specified locations"
+title: "`{traveltime}`: an R package to calculate walking and motorised travel time from user-specified locations"
 format:
@@
 
-Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@weiss2020global] (@ref-tt_healthcare) and likely a myriad others we haven't thought of.
-Here we present a software package --- `traveltime` --- written in and for the language R [@Rref].
-`traveltime` enables a user to supply a set of geographic point locations and an extent of interest, and calculate the motorised or walking travel time over that extent.
-The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest.
+Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@weiss2020global] (@ref-tt_healthcare) and likely a myriad others.
+Here we present a software package --- `{traveltime}` --- written in and for the language R [@Rref].
+`{traveltime}` enables a user to supply a set of geographic point locations and an extent of interest, to then calculate the motorised or walking travel time over that extent.
+The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest (e.g., travel time from health care facilities @ref-tt_healthcare from @weiss2020global).
 
@@
 
-`traveltime` provides a spatial interface using object classes from the `terra` package `[@terra]`.
-It accepts points as either `data.frame` or `SpatVector` class objects; extents as `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` objects; and return the results as a `SpatRaster` object.
+`{traveltime}` provides a spatial interface using object classes from the `terra` package `[@terra]`.
+It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and extents as `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` objects; and returns the result as a `SpatRaster` object.
 The package uses `gdistance` [@gdistance2017] to calculate the minimum least-cost-distance for each cell from the points of interest.
 The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by @weiss2020global available via the `malariaAtlas` package [@pfeffer2018malariaatlas].
-Rather than house and manipulate these surfaces independently, `traveltime` relies on `malariaAtlas` to download the surfaces for the extent of interest.
+Rather than house and manipulate these surfaces independently, `{traveltime}` relies on `malariaAtlas` to download the surfaces for the extent of interest.
 
@@
 
-`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and reference documents at <https://idem-lab.github.io/traveltime/>.
-Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work by @weiss2018global; @weiss2020global.
+`{traveltime}` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and has reference documents at <https://idem-lab.github.io/traveltime/>.
+Although this article is intended to be the key reference for the `{traveltime}` package, we suggest citations of the package should be accompanied by citing the underlying work by @weiss2018global; @weiss2020global as well.
 
@@
 The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location.
-The isochron polygons are most comparable to what the `traveltime` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across cells.
-TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repetitive iteration for all combinations of time and points, and additional calculations.
-Furthermore, TravelTime.com requires access keys, paid subscription beyond a short free period, caps queries, which add considerable friction to the use experience.
+The isochron polygons are most comparable to what the `{traveltime}` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across cells.
+TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repetitive iteration for all combinations of time and points, and additional calculations of the minimum value for each cell from all points.
+Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use.
 
-With `traveltime`, we provide internally complete software to estimate motorised or walking travel time from any number of user-supplied locations, across a complete area of interest.
+With `{traveltime}`, we provide free and open source software to estimate motorised or walking travel time from any number of user-supplied locations, across a complete area of interest.
 
-# Example workflow: walking home from public transport in Singapore
+# Example: walking from public transport in Singapore
 
-In this example workflow, we wish to calculate the walking travel time to the nearest mass transit station across the island nation of Singapore --- specifically MRT and LRT stations --- and create a map of this.
+In this example we wish to calculate the walking travel time from the nearest mass transit station across the island nation of Singapore --- specifically MRT and LRT stations --- and create a map of this.
 
@@
 
--   our area of interest; in this case, a map of Singapore, and
--   our points to calcute travel time from, that is, the locations of Singapore's MRT and LRT stations.
+-   our extent of interest --- in this case a map of Singapore, and
+-   our points to calculate travel time from --- here the locations of Singapore's MRT and LRT stations.
 
@@
 Here we download only the national boundary (`level = 0`) and at a low resolution (`resolution = 2`).
+Our boundary `sin` is a `SpatVector` class object.
 
 ```{r get basemap}
-#| label: "fig-basemap"
-#| fig-cap: "Area of interest: Singapore"
-library(terra)
 library(geodata)
-library(tidyterra)
 
@@
 
+sin
+```
+
+```{r plot basemap}
+#| label: "fig-basemap"
+#| fig-cap: "Area of interest: Singapore"
+
 ggplot() +
@@
   ) +
-  # scale_fill_viridis_c(
-  #   option = "A",
-  #   direction = -1,
-  #   na.value = "transparent"
-  # ) +
   scale_fill_manual(na.value = "transparent") +
@@
 
-The the `stations` data set included in the `traveltime` package is a 563 \$\*\$ 2 matrix containing the longitude (x) and latitude (y) of all LRT and MRT station exits in Singapore from @singdata:2024.
+The the `stations` data set included in the `{traveltime}` package is a 563 \$\*\$ 2 `matrix` containing the longitude (x) and latitude (y) of all LRT and MRT station exits in Singapore from @singdata:2024.
 
@@
 
-So now we have the two items of data we require. The next step is to download a friction surface for our area of interest.
-
+Now that we have the two items of data that require, the next step is to download a friction surface for our area of interest.
 We can pass in our basemap `sin` (@fig-basemap), a `SpatVector`, directly as the `extent`.
-It would also be possible pass in a `SpatRaster` with extent equal to our area of interest, or specify coordinates either as a vector, matrix, or `SpatExtent`.
-
 We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`.
-Alternatively, we could use `"motor2020"` for motorised travel if that was of interest.
+(Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.)
 
@@
 
+Thus we have our friction surface as a `SpatRaster`:
+
 ```{r show friction surface raster }
@@
 
-Taking a look at these data, below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points @fig-data.
+Below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points @fig-data.
 Higher values of friction indicate more time travelling across a given cell.
@@
 
-With all the data collected, the function `calculate_travel_time` takes the friction surface `friction_singapore` and the locations of interest `stations` and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station.
+With all the data collected, the function `calculate_travel_time` takes the friction surface `friction_singapore` and the locations of interest `stations`, and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station as a `SpatRaster`:
 
@@
 )
+
 travel_time_sin
@@
 
-We present those results in @fig-result:
+We present the resulting calculated travel times in @fig-result:
 
@@
 
-Note that the above raster include infinite (`Inf`) values.
+Note that the above raster `travel_time_sin` includes infinite (`Inf`) values.
 In @fig-data, a number of islands to the south are shown as cells unconnected with the mainland.
@@
 
-The `traveltime` package is immediately suitable to a range of applications where travel to custom locations of interest.
-But we nonetheless see opportunities to build the package utility into the future through two mechanisms: capability to distribute more friction surfaces, and additional methods to handle large spatial extents.
+The `{traveltime}` package is immediately suitable to a range of applications where travel to custom locations of interest.
+Nonetheless, we see opportunities to build the package utility into the future through two mechanisms: capability to distribute more a wider range friction surfaces, and additional methods to handle large spatial extents.
 
-Firstly, `traveltime` currently has access to friction surfaces for 2020, both at 30 arc-second resolution.
-As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available.
+Firstly, `{traveltime}` currently has access to friction surfaces for 2020, both at 30 arc-second resolution.
+As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available, though this is likely to occur first through `malariaAtlas`.
 Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses.
@@
 
-At the other end of the scale, the package can require large amounts of memory for analyses over large landscapes (e.g.
+At the other end of the scale, the package can require relatively large amounts of onboard memory for analyses over large landscapes (e.g.
 one analyses over Africa required \~ 72 GB RAM).
-Developing methods to handle large landscapes with less memory would be helpful to make such analyses accessible to those without access to larger computing resources.
+Developing methods to handle large landscapes either with less memory or via cloud resources would be helpful to make such analyses accessible to those without access to larger computing resources.
 

### 2025-01-28 R/calculate_travel_time.R
clarify docs wording for tibble points

@@ -5,9 +5,10 @@
 #'
 #' @param friction_surface A `SpatRaster` friction surface layer. See
 #'   `?get_friction_surface`
-#' @param points A two-column `matrix`, `data.frame`, or `tibble` with longitude
-#'   (x) in the first column and latitude (y) in the second, or a `SpatVector`,
-#'   in the same coordinate, reference system as `friction_surface`.
+#' @param points A two-column `matrix`, `data.frame`, (including `tibble` types)
+#'   with longitude (x) in the first column and latitude (y) in the second, or a
+#'   `SpatVector`, in the same coordinate, reference system as
+#'   `friction_surface`.
 #' @param filename `character`. Output file name with extension suitable for
 #'   `terra::writeRaster`
 #' @param overwrite `logical`. If `TRUE` `filename` is overwritten.

### 2025-01-28 paper/paper.qmd
Paper: Additional citations to tt surfaces per DW

@@
 
-Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@weiss2020global] (@ref-tt_healthcare) and likely a myriad others.
+Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [hulland2019travel; @weiss2020global] (@ref-tt_healthcare) and a myriad others [@nelson2019suite].
 Here we present a software package --- `{traveltime}` --- written in and for the language R [@Rref].
@@
 
-Global maps of travel time to cities [@weiss2018global] and health care facilities [@weiss2020global] (@fig-tt_healthcare) have generated significant interest and use.
-Collectively @weiss2018global ; @weiss2020global have over 1300 citations (Google Scholar, at 10th December 2024), and the city data set is available to R users through the widely-used `geodata` package [@geodata].
+Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] (@fig-tt_healthcare) have generated significant interest and use (collectively >1500 citations per google scholar at the 28th of January 2025), and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata].
 There is clear demand for these type of products.
@@
 
-The the `stations` data set included in the `{traveltime}` package is a 563 \$\*\$ 2 `matrix` containing the longitude (x) and latitude (y) of all LRT and MRT station exits in Singapore from @singdata:2024.
+The the `stations` data set included in the `{traveltime}` package is a 563 \$\*\$ 2 `matrix` containing the longitude (x) and latitude (y) of all LRT and MRT station exits in Singapore from @singdata.
 
@@
 
-We present the resulting calculated travel times in @fig-result:
+We present the resulting calculated travel times in @fig-result
 

### 2025-01-29 paper/paper.qmd
alternate label and caption method for cross refs - fails in html render but will it work in yml pdf render?

@@
 
-Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [hulland2019travel; @weiss2020global] (@ref-tt_healthcare) and a myriad others [@nelson2019suite].
+Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health (@hulland2019travel; @weiss2020global; \ref{fig:fig-tt_healthcare}) and a myriad others [@nelson2019suite].
 Here we present a software package --- `{traveltime}` --- written in and for the language R [@Rref].
 `{traveltime}` enables a user to supply a set of geographic point locations and an extent of interest, to then calculate the motorised or walking travel time over that extent.
-The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest (e.g., travel time from health care facilities @ref-tt_healthcare from @weiss2020global).
+The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest (e.g., travel time from health care facilities  \ref{fig:fig-tt_healthcare} from @weiss2020global).
 
@@
 
-```{r ttexample_image, echo=FALSE}
-#| label: "fig-tt_healthcare"
-#| fig-cap: "Existing surface of motorised travel time from healthcare facilities, exerpted to Cambodia (Weiss et al., 2020)"
+```{r ttexample_image, echo=FALSE, fig.cap = "\\label{fig:fig-tt_healthcare}Existing surface of motorised travel time from healthcare facilities, exerpted to Cambodia (Weiss et al., 2020)"}
 library(ggplot2)
@@
 
-@weiss2020global and colleagues made their code available as an R script to allow for reproduction and extension of their analyses (<https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt> ).
+@weiss2020globa made their code available as an R script to allow for reproduction and extension of their analyses (<https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt> ).
 To further enable extension of this work, here we have developed an R package based on that code to seamlessly calculate the travel time from any arbitrary set of locations.

### 2025-01-29 paper/paper.qmd
Alt figure labelling throughout and latex cross refs in text. Attempt at #43 Figures now rendering in html but cross refs not working. Will the pdf work out properly? :scream:

@@
 
-Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health (@hulland2019travel; @weiss2020global; \ref{fig:fig-tt_healthcare}) and a myriad others [@nelson2019suite].
+Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health (@hulland2019travel; @weiss2020global; Figure \ref{fig:tt_healthcare}) and a myriad others [@nelson2019suite].
 Here we present a software package --- `{traveltime}` --- written in and for the language R [@Rref].
 `{traveltime}` enables a user to supply a set of geographic point locations and an extent of interest, to then calculate the motorised or walking travel time over that extent.
-The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest (e.g., travel time from health care facilities  \ref{fig:fig-tt_healthcare} from @weiss2020global).
+The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest (e.g., travel time from health care facilities in Figure \ref{fig:tt_healthcare} from @weiss2020global).
 
@@
 
-```{r ttexample_image, echo=FALSE, fig.cap = "\\label{fig:fig-tt_healthcare}Existing surface of motorised travel time from healthcare facilities, exerpted to Cambodia (Weiss et al., 2020)"}
+```{r ttexample_image, echo=FALSE, fig.cap = "\\label{fig:tt_healthcare}Existing surface of motorised travel time from healthcare facilities, exerpted to Cambodia (Weiss et al., 2020)"}
 library(ggplot2)
@@
 
-```{r plot basemap}
-#| label: "fig-basemap"
-#| fig-cap: "Area of interest: Singapore"
+```{r plot basemap, fig.cap = "\\label{fig:basemap}Area of interest: Singapore."}
 
@@
 Now that we have the two items of data that require, the next step is to download a friction surface for our area of interest.
-We can pass in our basemap `sin` (@fig-basemap), a `SpatVector`, directly as the `extent`.
+We can pass in our basemap `sin` (Figure \ref{fig:basemap}), a `SpatVector`, directly as the `extent`.
 We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`.
@@
 
-```{r plot data}
-#| label: "fig-data"
-#| fig-cap: "Friction surface raster of Singapore, showing Singapore boundary in grey, and station locations as grey points"
+```{r plot data, fig.cap = "\\label{fig:data}Friction surface raster of Singapore, showing Singapore boundary in grey, and station locations as grey points."}
 
@@
 
-We present the resulting calculated travel times in @fig-result
+We present the resulting calculated travel times in Figure \ref{fig:data}
 
-```{r plot result}
-#| label: "fig-result"
-#| fig-cap: "Map of walking travel time in Singapore, in minutes from nearest MRT or LRT station."
-
+```{r plot result, fig.cap = "\\label{fig:result}Map of walking travel time in Singapore, in minutes from nearest MRT or LRT station."}
 ggplot() +
@@
 Note that the above raster `travel_time_sin` includes infinite (`Inf`) values.
-In @fig-data, a number of islands to the south are shown as cells unconnected with the mainland.
-These raster cells for these islands appear absent in @fig-result.
-Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in the plot @fig-result.
+In Figure \ref{fig:data}, a number of islands to the south are shown as cells unconnected with the mainland.
+These raster cells for these islands appear absent in Figure \ref{fig:result}.
+Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig:result}.
 

### 2025-01-29 paper/paper.qmd
get rid of previous figures 1 and 2 (healthcare tt from weiss 2020 and outline of SIN). Unnecessary

@@
 
-Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health (@hulland2019travel; @weiss2020global; Figure \ref{fig:tt_healthcare}) and a myriad others [@nelson2019suite].
+Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and a myriad others [@nelson2019suite].
 Here we present a software package --- `{traveltime}` --- written in and for the language R [@Rref].
 `{traveltime}` enables a user to supply a set of geographic point locations and an extent of interest, to then calculate the motorised or walking travel time over that extent.
-The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest (e.g., travel time from health care facilities in Figure \ref{fig:tt_healthcare} from @weiss2020global).
+The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest.
 
@@
 
-```{r ttexample,  include=FALSE}
-library(terra)
-library(malariaAtlas)
-
-boundary_kh <- getShp(ISO = "KHM")
-
-tt_kh <- getRaster(
-  dataset_id = "Accessibility__202001_Global_Motorized_Travel_Time_to_Healthcare",
-  shp = boundary_kh
-)
-```
-
-```{r ttexample_image, echo=FALSE, fig.cap = "\\label{fig:tt_healthcare}Existing surface of motorised travel time from healthcare facilities, exerpted to Cambodia (Weiss et al., 2020)"}
-library(ggplot2)
-library(tidyterra)
-
-ggplot() +
-  geom_spatraster(
-    data = tt_kh
-  ) +
-  scale_fill_viridis_c(
-    option = "A",
-    direction = -1,
-    na.value = "transparent"
-  ) +
-  theme_minimal() +
-  labs(fill = "minutes")
-```
-
 `{traveltime}` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and has reference documents at <https://idem-lab.github.io/traveltime/>.
-Although this article is intended to be the key reference for the `{traveltime}` package, we suggest citations of the package should be accompanied by citing the underlying work by @weiss2018global; @weiss2020global as well.
+Although this article is intended to be the key reference for the `{traveltime}` package, we suggest citations of the package should be accompanied by citing the underlying work [@weiss2018global; @weiss2020global] as well.
 
@@
 
-Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] (@fig-tt_healthcare) have generated significant interest and use (collectively >1500 citations per google scholar at the 28th of January 2025), and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata].
+Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] have generated significant interest and use (collectively >1500 citations per google scholar at the 28th of January 2025), and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata].
 There is clear demand for these type of products.
 
-@weiss2020globa made their code available as an R script to allow for reproduction and extension of their analyses (<https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt> ).
+@weiss2020global made their code available as an R script to allow for reproduction and extension of their analyses (<https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt> ).
 To further enable extension of this work, here we have developed an R package based on that code to seamlessly calculate the travel time from any arbitrary set of locations.
@@
 ```{r get basemap}
+library(terra)
 library(geodata)
@@
 
-```{r plot basemap, fig.cap = "\\label{fig:basemap}Area of interest: Singapore."}
-
-ggplot() +
-  geom_spatvector(
-    data = sin
-  ) +
-  scale_fill_manual(na.value = "transparent") +
-  theme_minimal()
-```
-
 The the `stations` data set included in the `{traveltime}` package is a 563 \$\*\$ 2 `matrix` containing the longitude (x) and latitude (y) of all LRT and MRT station exits in Singapore from @singdata.
@@
 Now that we have the two items of data that require, the next step is to download a friction surface for our area of interest.
-We can pass in our basemap `sin` (Figure \ref{fig:basemap}), a `SpatVector`, directly as the `extent`.
+We can pass in our basemap `sin`, a `SpatVector`, directly as the `extent`.
 We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`.
@@
 ```{r  download friction surface, echo=TRUE, output=FALSE}
-library(traveltime)
-
 friction_singapore <- get_friction_surface(
@@
 
-Below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points @fig-data.
+Below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points (Figure \ref{fig:data}).
 Higher values of friction indicate more time travelling across a given cell.
@@
 ```{r plot data, fig.cap = "\\label{fig:data}Friction surface raster of Singapore, showing Singapore boundary in grey, and station locations as grey points."}
-
 library(tidyterra)

### 2025-01-29 paper/paper.qmd
paper: polishing / tidying of ms and repo

@@
 authors:
-  - name: Gerard E Ryan
+  - name: Gerard E. Ryan
     orcid: 0000-0003-0183-7630
@@
     affiliation: "1, 2"
-  - name: Nick Tierney
+  - name: Nicholas Tierney
     orcid: 0000-0003-1460-8722
@@
     affiliation: "1, 3 "
-  - name: Daniel Weiss
+  - name: Daniel J. Weiss
     orcid: 0000-0002-6175-5648
@@
 
-Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and a myriad others [@nelson2019suite].
+Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite].
 Here we present a software package --- `{traveltime}` --- written in and for the language R [@Rref].
-`{traveltime}` enables a user to supply a set of geographic point locations and an extent of interest, to then calculate the motorised or walking travel time over that extent.
-The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest.
+`{traveltime}` enables a user to calculate the motorised or walking travel time over an area of interest from a user-specified set of geographic coordinates.
+The result is a raster of the area of interest, of 30 arcsecond resolution[^1] where the value in each cell is the lowest travel time in minutes to any of the specified locations.
+
+[^1]: Approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator
 
@@
 `{traveltime}` provides a spatial interface using object classes from the `terra` package `[@terra]`.
-It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and extents as `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` objects; and returns the result as a `SpatRaster` object.
+It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` object; and returns the result as a `SpatRaster` object.
 The package uses `gdistance` [@gdistance2017] to calculate the minimum least-cost-distance for each cell from the points of interest.
-The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by @weiss2020global available via the `malariaAtlas` package [@pfeffer2018malariaatlas].
-Rather than house and manipulate these surfaces independently, `{traveltime}` relies on `malariaAtlas` to download the surfaces for the extent of interest.
+The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by @weiss2020global and available from the `malariaAtlas` package [@pfeffer2018malariaatlas].
+Rather than house and manipulate these surfaces independently, `{traveltime}` relies internally on `malariaAtlas` to download the surfaces for the area of interest.
 
@@
 
-Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] have generated significant interest and use (collectively >1500 citations per google scholar at the 28th of January 2025), and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata].
+Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] have generated significant interest and use[^2], and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata].
 There is clear demand for these type of products.
+
+[^2]: Collectively >1600 citations per Google Scholar at the 28th of January 2025.
 
@@
 Other R packages produce superficially similar though fundamentally different functionality.
-A gaggle of R packages provide interfaces to the <https://www.TravelTime.com> API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh].
+A gaggle of R packages provide interfaces to the [TravelTime.com](https://www.TravelTime.com) API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh].
 The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location.
 The isochron polygons are most comparable to what the `{traveltime}` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across cells.
-TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repetitive iteration for all combinations of time and points, and additional calculations of the minimum value for each cell from all points.
-Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use.
+TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, and additional calculations of the minimum value for each cell from all points.
+Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use of this resource.
 
@@
 
--   our extent of interest --- in this case a map of Singapore, and
+-   our area of interest --- in this case a map of Singapore, and
 -   our points to calculate travel time from --- here the locations of Singapore's MRT and LRT stations.
@@
 
-The the `stations` data set included in the `{traveltime}` package is a 563 \$\*\$ 2 `matrix` containing the longitude (x) and latitude (y) of all LRT and MRT station exits in Singapore from @singdata.
+The the `stations` data set included in the `{traveltime}` package is a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from @singdata.
 
@@
 We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`.
-(Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.)
-
-We're also only interested in walking *on land* so we mask out areas outside of `sin`, that are within the extent of the raster:
+(Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we mask out areas outside of `sin`, that are within the extent of the raster:
 
@@
 
-We present the resulting calculated travel times in Figure \ref{fig:data}
+We present the resulting calculated travel times in Figure \ref{fig:result}.
+Note that the results in `travel_time_sin` include infinite (`Inf`) values.
+In Figure \ref{fig:data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
+These raster cells for these islands appear absent in Figure \ref{fig:result}.
+Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig:result}.
 
@@
   theme_minimal() +
-  labs(fill = "minutes") +
+  labs(fill = "Minutes") +
   geom_spatvector(
@@
 
-Note that the above raster `travel_time_sin` includes infinite (`Inf`) values.
-In Figure \ref{fig:data}, a number of islands to the south are shown as cells unconnected with the mainland.
-These raster cells for these islands appear absent in Figure \ref{fig:result}.
-Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig:result}.
-
 # Opportunities for future development
@@
 
-Firstly, `{traveltime}` currently has access to friction surfaces for 2020, both at 30 arc-second resolution.
+Firstly, `{traveltime}` currently has access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution.
 As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available, though this is likely to occur first through `malariaAtlas`.
 Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses.
-For instance, although the example here was chosen for it's simplicity and low computational demands, a 5 km cell size is a relatively large area to walk across, and thus actual waking times are likley to vary significantly within each cell.
+For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km$^2$ cell size is a relatively large area to walk across, and thus actual waking times are likley to vary significantly within each cell.
+We note though that the capability to calculate travel time over any friction surface exists presently, it is only the availability of additional surfaces that is presently restrictive.
 
-At the other end of the scale, the package can require relatively large amounts of onboard memory for analyses over large landscapes (e.g.
+At the other end of the scale, the calculations can require relatively large amounts of onboard memory for analyses over large landscapes (e.g.
 one analyses over Africa required \~ 72 GB RAM).

### 2025-01-29 paper/paper.qmd
paper - further polishing - get rid of ggplot code

@@
 
-`{traveltime}` provides a spatial interface using object classes from the `terra` package `[@terra]`.
+`{traveltime}` provides a spatial interface using object classes from the `terra` package [@terra].
 It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` object; and returns the result as a `SpatRaster` object.
@@
 
-Other R packages produce superficially similar though fundamentally different functionality.
+Other R packages provide superficially similar though fundamentally different functionality.
 A gaggle of R packages provide interfaces to the [TravelTime.com](https://www.TravelTime.com) API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh].
-The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location.
-The isochron polygons are most comparable to what the `{traveltime}` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across cells.
-TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, and additional calculations of the minimum value for each cell from all points.
+The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons --- areas reachable within a given time from a given location.
+The isochron polygons are most comparable to what the `{traveltime}` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across a landscape for multiple points.
+TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, plus additional calculation of the minimum value for each cell from all points.
 Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use of this resource.
@@
 We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`.
-(Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we mask out areas outside of `sin`, that are within the extent of the raster:
+(Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we then mask out areas outside of `sin`, that are within the extent of the friction surface raster:
 
@@
 
-Below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points (Figure \ref{fig:data}).
+Below we plot the friction surface raster `friction_singapore`, with the vector boundary `sin` as a grey line, and `stations` as grey points (Figure \ref{fig:data}).
 Higher values of friction indicate more time travelling across a given cell.
 
-```{r plot data, fig.cap = "\\label{fig:data}Friction surface raster of Singapore, showing Singapore boundary in grey, and station locations as grey points."}
+```{r plot data, echo=FALSE, fig.cap = "\\label{fig:data}Friction surface raster of Singapore, showing Singapore boundary in grey, and station locations as grey points."}
 library(tidyterra)
@@
 
-With all the data collected, the function `calculate_travel_time` takes the friction surface `friction_singapore` and the locations of interest `stations`, and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station as a `SpatRaster`:
+With all the data collected, the function `calculate_travel_time` takes the friction surface `friction_singapore` and the locations of interest `stations`, and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station:
 
@@
 In Figure \ref{fig:data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
-These raster cells for these islands appear absent in Figure \ref{fig:result}.
+The raster cells for these islands appear absent in Figure \ref{fig:result}.
 Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig:result}.
 
-```{r plot result, fig.cap = "\\label{fig:result}Map of walking travel time in Singapore, in minutes from nearest MRT or LRT station."}
+```{r plot result, echo=FALSE,  fig.cap = "\\label{fig:result}Map of walking travel time in Singapore, in minutes from nearest MRT or LRT station."}
 ggplot() +

### 2025-01-30 paper/paper.qmd
fix r cmd check warning by replacing spaces in chunk names so file names have no spaces (and are "portable")

@@
 
-```{r plot data, echo=FALSE, fig.cap = "\\label{fig:data}Friction surface raster of Singapore, showing Singapore boundary in grey, and station locations as grey points."}
+```{r plot_data, echo=FALSE, fig.cap = "\\label{fig:data}Friction surface raster of Singapore, showing Singapore boundary in grey, and station locations as grey points."}
 library(tidyterra)
@@
 
-```{r plot result, echo=FALSE,  fig.cap = "\\label{fig:result}Map of walking travel time in Singapore, in minutes from nearest MRT or LRT station."}
+```{r plot_result, echo=FALSE,  fig.cap = "\\label{fig:result}Map of walking travel time in Singapore, in minutes from nearest MRT or LRT station."}
 ggplot() +

### 2025-02-04 paper/paper.qmd
updates following NG comments #45 still incomplete

@@
 ---
-title: "`{traveltime}`: an R package to calculate walking and motorised travel time from user-specified locations"
+title: "`traveltime`: an R package to calculate walking and motorised travel time from user-specified locations"
 format:
@@
     orcid: 0000-0001-8916-5570
-    affiliation: "1, 3 "
+    affiliation: "1, 4"
   - name: Daniel J. Weiss
@@
    index: 3
+ - name: University of Western Australia, WA, Australia
+   index: 4
 date: In prep 2025
@@
 Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite].
-Here we present a software package --- `{traveltime}` --- written in and for the language R [@Rref].
-`{traveltime}` enables a user to calculate the motorised or walking travel time over an area of interest from a user-specified set of geographic coordinates.
-The result is a raster of the area of interest, of 30 arcsecond resolution[^1] where the value in each cell is the lowest travel time in minutes to any of the specified locations.
-
-[^1]: Approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator
+Here we present a software package --- `traveltime` --- written in and for the language R [@Rref].
+`traveltime` enables a user to create a map of the motorised or walking travel time over an area of interest from a user-specified set of geographic coordinates.
+The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations.
+We envisage these maps as having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias, @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], or understanding access to agricultural processing facilities [@zhao2023replanting].
 
@@
 
--   downloading a 'friction surface' for the area of interest, and then
+-   preparing a 'friction surface' for the area of interest, and then
 -   calculating travel time over that surface for the points of interest.
 
-`{traveltime}` provides a spatial interface using object classes from the `terra` package [@terra].
+`traveltime` provides a spatial interface using object classes from the `terra` package [@terra].
 It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` object; and returns the result as a `SpatRaster` object.
-The package uses `gdistance` [@gdistance2017] to calculate the minimum least-cost-distance for each cell from the points of interest.
-The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by @weiss2020global and available from the `malariaAtlas` package [@pfeffer2018malariaatlas].
-Rather than house and manipulate these surfaces independently, `{traveltime}` relies internally on `malariaAtlas` to download the surfaces for the area of interest.
+Internally, `traveltime` uses a number of existing R packages to perform specific operations.
+`gdistance` [@gdistance2017] is used to calculate the minimum least-cost-distance for each cell from the points of interest.
+The travel time is calculated as movement over any 'friction surface' the user supplies.
+To provide easy access to the existing friction surfaces generated by @weiss2020global.
+`traveltime` uses the package `malariaAtlas` to access friction surfaces for the desired extent, potentially saving the user downloading global surfaces and cropping them separately.
 
-`{traveltime}` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and has reference documents at <https://idem-lab.github.io/traveltime/>.
-Although this article is intended to be the key reference for the `{traveltime}` package, we suggest citations of the package should be accompanied by citing the underlying work [@weiss2018global; @weiss2020global] as well.
+`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and has reference documents at <https://idem-lab.github.io/traveltime/>.
+Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work [@weiss2018global; @weiss2020global] as well.
 
@@
 
-Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] have generated significant interest and use[^2], and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata].
+Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] have generated significant interest and use[^1], and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata].
 There is clear demand for these type of products.
 
-[^2]: Collectively >1600 citations per Google Scholar at the 28th of January 2025.
+[^1]: Collectively >1600 citations per Google Scholar at the 28th of January 2025.
 
@@
 The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons --- areas reachable within a given time from a given location.
-The isochron polygons are most comparable to what the `{traveltime}` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across a landscape for multiple points.
+The isochron polygons are most comparable to what the `traveltime` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across a landscape for multiple points.
 TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, plus additional calculation of the minimum value for each cell from all points.
@@
 
-With `{traveltime}`, we provide free and open source software to estimate motorised or walking travel time from any number of user-supplied locations, across a complete area of interest.
+With `traveltime`, we provide free and open source software to estimate motorised or walking travel time from any number of user-supplied locations, across a complete area of interest.
 
@@
 
-The the `stations` data set included in the `{traveltime}` package is a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from @singdata.
+The the `stations` data set included in the `traveltime` package is a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from @singdata.
 
@@
 
-Now that we have the two items of data that require, the next step is to download a friction surface for our area of interest.
+Now that we have the two items of data that we require, the next step is to download a friction surface for our area of interest.
 We can pass in our basemap `sin`, a `SpatVector`, directly as the `extent`.
@@
 
-The `{traveltime}` package is immediately suitable to a range of applications where travel to custom locations of interest.
+The `traveltime` package is immediately suitable to a range of applications where travel to custom locations of interest.
 Nonetheless, we see opportunities to build the package utility into the future through two mechanisms: capability to distribute more a wider range friction surfaces, and additional methods to handle large spatial extents.
 
-Firstly, `{traveltime}` currently has access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution.
+Firstly, `traveltime` currently has access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution[^2].
 As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available, though this is likely to occur first through `malariaAtlas`.
 Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses.
-For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km$^2$ cell size is a relatively large area to walk across, and thus actual waking times are likley to vary significantly within each cell.
+For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km$^2$ cell size is a relatively large area to walk across, and thus actual waking times are likeley to vary significantly within each cell.
 We note though that the capability to calculate travel time over any friction surface exists presently, it is only the availability of additional surfaces that is presently restrictive.
+
+[^2]: Approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator
 

### 2025-02-05 paper/paper.qmd
more updates re #45

@@
 Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses.
-For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km$^2$ cell size is a relatively large area to walk across, and thus actual waking times are likeley to vary significantly within each cell.
+For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km$^2$ cell size is a relatively large area to walk across, and thus actual waking times are likely to vary significantly within each cell.
 We note though that the capability to calculate travel time over any friction surface exists presently, it is only the availability of additional surfaces that is presently restrictive.
@@
 
-GR is supported by the Gates Foundation grant for the Vector Atlas.
+This work was supported, in whole or in part, by the Bill & Melinda Gates Foundation [INV-021972].
+The conclusions and opinions expressed in this work are those of the author(s) alone and shall not be attributed to the Foundation.
+Under the grant conditions of the Foundation, a Creative Commons Attribution 4.0 License has already been assigned to the Author Accepted Manuscript version that might arise from this submission.
+Please note works submitted as a preprint have not undergone a peer review process.
 

### 2025-02-06 paper/paper.qmd
A few minor changes to the quarto document:
* Use #| labels instead of inline code chunk
* labels use `-` as separator (so "get basemap" --> "get-basemap")
* Define MRT and LRT acronym
* Use singapore_shapefile` rather than `sin`
* remove use of `\ref{fig...}` in quarto text, instead using `@fig-<chunk-label>-` syntax.
* Use `^2^` instead of `$^2$` to use superscript

@@
 
-```{r setup, include=FALSE}
+```{r}
+#| label: setup
+#| include: false
 knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
@@
 
-In this example we wish to calculate the walking travel time from the nearest mass transit station across the island nation of Singapore --- specifically MRT and LRT stations --- and create a map of this.
+In this example we wish to calculate the walking travel time from the nearest mass transit station across the island nation of Singapore --- specifically Mass Rapid Transit (MRT) and Light Rail Transit (LRT) stations --- and create a map of this.
 
@@
 Here we download only the national boundary (`level = 0`) and at a low resolution (`resolution = 2`).
-Our boundary `sin` is a `SpatVector` class object.
+Our boundary `singapore_shapefile` is a `SpatVector` class object.
 
-```{r get basemap}
+```{r}
+#| label: get-basemap
 library(terra)
@@
 
-sin <- gadm(
+singapore_shapefile <- gadm(
   country = "Singapore",
@@
 
-sin
+singapore_shapefile
 ```
 
-The the `stations` data set included in the `traveltime` package is a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from @singdata.
+The `stations` data set included in the `traveltime` package is a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from @singdata:
 
-```{r show stations}
+```{r}
+#| label: show-stations
 library(traveltime)
@@
 Now that we have the two items of data that we require, the next step is to download a friction surface for our area of interest.
-We can pass in our basemap `sin`, a `SpatVector`, directly as the `extent`.
+We can pass in our basemap `singapore_shapefile`, a `SpatVector`, directly as the `extent`.
 We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`.
-(Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we then mask out areas outside of `sin`, that are within the extent of the friction surface raster:
+(Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we then mask out areas outside of `singapore_shapefile`, that are within the extent of the friction surface raster:
 
-```{r  download friction surface, echo=TRUE, output=FALSE}
+```{r}
+#| label: download-friction-surface
+#| echo: true
+#| output: false
 friction_singapore <- get_friction_surface(
     surface = "walk2020",
-    extent = sin
+    extent = singapore_shapefile
   )|> 
-  mask(sin)
+  mask(singapore_shapefile)
 ```
@@
 
-```{r show friction surface raster }
+```{r }
+#| label: show-friction-surface-raster
 friction_singapore
@@
 
-Below we plot the friction surface raster `friction_singapore`, with the vector boundary `sin` as a grey line, and `stations` as grey points (Figure \ref{fig:data}).
+Below we plot the friction surface raster `friction_singapore`, with the vector boundary `singapore_shapefile` as a grey line, and `stations` as grey points (Figure @fig-data).
 Higher values of friction indicate more time travelling across a given cell.
 
-```{r plot_data, echo=FALSE, fig.cap = "\\label{fig:data}Friction surface raster of Singapore, showing Singapore boundary in grey, and station locations as grey points."}
+```{r}
+#| label: fig-data
+#| echo: false
+#| fig-cap: "Friction surface raster of Singapore, showing Singapore boundary
+#|   in grey, and station locations as grey points."
 library(tidyterra)
@@
   geom_spatvector(
-    data = sin,
+    data = singapore_shapefile,
     fill = "transparent",
@@
 
-With all the data collected, the function `calculate_travel_time` takes the friction surface `friction_singapore` and the locations of interest `stations`, and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station:
+With all the data collected, the function `calculate_travel_time()` takes the friction surface `friction_singapore()` and the locations of interest `stations`, and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station:
 
-```{r calculate result}
-travel_time_sin <- calculate_travel_time(
+```{r}
+#| label: calculate-result
+travel_timesingapore <- calculate_travel_time(
   friction_surface = friction_singapore,
@@
 
-travel_time_sin
+travel_timesingapore
 ```
 
-We present the resulting calculated travel times in Figure \ref{fig:result}.
-Note that the results in `travel_time_sin` include infinite (`Inf`) values.
-In Figure \ref{fig:data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
-The raster cells for these islands appear absent in Figure \ref{fig:result}.
-Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig:result}.
+We present the resulting calculated travel times in Figure @fig-result.
+Note that the results in `travel_timesingapore` include infinite (`Inf`) values.
+In Figure @fig-data, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
+The raster cells for these islands appear absent in Figure @fig-result.
+Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure @fig-result.
 
-```{r plot_result, echo=FALSE,  fig.cap = "\\label{fig:result}Map of walking travel time in Singapore, in minutes from nearest MRT or LRT station."}
+```{r}
+#| label: fig-result
+#| echo: false
+#| fig-cap: "Map of walking travel time in Singapore, in minutes from 
+#|   nearest MRT or LRT station."
 ggplot() +
   geom_spatraster(
-    data = travel_time_sin
+    data = travel_timesingapore
   ) +
@@
 
-The `traveltime` package is immediately suitable to a range of applications where travel to custom locations of interest.
-Nonetheless, we see opportunities to build the package utility into the future through two mechanisms: capability to distribute more a wider range friction surfaces, and additional methods to handle large spatial extents.
+The `traveltime` package is immediately suitable to a range of applications where travel to custom locations is of interest.
+Nonetheless, we see opportunities to build the package utility into the future through two mechanisms: (1) capability to better distribute a wider range friction surfaces, and (2) additional methods to handle large spatial extents.
 
@@
 Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses.
-For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km$^2$ cell size is a relatively large area to walk across, and thus actual waking times are likely to vary significantly within each cell.
+For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km^2^ cell size is a relatively large area to walk across, and thus actual waking times are likely to vary significantly within each cell.
 We note though that the capability to calculate travel time over any friction surface exists presently, it is only the availability of additional surfaces that is presently restrictive.
 
-[^2]: Approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator
+[^2]: Approximately 0.008333 decimal degrees, or just below 1 km^2^ at the equator
 

### 2025-02-06 paper/paper.qmd
minor typos

--- 2025-02-05 paper/paper.qmd
@@
 The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations.
-We envisage these maps as having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias, @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], or understanding access to agricultural processing facilities [@zhao2023replanting].
+We envisage this software having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias, @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], or understanding access to agricultural processing facilities [@zhao2023replanting].
 
@@
 This work was supported, in whole or in part, by the Bill & Melinda Gates Foundation [INV-021972].
-The conclusions and opinions expressed in this work are those of the author(s) alone and shall not be attributed to the Foundation.
+The conclusions and opinions expressed in this work are those of the authors alone and shall not be attributed to the Foundation.
 Under the grant conditions of the Foundation, a Creative Commons Attribution 4.0 License has already been assigned to the Author Accepted Manuscript version that might arise from this submission.

### 2025-02-06 paper/paper.qmd
Roll back figure cross refs from #47 in attempt to address #43 - use \ref{figure} format instead of @figure which quarto likes - this breaks the html render by I think it should make the draft-yml action render the cross refs.
If this fails might need to roll back chunks

@@
 
-Below we plot the friction surface raster `friction_singapore`, with the vector boundary `singapore_shapefile` as a grey line, and `stations` as grey points (Figure @fig-data).
+Below we plot the friction surface raster `friction_singapore`, with the vector boundary `singapore_shapefile` as a grey line, and `stations` as grey points (Figure \ref{fig-data}).
 Higher values of friction indicate more time travelling across a given cell.
@@
 
-We present the resulting calculated travel times in Figure @fig-result.
+We present the resulting calculated travel times in Figure \ref{fig-result}.
 Note that the results in `travel_timesingapore` include infinite (`Inf`) values.
-In Figure @fig-data, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
-The raster cells for these islands appear absent in Figure @fig-result.
-Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure @fig-result.
+In Figure \ref{fig-data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
+The raster cells for these islands appear absent in Figure \ref{fig-result}.
+Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig-result}.
 

### 2025-02-06 paper/paper.qmd
replacing `travel_timesingapore` with `travel_time_singapore`
I cannot cope with the underscore then no space between words in one object name, sorry @njtierney

@@
 #| label: calculate-result
-travel_timesingapore <- calculate_travel_time(
+trave_time_singapore <- calculate_travel_time(
   friction_surface = friction_singapore,
@@
 
-travel_timesingapore
+trave_time_singapore
 ```
@@
 We present the resulting calculated travel times in Figure \ref{fig-result}.
-Note that the results in `travel_timesingapore` include infinite (`Inf`) values.
+Note that the results in `trave_time_singapore` include infinite (`Inf`) values.
 In Figure \ref{fig-data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
@@
   geom_spatraster(
-    data = travel_timesingapore
+    data = trave_time_singapore
   ) +

### 2025-02-10 R/calculate_travel_time.R
tidy and link documentation for `calculate_travel_time`

@@ -3,8 +3,8 @@
 #'   surface.
 #'
 #'
-#' @param friction_surface A `SpatRaster` friction surface layer. See
-#'   `?get_friction_surface`
+#' @param friction_surface A `SpatRaster` friction surface layer with in
+#'   resistance units. See Details.
 #' @param points A two-column `matrix`, `data.frame`, (including `tibble` types)
 #'   with longitude (x) in the first column and latitude (y) in the second, or a
 #'   `SpatVector`, in the same coordinate, reference system as
@@ -19,20 +19,32 @@
 #'   Over large areas this function can require significant RAM and will be
 #'   slow.
 #'
-#'   Citations: D. J. Weiss, A. Nelson, C. A. Vargas-Ruiz, K. Gligoric, S.,
-#'   Bavadekar, E. Gabrilovich, A. Bertozzi-Villa, J. Rozier, H. S. Gibson, T.,
-#'   Shekel, C. Kamath, A. Lieber, K. Schulman, Y. Shao, V. Qarkaxhija, A. K.
-#'   Nandi, S. H. Keddie, S. Rumisha, P. Amratia, R. Arambepola, E. G.
-#'   Chestnutt, J. J. Millar, T. L. Symons, E. Cameron, K. E. Battle, S. Bhatt,
-#'   and P. W. Gething. Global maps of travel time to healthcare facilities.
-#'   (2020) Nature Medicine. https://doi.org/10.1038/s41591-020-1059-1
+#'   Pre-prepared walking or motorised friction surfaces can be obtained
+#'   with`?get_friction_surface`. User can also provide their own friction
+#'   surface. This surface but must be in resistance units (1/conductance), e.g.
+#'   minutes/meter. See Van Etten 2017.
+#'
+#'   Citations:
+#'
+#'   D. J. Weiss, A. Nelson, C. A. Vargas-Ruiz, K. Gligoric, S., Bavadekar, E.
+#'   Gabrilovich, A. Bertozzi-Villa, J. Rozier, H. S. Gibson, T., Shekel, C.
+#'   Kamath, A. Lieber, K. Schulman, Y. Shao, V. Qarkaxhija, A. K. Nandi, S. H.
+#'   Keddie, S. Rumisha, P. Amratia, R. Arambepola, E. G. Chestnutt, J. J.
+#'   Millar, T. L. Symons, E. Cameron, K. E. Battle, S. Bhatt, and P. W.
[…]

### 2025-02-10 R/calculate_travel_time.R
remove unused code an make overdue updates from @njtierney review in `calculate_travel_time`

@@ -95,34 +95,30 @@ calculate_travel_time <- function(
     )
   }
 
-  # explaining variable
-  if(!any(c("matrix", "data.frame", "SpatVector") %in% class(points))){
+  point_format_ok <- any(c("matrix", "data.frame", "SpatVector") %in% class(points))
+
+  if(!point_format_ok){
     stop("points must be a SpatVector, data.frame, or matrix")
   }
 
-  # use inherits
-  if("SpatVector" %in% class(points)){
+  if(inherits(points, "SpatVector")){
     points <- terra::geom(points)[,c("x", "y")]
   }
-
-  #npoints <- nrow(points)
 
   friction <- raster::raster(friction_surface)
 
   tsn <- gdistance::transition(friction, function(x) 1/mean(x), 8)
-  tgc <- gdistance::geoCorrection(tsn)
 
-  # xy.data.frame <- data.frame()
-  # xy.data.frame[1:npoints,1] <- points[,1]
-  # xy.data.frame[1:npoints,2] <- points[,2]
-  # xy.matrix <- as.matrix(xy.data.frame)
+  tgc <- gdistance::geoCorrection(tsn)
 
   xy.matrix <- as.matrix(points[,1:2])
 
   travel_time <- gdistance::accCost(tgc, xy.matrix)
+
   names(travel_time) <- "travel_time"
 
   if(!is.null(filename)){
+
     raster::writeRaster(
       travel_time,
       filename,
@@ -130,8 +126,11 @@ calculate_travel_time <- function(
     )
 
     return(terra::rast(filename))
+
   } else {
+
     terra::rast(travel_time)
+
   }
 
 }

### 2025-02-10 paper/paper.qmd
extra examples of potential applications of the package

@@
 The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations.
-We envisage this software having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias, @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], or understanding access to agricultural processing facilities [@zhao2023replanting].
+We envisage this software having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias, @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], allocating public defibrillators [tierney2018novel], setting rehabilitation districts for stroke patients [padgham2019introduction], or understanding access to agricultural processing facilities [@zhao2023replanting].
 
@@
 It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` object; and returns the result as a `SpatRaster` object.
-Internally, `traveltime` uses a number of existing R packages to perform specific operations.
-`gdistance` [@gdistance2017] is used to calculate the minimum least-cost-distance for each cell from the points of interest.
-The travel time is calculated as movement over any 'friction surface' the user supplies.
+The travel time is calculated as movement over a resistance 'friction surface' [@gdistance2017].
 To provide easy access to the existing friction surfaces generated by @weiss2020global.
-`traveltime` uses the package `malariaAtlas` to access friction surfaces for the desired extent, potentially saving the user downloading global surfaces and cropping them separately.
+`traveltime` uses the package `malariaAtlas` to access friction surfaces for the desired extent.
+Users can also supply other friction surfaces.
+Functionality from `gdistance` [@gdistance2017] is also used internally to calculate the minimum least-cost-distance for each cell from the points of interest.
 

### 2025-02-10 paper/paper.qmd
more refinements to paper. NB package not passing R CMD CHECK because `malariaAtlas` AWS server down so will need to rerun action to check draft

@@
 
-For this excercise, we need two items of data:
+For this exercise, we need two items of data:
 
@@
 Below we plot the friction surface raster `friction_singapore`, with the vector boundary `singapore_shapefile` as a grey line, and `stations` as grey points (Figure \ref{fig-data}).
-Higher values of friction indicate more time travelling across a given cell.
+`traveltime` takes resistance values of friction [@gdistance2017], so higher values of friction indicate more time travelling across a given cell.
+In this case walking friction is fairly uniform across the islands barring a few small inaccesible locations.
+More heterogeneity would be expected in less developed areas with fewer transport corridors.
 
@@
   labs(
-    fill = "Friction",
+    fill = "Resistance",
     x = element_blank(),
@@
 Firstly, `traveltime` currently has access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution[^2].
+Although the user can supply their own friction surface, we expect most applications will use these surfaces.
 As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available, though this is likely to occur first through `malariaAtlas`.
 Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses.
-For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km^2^ cell size is a relatively large area to walk across, and thus actual waking times are likely to vary significantly within each cell.
-We note though that the capability to calculate travel time over any friction surface exists presently, it is only the availability of additional surfaces that is presently restrictive.
+For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km^2^ cell size is a relatively large area to walk across, and thus actual waking times may vary significantly within each cell.
 
@@
 At the other end of the scale, the calculations can require relatively large amounts of onboard memory for analyses over large landscapes (e.g.
-one analyses over Africa required \~ 72 GB RAM).
+analyses over Africa required \~ 72 GB RAM).
 Developing methods to handle large landscapes either with less memory or via cloud resources would be helpful to make such analyses accessible to those without access to larger computing resources.

### 2025-02-10 R/calculate_travel_time.R
comma begone

@@ -7,7 +7,7 @@
 #'   resistance units. See Details.
 #' @param points A two-column `matrix`, `data.frame`, (including `tibble` types)
 #'   with longitude (x) in the first column and latitude (y) in the second, or a
-#'   `SpatVector`, in the same coordinate, reference system as
+#'   `SpatVector`, in the same coordinate reference system as
 #'   `friction_surface`.
 #' @param filename `character`. Output file name with extension suitable for
 #'   `terra::writeRaster`

### 2025-02-11 paper/paper.qmd
typos and polish

@@
 The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations.
-We envisage this software having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias, @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], allocating public defibrillators [tierney2018novel], setting rehabilitation districts for stroke patients [padgham2019introduction], or understanding access to agricultural processing facilities [@zhao2023replanting].
+We envisage this software having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias, @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], allocating public defibrillators [@tierney2018novel], setting rehabilitation districts for stroke patients [@padgham2019introduction], or understanding access to agricultural processing facilities [@zhao2023replanting].
 
@@
 `traveltime` provides a spatial interface using object classes from the `terra` package [@terra].
-It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` object; and returns the result as a `SpatRaster` object.
+It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` class object; and returns the result as a `SpatRaster` class object.
 The travel time is calculated as movement over a resistance 'friction surface' [@gdistance2017].
-To provide easy access to the existing friction surfaces generated by @weiss2020global.
-`traveltime` uses the package `malariaAtlas` to access friction surfaces for the desired extent.
-Users can also supply other friction surfaces.
+To provide easy access to the existing friction surfaces generated by @weiss2020global, `traveltime` uses the package `malariaAtlas` to allow users to access friction surfaces for the area of interest; though users can also supply any other friction surfaces to `traveltime`.
 Functionality from `gdistance` [@gdistance2017] is also used internally to calculate the minimum least-cost-distance for each cell from the points of interest.
@@
 `traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and has reference documents at <https://idem-lab.github.io/traveltime/>.
-Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work [@weiss2018global; @weiss2020global] as well.
+Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying methodological work [@weiss2018global; @weiss2020global] as well.
 
@@
 The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons --- areas reachable within a given time from a given location.
-The isochron polygons are most comparable to what the `traveltime` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across a landscape for multiple points.
+The isochron polygons are most comparable to what `traveltime::calculate_travel_time()` calculates, though each isochron polygon is a single polygon calculated is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than a raster of gradation across a landscape for an arbitrary number of points as in `traveltime`.
 TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, plus additional calculation of the minimum value for each cell from all points.
@@
 
-With `traveltime`, we provide free and open source software to estimate motorised or walking travel time from any number of user-supplied locations, across a complete area of interest.
+With `traveltime`, we provide free and open source software to estimate travel time from any number of user-supplied locations, across a complete area of interest, and with convenient access to motorised or walking friction surfaces with global coverage.
 
@@
 
-In this example we wish to calculate the walking travel time from the nearest mass transit station across the island nation of Singapore --- specifically Mass Rapid Transit (MRT) and Light Rail Transit (LRT) stations --- and create a map of this.
+In this example we will calculate the walking travel time from the nearest mass transit station across the island nation of Singapore --- specifically Mass Rapid Transit (MRT) and Light Rail Transit (LRT) stations --- and create a map of this.
 
-## Prepare the data and download friction surface
+## Prepare the data and friction surface
 
@@
 
-We can download a national-level polygon of Singapore from the GADM [@gadm] database using the `geodata` package [@geodata].
+We can download a national-level polygon boundary of Singapore from the GADM [@gadm] database using the `geodata` package [@geodata].
 Here we download only the national boundary (`level = 0`) and at a low resolution (`resolution = 2`).
@@
 
-Now that we have the two items of data that we require, the next step is to download a friction surface for our area of interest.
+Now that we have the two items of data that we require, the next step is to prepare a friction surface for our area of interest.
+We will use the friction surface from @weiss2020 that can be downloaded by `traveltime` with the function `get_friction_surface()`.
 We can pass in our basemap `singapore_shapefile`, a `SpatVector`, directly as the `extent`.
 We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`.
-(Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we then mask out areas outside of `singapore_shapefile`, that are within the extent of the friction surface raster:
+(Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we then mask out areas outside of the land boundary in `singapore_shapefile`:
 
@@
 In this case walking friction is fairly uniform across the islands barring a few small inaccesible locations.
-More heterogeneity would be expected in less developed areas with fewer transport corridors.
+More heterogeneity would be expected in less developed locations with fewer transport corridors.
 
@@
 
-With all the data collected, the function `calculate_travel_time()` takes the friction surface `friction_singapore()` and the locations of interest `stations`, and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station:
+With all the data collected, the function `calculate_travel_time()` takes the friction surface `friction_singapore` and the points of interest in `stations`, and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station:
 
@@
 
-We present the resulting calculated travel times in Figure \ref{fig-result}.
-Note that the results in `trave_time_singapore` include infinite (`Inf`) values.
+We present the resulting calculated travel times in Figure \ref{fig-result} where, as expected, the travel times are lowest near station exits (per Figure \ref{fig-data}) and progressively higher further away.
+Note that the results in `trave_time_singapore` include infinite values (`Inf`  above).
 In Figure \ref{fig-data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
@@
 Firstly, `traveltime` currently has access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution[^2].
-Although the user can supply their own friction surface, we expect most applications will use these surfaces.
+Although the user can supply their own friction surface, we expect most applications will use these existing surfaces given the extensive work needed in creating new ones [@weiss2018global; @weiss2020global].
 As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available, though this is likely to occur first through `malariaAtlas`.
@@
 For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km^2^ cell size is a relatively large area to walk across, and thus actual waking times may vary significantly within each cell.
+We underline however that a user can provide their own higher resolution friction surface at present if desired.
 

### 2025-02-11 paper/paper.tex
remove gh action and add latex templates

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Summary},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\date{January 1, 2025}
\title{\bf Summary}
\author{
Gerard E. Ryan\\
1, 2\\
and\\Nicholas Tierney\\
1, 3\\
and\\Nick Golding\\
1, 4\\
and\\Daniel J. Weiss\\
1, 3\\
}
\maketitle

\bigskip
\bigskip
\begin{abstract}

\end{abstract}

\newpage
\spacingset{1.9} % DON'T change the spacing!

Understanding and mapping the time to travel among locations is useful
for many activities from urban planning \citep{zahavi1974traveltime} to
public health \citep{hulland2019travel, weiss2020global} and myriad
others \citep{nelson2019suite}. Here we present a software package ---
\texttt{traveltime} --- written in and for the language R \citep{Rref}.
\texttt{traveltime} enables a user to create a map of the motorised or
walking travel time over an area of interest from a user-specified set
of geographic coordinates. The result is a raster of the area of
interest where the value in each cell is the lowest travel time in
minutes to any of the specified locations. We envisage this software
having diverse applications including: estimating sampling bias in
species occurrence data \citep[@reddy2003geographical]{dennis2000bias},
mapping electric vehicle charger accessibility
\citep{falchetta2021electric}, allocating public defibrillators
\citep{tierney2018novel}, setting rehabilitation districts for stroke
patients \citep{padgham2019introduction}, or understanding access to
agricultural processing facilities \citep{zhao2023replanting}.

The work-flow requires two steps:

\begin{itemize}
\tightlist
\item
  preparing a `friction surface' for the area of interest, and then
\item
  calculating travel time over that surface for the points of interest.
\end{itemize}

\texttt{traveltime} provides a spatial interface using object classes
from the \texttt{terra} package \citep{terra}. It accepts points as
\texttt{matrix}, \texttt{data.frame}, or \texttt{SpatVector} class
objects; and the area of interest as an extent in the form of a
\texttt{numeric}, \texttt{SpatExtent}, \texttt{SpatVector}, or
\texttt{SpatRaster} class object; and returns the result as a
\texttt{SpatRaster} class object. The travel time is calculated as
movement over a resistance `friction surface' \citep{gdistance2017}. To
provide easy access to the existing friction surfaces generated by
\citet{weiss2020global}, \texttt{traveltime} uses the package
\texttt{malariaAtlas} to allow users to access friction surfaces for the
area of interest; though users can also supply any other friction
surfaces to \texttt{traveltime}. Functionality from \texttt{gdistance}
\citep{gdistance2017} is also used internally to calculate the minimum
least-cost-distance for each cell from the points of interest.

\texttt{traveltime} is available from
\href{https://idem-lab.r-universe.dev/traveltime}{R-Universe} and
\href{https://github.com/idem-lab/traveltime}{GitHub}, and has reference
documents at \url{https://idem-lab.github.io/traveltime/}. Although this
article is intended to be the key reference for the \texttt{traveltime}
package, we suggest citations of the package should be accompanied by
citing the underlying methodological work
\citep{weiss2018global, weiss2020global} as well.

Statement of need

Global maps of travel time to cities
\citep{weiss2018global, nelson2019suite} and health care facilities
\citep{hulland2019travel, weiss2020global} have generated significant
interest and use\footnote{Collectively \textgreater1600 citations per
  Google Scholar at the 28th of January 2025.}, and the city data set of
\citet{nelson2019suite} is available to R users through the widely-used
\texttt{geodata} package \citep{geodata}. There is clear demand for
these type of products.

\citet{weiss2020global} made their code available as an R script to
allow for reproduction and extension of their analyses
(\url{https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt}
). To further enable extension of this work, here we have developed an R
package based on that code to seamlessly calculate the travel time from
any arbitrary set of locations.

Other R packages provide superficially similar though fundamentally
different functionality. A gaggle of R packages provide interfaces to
the \href{https://www.TravelTime.com}{TravelTime.com} API
\citep{traveltimeapi, traveltimeR, rtraveltime, traveltime_gh}. The
TravelTime.com platform provides travel time and routes between pairs of
locations, and `isochron' polygons --- areas reachable within a given
time from a given location. The isochron polygons are most comparable to
what \texttt{traveltime::calculate\_travel\_time()} calculates, though
each isochron polygon is a single polygon calculated is for a single
point location and specified maximum travel time, and the result
provides a maximum reachable extent for a specified time, rather than a
raster of gradation across a landscape for an arbitrary number of points
as in \texttt{traveltime}. TravelTime.com cannot provide a single result
surface for time to the nearest of a group of points, and continuous
time scale without extensive repeated iteration for all combinations of
time and points, plus additional calculation of the minimum value for
each cell from all points. Furthermore, TravelTime.com requires access
keys, a paid subscription beyond a short free period, and caps queries,
which add considerable friction to the use of this resource.

With \texttt{traveltime}, we provide free and open source software to
estimate travel time from any number of user-supplied locations, across
a complete area of interest, and with convenient access to motorised or
walking friction surfaces with global coverage.

Example: walking from public transport in Singapore

In this example we will calculate the walking travel time from the
nearest mass transit station across the island nation of Singapore ---
specifically Mass Rapid Transit (MRT) and Light Rail Transit (LRT)
stations --- and create a map of this.

\section{Prepare the data and friction
surface}\label{prepare-the-data-and-friction-surface}

For this exercise, we need two items of data:

\begin{itemize}
\tightlist
\item
  our area of interest --- in this case a map of Singapore, and
\item
  our points to calculate travel time from --- here the locations of
  Singapore's MRT and LRT stations.
\end{itemize}

We can download a national-level polygon boundary of Singapore from the
GADM \citep{gadm} database using the \texttt{geodata} package
\citep{geodata}. Here we download only the national boundary
(\texttt{level\ =\ 0}) and at a low resolution
(\texttt{resolution\ =\ 2}). Our boundary \texttt{singapore\_shapefile}
is a \texttt{SpatVector} class object.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(terra)}
\FunctionTok{library}\NormalTok{(geodata)}

\NormalTok{singapore\_shapefile }\OtherTok{\textless{}{-}} \FunctionTok{gadm}\NormalTok{(}
  \AttributeTok{country =} \StringTok{"Singapore"}\NormalTok{,}
  \AttributeTok{level =} \DecValTok{0}\NormalTok{,}
  \AttributeTok{path =} \FunctionTok{tempdir}\NormalTok{(),}
  \AttributeTok{resolution =} \DecValTok{2}
\NormalTok{)}

\NormalTok{singapore\_shapefile}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
 class       : SpatVector 
 geometry    : polygons 
 dimensions  : 1, 2  (geometries, attributes)
 extent      : 103.6091, 104.0858, 1.1664, 1.4714  (xmin, xmax, ymin, ymax)
 coord. ref. : lon/lat WGS 84 (EPSG:4326) 
 names       : GID_0   COUNTRY
 type        : <chr>     <chr>
 values      :   SGP Singapore
\end{verbatim}

The \texttt{stations} data set included in the \texttt{traveltime}
package is a 563 row, 2 column \texttt{matrix} containing the longitude
(\texttt{x}) and latitude (\texttt{y}) of all LRT and MRT station exits
in Singapore from \citet{singdata}:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(traveltime)}
\FunctionTok{head}\NormalTok{(stations)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
            x        y
[1,] 103.9091 1.334922
[2,] 103.9335 1.336555
[3,] 103.8493 1.297699
[4,] 103.8508 1.299195
[5,] 103.9094 1.335311
[6,] 103.9389 1.344999
\end{verbatim}

Now that we have the two items of data that we require, the next step is
to prepare a friction surface for our area of interest. We will use the
friction surface from \citet{weiss2020} that can be downloaded by
\texttt{traveltime} with the function \texttt{get\_friction\_surface()}.
We can pass in our basemap \texttt{singapore\_shapefile}, a
\texttt{SpatVector}, directly as the \texttt{extent}. We're interested
in walking time from a station, so we'll download the walking friction
surface by specifying \texttt{surface\ =\ "walk2020"}. (Alternatively,
we could use \texttt{surface\ =\ "motor2020"} for motorised travel if
that were of interest.) We're also only interested in walking \emph{on
land} so we then mask out areas outside of the land boundary in
\texttt{singapore\_shapefile}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{friction\_singapore }\OtherTok{\textless{}{-}} \FunctionTok{get\_friction\_surface}\NormalTok{(}
    \AttributeTok{surface =} \StringTok{"walk2020"}\NormalTok{,}
    \AttributeTok{extent =}\NormalTok{ singapore\_shapefile}
\NormalTok{  )}\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{mask}\NormalTok{(singapore\_shapefile)}
\end{Highlighting}
\end{Shaded}

Thus we have our friction surface as a \texttt{SpatRaster}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{friction\_singapore}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
class       : SpatRaster 
dimensions  : 37, 57, 1  (nrow, ncol, nlyr)
resolution  : 0.008333333, 0.008333333  (x, y)
extent      : 103.6083, 104.0833, 1.166667, 1.475  (xmin, xmax, ymin, ymax)
coord. ref. : lon/lat WGS 84 (EPSG:4326) 
source(s)   : memory
varname     : Accessibility__202001_Global_Walking_Only_Friction_Surface_1.1664,103.6091,1.4714,104.0858 
name        : friction_surface 
min value   :       0.01200000 
max value   :       0.06192715 
\end{verbatim}

Below we plot the friction surface raster \texttt{friction\_singapore},
with the vector boundary \texttt{singapore\_shapefile} as a grey line,
and \texttt{stations} as grey points (Figure Figure~\ref{fig-data}).
\texttt{traveltime} takes resistance values of friction
\citep{gdistance2017}, so higher values of friction indicate more time
travelling across a given cell. In this case walking friction is fairly
uniform across the islands barring a few small inaccesible locations.
More heterogeneity would be expected in less developed locations with
fewer transport corridors.

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-data-1.pdf}

}

\caption{\label{fig-data}Friction surface raster of Singapore, showing
Singapore boundary in grey, and station locations as grey points.}

\end{figure}%

\section{Calculate and plot the travel
time}\label{calculate-and-plot-the-travel-time}

With all the data collected, the function
\texttt{calculate\_travel\_time()} takes the friction surface
\texttt{friction\_singapore} and the points of interest in
\texttt{stations}, and returns a \texttt{SpatRaster} of walking time in
minutes to each cell from the nearest station:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{trave\_time\_singapore }\OtherTok{\textless{}{-}} \FunctionTok{calculate\_travel\_time}\NormalTok{(}
  \AttributeTok{friction\_surface =}\NormalTok{ friction\_singapore,}
  \AttributeTok{points =}\NormalTok{ stations}
\NormalTok{)}

\NormalTok{trave\_time\_singapore}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
class       : SpatRaster 
dimensions  : 37, 57, 1  (nrow, ncol, nlyr)
resolution  : 0.008333333, 0.008333333  (x, y)
extent      : 103.6083, 104.0833, 1.166667, 1.475  (xmin, xmax, ymin, ymax)
coord. ref. :  
source(s)   : memory
name        : travel_time 
min value   :           0 
max value   :         Inf 
\end{verbatim}

We present the resulting calculated travel times in Figure
Figure~\ref{fig-result} where, as expected, the travel times are lowest
near station exits (per Figure Figure~\ref{fig-data}) and progressively
higher further away. Note that the results in
\texttt{trave\_time\_singapore} include infinite values (\texttt{Inf}
above). In Figure Figure~\ref{fig-data}, the islands to the south and
north-east are shown as filled cells, but unconnected with the mainland.
The raster cells for these islands appear absent in Figure
Figure~\ref{fig-result}. Because they are not connected to any cells
with a station, the calculated travel time is infinite, and so these
cells do not appear in Figure Figure~\ref{fig-result}.

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-result-1.pdf}

}

\caption{\label{fig-result}Map of walking travel time in Singapore, in
minutes from nearest MRT or LRT station.}

\end{figure}%

Opportunities for future development

The \texttt{traveltime} package is immediately suitable to a range of
applications where travel to custom locations is of interest.
Nonetheless, we see opportunities to build the package utility into the
future through two mechanisms: (1) capability to better distribute a
wider range friction surfaces, and (2) additional methods to handle
large spatial extents.

Firstly, \texttt{traveltime} currently has access to walking and
motorised friction surfaces for 2020, both at 30 arc-second
resolution\footnote{Approximately 0.008333 decimal degrees, or just
  below 1 km\textsuperscript{2} at the equator}. Although the user can
supply their own friction surface, we expect most applications will use
these existing surfaces given the extensive work needed in creating new
ones \citep{weiss2018global, weiss2020global}. As landscapes are not
dynamic, it may be useful to incorporate updated versions of these
friction surfaces if and when they are available, though this is likely
to occur first through \texttt{malariaAtlas}. Furthermore, although the
resolution of these data is likely to be suitable for larger landscape
foci, higher resolution data may be helpful for more locally focussed
analyses. For instance, although the example here was chosen for it's
simplicity and low computational demands, a \textasciitilde1
km\textsuperscript{2} cell size is a relatively large area to walk
across, and thus actual waking times may vary significantly within each
cell. We underline however that a user can provide their own higher
resolution friction surface at present if desired.

At the other end of the scale, the calculations can require relatively
large amounts of onboard memory for analyses over large landscapes
(e.g.~analyses over Africa required \textasciitilde{} 72 GB RAM).
Developing methods to handle large landscapes either with less memory or
via cloud resources would be helpful to make such analyses accessible to
those without access to larger computing resources.

Acknowledgements

This work was supported, in whole or in part, by the Bill \& Melinda
Gates Foundation {[}INV-021972{]}. The conclusions and opinions
expressed in this work are those of the authors alone and shall not be
attributed to the Foundation. Under the grant conditions of the
Foundation, a Creative Commons Attribution 4.0 License has already been
assigned to the Author Accepted Manuscript version that might arise from
this submission. Please note works submitted as a preprint have not
undergone a peer review process.

References

  \bibliography{paper.bib}

\end{document}

### 2025-02-11 paper/paper.qmd
changes to internal referencing

@@
 format:
-  #md: default
-  html:
-    keep-md: true
-    fig-height: 4
-    fig-align: center
-    fig-format: png
-    dpi: 300
+  jasa-pdf:
+    keep-tex: true   
 tags: 
@@
 Now that we have the two items of data that we require, the next step is to prepare a friction surface for our area of interest.
-We will use the friction surface from @weiss2020 that can be downloaded by `traveltime` with the function `get_friction_surface()`.
+We will use the friction surface from @weiss2020global that can be downloaded by `traveltime` with the function `get_friction_surface()`.
 We can pass in our basemap `singapore_shapefile`, a `SpatVector`, directly as the `extent`.
@@
 
-Below we plot the friction surface raster `friction_singapore`, with the vector boundary `singapore_shapefile` as a grey line, and `stations` as grey points (Figure \ref{fig-data}).
+Below we plot the friction surface raster `friction_singapore`, with the vector boundary `singapore_shapefile` as a grey line, and `stations` as grey points (Figure @fig-data).
 `traveltime` takes resistance values of friction [@gdistance2017], so higher values of friction indicate more time travelling across a given cell.
@@
 
-We present the resulting calculated travel times in Figure \ref{fig-result} where, as expected, the travel times are lowest near station exits (per Figure \ref{fig-data}) and progressively higher further away.
+We present the resulting calculated travel times in Figure @fig-result where, as expected, the travel times are lowest near station exits (per Figure @fig-data) and progressively higher further away.
 Note that the results in `trave_time_singapore` include infinite values (`Inf`  above).
-In Figure \ref{fig-data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
-The raster cells for these islands appear absent in Figure \ref{fig-result}.
-Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig-result}.
+In Figure @fig-data, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
+The raster cells for these islands appear absent in Figure @fig-result.
+Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure @fig-result.
 

### 2025-02-11 paper/paper.qmd
use PLoS template which worx

@@
 format:
-  jasa-pdf:
+  plos-pdf:
     keep-tex: true   
@@
     corresponding: true
-    affiliation: "1, 2"
+    affiliation: 1, 2
   - name: Nicholas Tierney
     orcid: 0000-0003-1460-8722
-    affiliation: "1, 3"
+    affiliation: 1, 3
   - name: Nick Golding
     orcid: 0000-0001-8916-5570
-    affiliation: "1, 4"
+    affiliation: 1, 4
   - name: Daniel J. Weiss
     orcid: 0000-0002-6175-5648
-    affiliation: "1, 3"
+    affiliation: 1, 3
 affiliations:
@@
 
-[^2]: Approximately 0.008333 decimal degrees, or just below 1 km^2^ at the equator
+[^2]: Approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator
 
 At the other end of the scale, the calculations can require relatively large amounts of onboard memory for analyses over large landscapes (e.g.
-analyses over Africa required \~ 72 GB RAM).
+analyses over Africa required ~ 72 GB RAM).
 Developing methods to handle large landscapes either with less memory or via cloud resources would be helpful to make such analyses accessible to those without access to larger computing resources.

### 2025-02-11 paper/paper.tex
use PLoS template which worx

@@
 \PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
 %
+
+\PassOptionsToPackage{table}{xcolor}
+
 \documentclass[
-  12pt]{article}
+  10pt,
+  letterpaper,
+]{article}
 
@@
 \usepackage{xcolor}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
-% Make \paragraph and \subparagraph free-standing
-\makeatletter
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}{
-    \@ifstar
-      \xxxParagraphStar
-      \xxxParagraphNoStar
-  }
-  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
-  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}{
-    \@ifstar
-      \xxxSubParagraphStar
-      \xxxSubParagraphNoStar
-  }
-  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
-  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-\makeatother
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 
@@
 
-\addtolength{\oddsidemargin}{-.5in}%
-\addtolength{\evensidemargin}{-1in}%
-\addtolength{\textwidth}{1in}%
-\addtolength{\textheight}{1.7in}%
-\addtolength{\topmargin}{-1in}%
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% marvosym package for additional characters
+\usepackage{marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+% Using natbib instead
+% \usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr}
+\usepackage{epstopdf}
+%\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+%\setlength{\headheight}{27.023pt}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\today}
 \makeatletter
@@
 \fi
-\usepackage[]{natbib}
-\bibliographystyle{agsm}
+\usepackage[numbers,square,comma]{natbib}
+\bibliographystyle{plos2015}
 \usepackage{bookmark}
@@
 \hypersetup{
-  pdftitle={Summary},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
+  pdftitle={traveltime: an R package to calculate walking and motorised travel time from user-specified locations},
+  hidelinks,
   pdfcreator={LaTeX via pandoc}}
@@
 \begin{document}
+\vspace*{0.2in}
 
-\def\spacingset#1{\renewcommand{\baselinestretch}%
-{#1}\small\normalsize} \spacingset{1}
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{\texttt{traveltime}: an R package to calculate walking
+and motorised travel time from user-specified
+locations} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+\newline
+\\
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+Gerard E. Ryan\textsuperscript{1*}, Nicholas
+Tierney\textsuperscript{2}, Nick Golding\textsuperscript{3}, Daniel J.
+Weiss\textsuperscript{2}
+\\
+\bigskip
+\textbf{1} 1, 2, \\ \textbf{2} 1, 3, \\ \textbf{3} 1,
+4, \\ \textbf{4} The Kids Research Institute Australia, Nedlands 6009
+WA, Australia, \\ \textbf{5} Melbourne School of Population and Global
+Health, University of Melbourne, 3010, VIC,
+Australia, \\ \textbf{6} Curtin University, Bentley, WA,
+Australia, \\ \textbf{7} University of Western Australia, WA,
+Australia, 
+\bigskip
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
+%
+% Remove or comment out the author notes below if they aren't used.
+%
+% Primary Equal Contribution Note
+%\Yinyang These authors contributed equally to this work.
 
-\date{January 1, 2025}
-\title{\bf Summary}
-\author{
-Gerard E. Ryan\\
-1, 2\\
-and\\Nicholas Tierney\\
-1, 3\\
-and\\Nick Golding\\
-1, 4\\
-and\\Daniel J. Weiss\\
-1, 3\\
-}
-\maketitle
+% Additional Equal Contribution Note
+% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
+%\ddag These authors also contributed equally to this work.
 
-\bigskip
-\bigskip
-\begin{abstract}
+% Current address notes
+%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+% \textcurrency b Insert second current address
+% \textcurrency c Insert third current address
 
-\end{abstract}
+% Deceased author note
+%\dag Deceased
 
-\newpage
-\spacingset{1.9} % DON'T change the spacing!
+% Group/Consortium Author Note
+%\textpilcrow Membership list can be found in the Acknowledgments section.
+
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* \\ email \href{mailto:gerry.ryan@thekids.org.au}{gerry.ryan@thekids.org.au}
+
+\end{flushleft}
+
+\linenumbers
+
+\section{Summary}\label{summary}
 
@@
 
-Statement of need
+\section{Statement of need}\label{statement-of-need}
 
@@
 
-Example: walking from public transport in Singapore
+\section{Example: walking from public transport in
+Singapore}\label{example-walking-from-public-transport-in-singapore}
 
@@
 
-\section{Prepare the data and friction
+\subsection{Prepare the data and friction
 surface}\label{prepare-the-data-and-friction-surface}
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(terra)}
-\FunctionTok{library}\NormalTok{(geodata)}
-
-\NormalTok{singapore\_shapefile }\OtherTok{\textless{}{-}} \FunctionTok{gadm}\NormalTok{(}
-  \AttributeTok{country =} \StringTok{"Singapore"}\NormalTok{,}
-  \AttributeTok{level =} \DecValTok{0}\NormalTok{,}
-  \AttributeTok{path =} \FunctionTok{tempdir}\NormalTok{(),}
-  \AttributeTok{resolution =} \DecValTok{2}
-\NormalTok{)}
-
-\NormalTok{singapore\_shapefile}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{verbatim}
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(traveltime)}
-\FunctionTok{head}\NormalTok{(stations)}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{verbatim}
@@
 to prepare a friction surface for our area of interest. We will use the
-friction surface from \citet{weiss2020} that can be downloaded by
+friction surface from \citet{weiss2020global} that can be downloaded by
 \texttt{traveltime} with the function \texttt{get\_friction\_surface()}.
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{friction\_singapore}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{verbatim}
@@
 with the vector boundary \texttt{singapore\_shapefile} as a grey line,
-and \texttt{stations} as grey points (Figure Figure~\ref{fig-data}).
+and \texttt{stations} as grey points (Figure Fig~\ref{fig-data}).
 \texttt{traveltime} takes resistance values of friction
@@
 
-\section{Calculate and plot the travel
+\subsection{Calculate and plot the travel
 time}\label{calculate-and-plot-the-travel-time}
@@
 minutes to each cell from the nearest station:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{trave\_time\_singapore }\OtherTok{\textless{}{-}} \FunctionTok{calculate\_travel\_time}\NormalTok{(}
-  \AttributeTok{friction\_surface =}\NormalTok{ friction\_singapore,}
-  \AttributeTok{points =}\NormalTok{ stations}
-\NormalTok{)}
-
-\NormalTok{trave\_time\_singapore}
-\end{Highlighting}
-\end{Shaded}
 
@@
 We present the resulting calculated travel times in Figure
-Figure~\ref{fig-result} where, as expected, the travel times are lowest
-near station exits (per Figure Figure~\ref{fig-data}) and progressively
+Fig~\ref{fig-result} where, as expected, the travel times are lowest
+near station exits (per Figure Fig~\ref{fig-data}) and progressively
 higher further away. Note that the results in
 \texttt{trave\_time\_singapore} include infinite values (\texttt{Inf}
-above). In Figure Figure~\ref{fig-data}, the islands to the south and
+above). In Figure Fig~\ref{fig-data}, the islands to the south and
 north-east are shown as filled cells, but unconnected with the mainland.
 The raster cells for these islands appear absent in Figure
-Figure~\ref{fig-result}. Because they are not connected to any cells
-with a station, the calculated travel time is infinite, and so these
-cells do not appear in Figure Figure~\ref{fig-result}.
+Fig~\ref{fig-result}. Because they are not connected to any cells with a
+station, the calculated travel time is infinite, and so these cells do
+not appear in Figure Fig~\ref{fig-result}.
 
@@
 
-Opportunities for future development
+\section{Opportunities for future
+development}\label{opportunities-for-future-development}
 
@@
 resolution\footnote{Approximately 0.008333 decimal degrees, or just
-  below 1 km\textsuperscript{2} at the equator}. Although the user can
-supply their own friction surface, we expect most applications will use
-these existing surfaces given the extensive work needed in creating new
-ones \citep{weiss2018global, weiss2020global}. As landscapes are not
-dynamic, it may be useful to incorporate updated versions of these
-friction surfaces if and when they are available, though this is likely
-to occur first through \texttt{malariaAtlas}. Furthermore, although the
+  below 1 km\(^2\) at the equator}. Although the user can supply their
+own friction surface, we expect most applications will use these
+existing surfaces given the extensive work needed in creating new ones
+\citep{weiss2018global, weiss2020global}. As landscapes are not dynamic,
+it may be useful to incorporate updated versions of these friction
+surfaces if and when they are available, though this is likely to occur
+first through \texttt{malariaAtlas}. Furthermore, although the
 resolution of these data is likely to be suitable for larger landscape
@@
 
-Acknowledgements
+\section{Acknowledgements}\label{acknowledgements}
 
@@
 
-References
-
+\nolinenumbers
+\renewcommand\refname{References}
   \bibliography{paper.bib}

### 2025-02-11 paper/paper.qmd
fix affiliations for pdf to render them properly

@@
     corresponding: true
-    affiliation: 1, 2
+    affiliations:
+      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
+      - Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
   - name: Nicholas Tierney
     orcid: 0000-0003-1460-8722
-    affiliation: 1, 3
+    affiliations:
+      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
+      - Curtin University, Bentley, WA, Australia
   - name: Nick Golding
     orcid: 0000-0001-8916-5570
-    affiliation: 1, 4
+    affiliations:
+      - University of Western Australia, WA, Australia
+      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
   - name: Daniel J. Weiss
     orcid: 0000-0002-6175-5648
-    affiliation: 1, 3
     affiliations:
- - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
-   index: 1
- - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
-   index: 2
- - name: Curtin University, Bentley, WA, Australia
-   index: 3
- - name: University of Western Australia, WA, Australia
-   index: 4
+      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
+      - Curtin University, Bentley, WA, Australia
 date: In prep 2025

### 2025-02-11 paper/paper.tex
fix affiliations for pdf to render them properly

@@
 % Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-Gerard E. Ryan\textsuperscript{1*}, Nicholas
-Tierney\textsuperscript{2}, Nick Golding\textsuperscript{3}, Daniel J.
-Weiss\textsuperscript{2}
+Gerard E. Ryan\textsuperscript{1,2*}, Nicholas
+Tierney\textsuperscript{1,3}, Nick Golding\textsuperscript{4,1}, Daniel
+J. Weiss\textsuperscript{1,3}
 \\
 \bigskip
-\textbf{1} 1, 2, \\ \textbf{2} 1, 3, \\ \textbf{3} 1,
-4, \\ \textbf{4} The Kids Research Institute Australia, Nedlands 6009
-WA, Australia, \\ \textbf{5} Melbourne School of Population and Global
+\textbf{1} The Kids Research Institute Australia, Nedlands 6009 WA,
+Australia, \\ \textbf{2} Melbourne School of Population and Global
 Health, University of Melbourne, 3010, VIC,
-Australia, \\ \textbf{6} Curtin University, Bentley, WA,
-Australia, \\ \textbf{7} University of Western Australia, WA,
+Australia, \\ \textbf{3} Curtin University, Bentley, WA,
+Australia, \\ \textbf{4} University of Western Australia, WA,
 Australia, 

### 2025-02-12 paper/paper.qmd
this version successfully submitted to arXiv NB needed manual intervention: - commenting out of line number funtionality (package and commands) - in text author citations (e.g. Weiss et al. (2020), rather than (Weiss et al. 2020) needed edits in `paper.tex` changing `\citet{authoryear}` to `Author et al. \citep{authoryear}`

@@
 ---
-title: "`traveltime`: an R package to calculate walking and motorised travel time from user-specified locations"
+title: "`traveltime`: an R package to calculate travel time across a landscape from user-specified locations"
 format:
@@
     keep-tex: true
+    fig-height: 4
+    fig-align: center
+    fig-format: png
+    dpi: 300   
 tags: 
@@
 The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations.
-We envisage this software having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias, @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], allocating public defibrillators [@tierney2018novel], setting rehabilitation districts for stroke patients [@padgham2019introduction], or understanding access to agricultural processing facilities [@zhao2023replanting].
+We envisage this software having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias; @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], allocating public defibrillators [@tierney2018novel], setting rehabilitation districts for stroke patients [@padgham2019introduction], or understanding access to agricultural processing facilities [@zhao2023replanting].
 

### 2025-02-12 paper/paper.tex
this version successfully submitted to arXiv NB needed manual intervention: - commenting out of line number funtionality (package and commands) - in text author citations (e.g. Weiss et al. (2020), rather than (Weiss et al. 2020) needed edits in `paper.tex` changing `\citet{authoryear}` to `Author et al. \citep{authoryear}`

@@
 % line numbers
-\usepackage[right]{lineno}
+% \usepackage[right]{lineno}
 
@@
 \hypersetup{
-  pdftitle={traveltime: an R package to calculate walking and motorised travel time from user-specified locations},
+  pdftitle={traveltime: an R package to calculate travel time across a landscape from user-specified locations},
   hidelinks,
@@
 {\Large
-\textbf\newline{\texttt{traveltime}: an R package to calculate walking
-and motorised travel time from user-specified
+\textbf\newline{\texttt{traveltime}: an R package to calculate travel
+time across a landscape from user-specified
 locations} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
@@
 
-\linenumbers
+% \linenumbers
 
@@
 having diverse applications including: estimating sampling bias in
-species occurrence data \citep[@reddy2003geographical]{dennis2000bias},
+species occurrence data \citep{dennis2000bias, reddy2003geographical},
 mapping electric vehicle charger accessibility
@@
 provide easy access to the existing friction surfaces generated by
-\citet{weiss2020global}, \texttt{traveltime} uses the package
+Weiss et al. \citep{weiss2020global}, \texttt{traveltime} uses the package
 \texttt{malariaAtlas} to allow users to access friction surfaces for the
@@
   Google Scholar at the 28th of January 2025.}, and the city data set of
-\citet{nelson2019suite} is available to R users through the widely-used
+Nelson et al. \citep{nelson2019suite} is available to R users through the widely-used
 \texttt{geodata} package \citep{geodata}. There is clear demand for
@@
 
-\citet{weiss2020global} made their code available as an R script to
+Weiss et al. \citep{weiss2020global} made their code available as an R script to
 allow for reproduction and extension of their analyses
@@
 (\texttt{x}) and latitude (\texttt{y}) of all LRT and MRT station exits
-in Singapore from \citet{singdata}:
+in Singapore from Land Transport Authority \citep{singdata}:
 
@@
 to prepare a friction surface for our area of interest. We will use the
-friction surface from \citet{weiss2020global} that can be downloaded by
+friction surface from Weiss et al. \citep{weiss2020global} that can be downloaded by
 \texttt{traveltime} with the function \texttt{get\_friction\_surface()}.
@@
 
-\includegraphics{paper_files/figure-pdf/fig-data-1.pdf}
+\includegraphics{paper_files/figure-pdf/fig-data-1.png}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-result-1.pdf}
+\includegraphics{paper_files/figure-pdf/fig-result-1.png}
 
@@
 
-\nolinenumbers
+% \nolinenumbers
 \renewcommand\refname{References}

### 2025-02-12 paper/paper.qmd
resolving #50 for this branch

@@
 ---
-title: "`traveltime`: an R package to calculate travel time across a landscape from user-specified locations"
+title: "`traveltime`: an R package to calculate walking and motorised travel time from user-specified locations"
 format:
-  plos-pdf:
-    keep-tex: true
+  #md: default
+  html:
+    keep-md: true
     fig-height: 4
@@
     corresponding: true
-    affiliations:
-      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
-      - Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
+    affiliation: "1, 2"
   - name: Nicholas Tierney
     orcid: 0000-0003-1460-8722
-    affiliations:
-      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
-      - Curtin University, Bentley, WA, Australia
+    affiliation: "1, 3"
   - name: Nick Golding
     orcid: 0000-0001-8916-5570
-    affiliations:
-      - University of Western Australia, WA, Australia
-      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
+    affiliation: "1, 4"
   - name: Daniel J. Weiss
     orcid: 0000-0002-6175-5648
+    affiliation: "1, 3"
 affiliations:
-      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
-      - Curtin University, Bentley, WA, Australia
+ - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
+   index: 1
+ - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
+   index: 2
+ - name: Curtin University, Bentley, WA, Australia
+   index: 3
+ - name: University of Western Australia, WA, Australia
+   index: 4
 date: In prep 2025
@@
 
-Below we plot the friction surface raster `friction_singapore`, with the vector boundary `singapore_shapefile` as a grey line, and `stations` as grey points (Figure @fig-data).
+Below we plot the friction surface raster `friction_singapore`, with the vector boundary `singapore_shapefile` as a grey line, and `stations` as grey points (Figure \ref{fig-data}).
 `traveltime` takes resistance values of friction [@gdistance2017], so higher values of friction indicate more time travelling across a given cell.
@@
 
-We present the resulting calculated travel times in Figure @fig-result where, as expected, the travel times are lowest near station exits (per Figure @fig-data) and progressively higher further away.
+We present the resulting calculated travel times in Figure \ref{fig-result} where, as expected, the travel times are lowest near station exits (per Figure \ref{fig-data}) and progressively higher further away.
 Note that the results in `trave_time_singapore` include infinite values (`Inf`  above).
-In Figure @fig-data, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
-The raster cells for these islands appear absent in Figure @fig-result.
-Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure @fig-result.
+In Figure \ref{fig-data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
+The raster cells for these islands appear absent in Figure \ref{fig-result}.
+Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig-result}.
 
@@
 
-[^2]: Approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator
+[^2]: Approximately 0.008333 decimal degrees, or just below 1 km^2^ at the equator
 
 At the other end of the scale, the calculations can require relatively large amounts of onboard memory for analyses over large landscapes (e.g.
-analyses over Africa required ~ 72 GB RAM).
+analyses over Africa required \~ 72 GB RAM).
 Developing methods to handle large landscapes either with less memory or via cloud resources would be helpful to make such analyses accessible to those without access to larger computing resources.

### 2025-02-13 paper/paper.qmd
Add hard code author names into `.qmd` for in-text named citations so it renders properly in `.tex`

--- 2025-02-12 paper/paper.qmd
@@
 The travel time is calculated as movement over a resistance 'friction surface' [@gdistance2017].
-To provide easy access to the existing friction surfaces generated by @weiss2020global, `traveltime` uses the package `malariaAtlas` to allow users to access friction surfaces for the area of interest; though users can also supply any other friction surfaces to `traveltime`.
+To provide easy access to the existing friction surfaces generated by Weiss et al.
+[@weiss2020global], `traveltime` uses the package `malariaAtlas` to allow users to access friction surfaces for the area of interest; though users can also supply any other friction surfaces to `traveltime`.
 Functionality from `gdistance` [@gdistance2017] is also used internally to calculate the minimum least-cost-distance for each cell from the points of interest.
@@
 
-Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] have generated significant interest and use[^1], and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata].
+Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] have generated significant interest and use[^1], and the city data set of Nelson et al.
+[@nelson2019suite] is available to R users through the widely-used `geodata` package [@geodata].
 There is clear demand for these type of products.
@@
 
-@weiss2020global made their code available as an R script to allow for reproduction and extension of their analyses (<https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt> ).
+Weiss et al.
+[@weiss2020global] made their code available as an R script to allow for reproduction and extension of their analyses (<https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt> ).
 To further enable extension of this work, here we have developed an R package based on that code to seamlessly calculate the travel time from any arbitrary set of locations.
@@
 ```{r}
+#| echo: true
 #| label: get-basemap
@@
 
-The `stations` data set included in the `traveltime` package is a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from @singdata:
+The `stations` data set included in the `traveltime` package is a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from Land Transport Authority @singdata:
 
@@
 Now that we have the two items of data that we require, the next step is to prepare a friction surface for our area of interest.
-We will use the friction surface from @weiss2020global that can be downloaded by `traveltime` with the function `get_friction_surface()`.
+We will use the friction surface from Weiss et al.
+[@weiss2020global] that can be downloaded by `traveltime` with the function `get_friction_surface()`.
 We can pass in our basemap `singapore_shapefile`, a `SpatVector`, directly as the `extent`.

### 2025-02-13 paper/paper.tex
Add hard code author names into `.qmd` for in-text named citations so it renders properly in `.tex`

@@
 movement over a resistance `friction surface' \citep{gdistance2017}. To
-provide easy access to the existing friction surfaces generated by
-Weiss et al. \citep{weiss2020global}, \texttt{traveltime} uses the package
+provide easy access to the existing friction surfaces generated by Weiss
+et al. \citep{weiss2020global}, \texttt{traveltime} uses the package
 \texttt{malariaAtlas} to allow users to access friction surfaces for the
@@
   Google Scholar at the 28th of January 2025.}, and the city data set of
-Nelson et al. \citep{nelson2019suite} is available to R users through the widely-used
-\texttt{geodata} package \citep{geodata}. There is clear demand for
-these type of products.
+Nelson et al. \citep{nelson2019suite} is available to R users through
+the widely-used \texttt{geodata} package \citep{geodata}. There is clear
+demand for these type of products.
 
-Weiss et al. \citep{weiss2020global} made their code available as an R script to
-allow for reproduction and extension of their analyses
+Weiss et al. \citep{weiss2020global} made their code available as an R
+script to allow for reproduction and extension of their analyses
 (\url{https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt}
@@
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(terra)}
+\FunctionTok{library}\NormalTok{(geodata)}
+
+\NormalTok{singapore\_shapefile }\OtherTok{\textless{}{-}} \FunctionTok{gadm}\NormalTok{(}
+  \AttributeTok{country =} \StringTok{"Singapore"}\NormalTok{,}
+  \AttributeTok{level =} \DecValTok{0}\NormalTok{,}
+  \AttributeTok{path =} \FunctionTok{tempdir}\NormalTok{(),}
+  \AttributeTok{resolution =} \DecValTok{2}
+\NormalTok{)}
+
+\NormalTok{singapore\_shapefile}
+\end{Highlighting}
+\end{Shaded}
+
 \begin{verbatim}
@@
 (\texttt{x}) and latitude (\texttt{y}) of all LRT and MRT station exits
-in Singapore from Land Transport Authority \citep{singdata}:
+in Singapore from Land Transport Authority \citet{singdata}:
 
@@
 to prepare a friction surface for our area of interest. We will use the
-friction surface from Weiss et al. \citep{weiss2020global} that can be downloaded by
-\texttt{traveltime} with the function \texttt{get\_friction\_surface()}.
-We can pass in our basemap \texttt{singapore\_shapefile}, a
-\texttt{SpatVector}, directly as the \texttt{extent}. We're interested
-in walking time from a station, so we'll download the walking friction
-surface by specifying \texttt{surface\ =\ "walk2020"}. (Alternatively,
-we could use \texttt{surface\ =\ "motor2020"} for motorised travel if
-that were of interest.) We're also only interested in walking \emph{on
-land} so we then mask out areas outside of the land boundary in
+friction surface from Weiss et al. \citep{weiss2020global} that can be
+downloaded by \texttt{traveltime} with the function
+\texttt{get\_friction\_surface()}. We can pass in our basemap
+\texttt{singapore\_shapefile}, a \texttt{SpatVector}, directly as the
+\texttt{extent}. We're interested in walking time from a station, so
+we'll download the walking friction surface by specifying
+\texttt{surface\ =\ "walk2020"}. (Alternatively, we could use
+\texttt{surface\ =\ "motor2020"} for motorised travel if that were of
+interest.) We're also only interested in walking \emph{on land} so we
+then mask out areas outside of the land boundary in
 \texttt{singapore\_shapefile}:

### 2025-02-13 paper/paper.qmd
make code echo and fix referencing of in-text author citations. For preprint branch only

@@
 
-The `stations` data set included in the `traveltime` package is a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from Land Transport Authority @singdata:
+The `stations` data set included in the `traveltime` package is a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from Land Transport Authority [@singdata]:
 
 ```{r}
+#| echo: true
 #| label: show-stations
@@
 ```{r }
+#| echo: true
 #| label: show-friction-surface-raster
@@
 ```{r}
+#| echo: true
 #| label: calculate-result

### 2025-02-13 paper/paper.tex
make code echo and fix referencing of in-text author citations. For preprint branch only

@@
 (\texttt{x}) and latitude (\texttt{y}) of all LRT and MRT station exits
-in Singapore from Land Transport Authority \citet{singdata}:
+in Singapore from Land Transport Authority \citep{singdata}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(traveltime)}
+\FunctionTok{head}\NormalTok{(stations)}
+\end{Highlighting}
+\end{Shaded}
 
@@
 Thus we have our friction surface as a \texttt{SpatRaster}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{friction\_singapore}
+\end{Highlighting}
+\end{Shaded}
 
@@
 minutes to each cell from the nearest station:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{trave\_time\_singapore }\OtherTok{\textless{}{-}} \FunctionTok{calculate\_travel\_time}\NormalTok{(}
+  \AttributeTok{friction\_surface =}\NormalTok{ friction\_singapore,}
+  \AttributeTok{points =}\NormalTok{ stations}
+\NormalTok{)}
+
+\NormalTok{trave\_time\_singapore}
+\end{Highlighting}
+\end{Shaded}
 

### 2025-02-14 README.Rmd
update readme to align with paper per #36

@@
   fig.path = "man/figures/README-",
-  out.width = "100%",
+  out.width = "80%",
   warning = NA
@@
 ![GitHub License](https://img.shields.io/github/license/geryan/traveltime)
-[![Lifecycle:](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
 ![GitHub commit activity](https://img.shields.io/github/commit-activity/w/geryan/traveltime)
-![GitHub Downloads (all assets, all releases)](https://img.shields.io/github/downloads/geryan/traveltime/total)
 ![GitHub last commit](https://img.shields.io/github/last-commit/geryan/traveltime)
@@
 [![Codecov test coverage](https://codecov.io/gh/idem-lab/traveltime/graph/badge.svg)](https://app.codecov.io/gh/idem-lab/traveltime)
+[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
 <!-- badges: end -->
 
-Implements methods from Weiss et al. 2018, 2020 to calculate travel time from
-given locations over a friction surface.
+`traveltime` enables a user to create a map of travel time over an area of interest from a user-specified set of geographic coordinates and friction surface.
+The package provides convenient access to global friction surfaces for walking and motorised travel for the year 2020.
+The final result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations.
 
-Citations:
-
-*D. J. Weiss, A. Nelson, C. A. Vargas-Ruiz, K. Gligoric, S., Bavadekar, E.
-Gabrilovich, A. Bertozzi-Villa, J. Rozier, H. S. Gibson, T., Shekel, C. Kamath,
-A. Lieber, K. Schulman, Y. Shao, V. Qarkaxhija, A. K. Nandi, S. H. Keddie, S.
-Rumisha, P. Amratia, R. Arambepola, E. G. Chestnutt, J. J. Millar, T. L. Symons,
-E. Cameron, K. E. Battle, S. Bhatt, and P. W. Gething.* **Global maps of travel
-time to healthcare facilities.** (2020) Nature Medicine.
-<https://doi.org/10.1038/s41591-020-1059-1>
-
-*D. J. Weiss, A. Nelson, H.S. Gibson, W. Temperley, S. Peedell, A. Lieber, M.
-Hancher, E. Poyart, S. Belchior, N. Fullman, B. Mappin, U. Dalrymple, J. Rozier,
-T.C.D. Lucas, R.E. Howes, L.S. Tusting, S.Y. Kang, E. Cameron, D. Bisanzio, K.E.
-Battle, S. Bhatt, and P.W. Gething.* **A global map of travel time to cities to
-assess inequalities in accessibility in 2015.** (2018). Nature.
-<https://doi.org/10.1038/nature25181>
+![Walking travel time from rail transport in Singapore](man/figures/README-fig-result-1.png){width=60% fig-align="right"}
 
 ## Installation
-
-You can install `traveltime` with:
 
@@
 ```
-
-## Let's calculate some travel times
-
-First download a friction surface --— here we are using the motorised travel
-time from Weiss *et al.* 2020. We use the function
-`traveltime::get_friction_surface`, specify the surface (type) as `"motor2020"`,
- and provide the spatial extent of interest:
-```{r }
-library(traveltime)
-library(terra)
-
-friction_surface <- get_friction_surface(
-    surface = "motor2020",
-    extent = c(111,112,0,1)
-  )
-friction_surface
+or 
+```r
+remotes::install_github("idem-lab/traveltime")
 ```
 
-Let's have a look at that `SpatRaster`:
+## Documentation
+
+### Website
+
+<https://idem-lab.github.io/traveltime/>
+
+### Paper
+
+Ryan, G.E., Tierney, N., Golding, N., and Weiss, D.J (2025).
+[traveltime: an R package to calculate travel time across a landscape from user-specified locations](https://doi.org/10.31223/X56M74){target="_blank"}.
+**EarthArXiv** 10.31223/X56M74
+
+## What does this thing do?
+
+The `traveltime` workflow starts with:
+
+- a set of points you are interested in,
+
+and either
+
+- you supply a friction surface for the area you are interested in, or 
+- you supply your area of interest and use `get_friction_surface` to retrieve a pre-prepared walking or motorised travel friction surface from Weiss et al.
+(2020) [^1] --- this will probably be the case in most applications.
+
+Then, running `calculate_travel_time` produces a raster as a `terra` `SpatRaster` with the travel time to the (temporally) nearest of the supplied points over the area of interest.
+
+### A practical example: walking from public transport in Singapore
+
+Here we will calculate the walking travel time from the nearest mass transit station across the island nation of Singapore --- specifically Mass Rapid Transit (MRT) and Light Rail Transit (LRT) stations --- and create a map of this.
+
+#### Prepare the data
+
+For this exercise, we need two items of data:
+
+-   our points to calculate travel time from --- here the locations of Singapore's MRT and LRT stations, and
+-   our area of interest --- in this case a map of Singapore.
+
+##### Points
+
+Our points of interest will be the `stations` data set included in `traveltime`; a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from [^2]:
+
 ```{r}
-plot(friction_surface)
+#| label: show-stations
+library(traveltime)
+head(stations)
 ```
 
-Now, prepare points that we would like to calculate travel time from:
-```{r}
-from_here <- tibble::tibble(
-  x = c(111.2, 111.9),
-  y = c(0.2, 0.35)
-)
-from_here
-```
+##### Area of interest
 
-And calculate the travel time from our points `from_here` over the friction 
-surface `friction_surface` using the function
-`traveltime::calculate_travel_time`:
-```{r}
-travel_time <- calculate_travel_time(
-  friction_surface = friction_surface,
-  points = from_here
-)
-travel_time
-```
-
-Et voila! Here is the motorised travel time in minutes for each cell, with our
-points in pink.
-```{r}
-plot(travel_time)
-points(from_here, pch = 19, col = "hotpink")
-```
-
-## A more tangible example: Walking in Singapore
-
-Let's create a map of the walking time across the island of Singapore from the 
-nearest
-[MRT or LRT](https://en.wikipedia.org/wiki/Transport_in_Singapore#Rail_transport)
-station.
-
-To do this, we need:
-
-- a map of Singapore
-- locations of the stations
-
-Here's our basemap via `geodata`:
+To obtain our area of interest, we download a national-level polygon boundary of Singapore using the `geodata` package.
+Here we download only the national boundary (`level = 0`) at a low resolution (`resolution = 2`).
+Our boundary `singapore_shapefile` is a `SpatVector` class object from the package `terra`.
 
 ```{r}
-#install.packages("geodata")
+#| label: get-basemap
+library(terra)
 library(geodata)
-sin <- gadm(
+
+singapore_shapefile <- gadm(
   country = "Singapore",
@@
 )
-plot(sin)
+
+singapore_shapefile
 ```
 
-We're going to see how long it takes to walk home from a station, so we'll
-download the walking-only friction surface this time by specifying 
-`surface = "walk2020`.
+##### Friction surface
 
-We can also pass in our basemap `sin`, a `SpatVector`, directly as the `extent`,
-instead of specifying by hand as above. We're also only interested in walking 
-*on land* so we mask out areas outside of `sin`, that are within the extent of 
-the raster:
+Now that we have the two items of data that we require initially, the next step is to prepare a friction surface for our area of interest.
+
+We will use the friction surface from Weiss et al (2020) that can be downloaded by `traveltime` with the function `get_friction_surface()`.
+This function takes extents in a variety of formats and returns the surface for that extent only.
+
+We can pass in our basemap `singapore_shapefile`, a `SpatVector`, directly as the `extent`.
+We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`.
+
+(Alternatively, we could use `surface = "motor2020"` for motorised travel).
+
+We're only interested in walking *on land*, so we then mask out areas outside of the land boundary of `singapore_shapefile`:
+
+```{r}
+#| label: download-friction-surface
+#| echo: true
+#| output: false
+friction_singapore <- get_friction_surface(
+    surface = "walk2020",
+    extent = singapore_shapefile
+  )|> 
+  mask(singapore_shapefile)
+```
+
+Thus we have our friction surface as a `SpatRaster`:
 
 ```{r }
-library(traveltime)
-library(terra)
-
-friction_singapore <- get_friction_surface(
-    surface = "walk2020",
-    extent = sin
-  )|> 
-  mask(sin)
-
+#| label: show-friction-surface-raster
+#| echo: true
 friction_singapore
@@
 
-The the `stations` data set in this package contains the longitude and latitude
-of all LRT and MRT station exits in Singapore[^1].
+##### Input data
+
+Below we plot the friction surface raster `friction_singapore`, with the vector boundary `singapore_shapefile` as a grey line, and `stations` as grey points.
+`traveltime` takes resistance values of friction (see paper for more details), so higher values of friction indicate more time travelling across a given cell.
 
 ```{r}
-head(stations)
-```
-
-Let's look at our data now.
-
-Below we plot the friction surface raster `friction_singapore`, with the vector
-boundary of `sin` as a dashed grey line, and `stations` as grey points:
-
-```{r}
+#| label: fig-data
+#| echo: true
+#| fig-cap: "Friction surface raster of Singapore, showing Singapore boundary
+#|   in grey, and station locations as grey points."
 library(tidyterra)
@@
   geom_spatvector(
-    data = sin,
+    data = singapore_shapefile,
     fill = "transparent",
-    col = "grey50",
-    lty = 2
+    col = "grey50"
   ) +
@@
   labs(
-    fill = "Friction"
-  )
-
+    fill = "Resistance",
+    x = element_blank(),
+    y = element_blank()
+  ) +
+  theme_minimal()
 ```
 
-OK, now we want to calculate the walking travel time in minutes across the 
-friction surface from the nearest station exit:
+#### Calculate travel time
+
+With all the data collected, the function `calculate_travel_time()` takes the friction surface `friction_singapore` and the points of interest in `stations`, and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station:
 
 ```{r}
-travel_time_sin <- calculate_travel_time(
+#| label: calculate-result
+#| echo: true
+trave_time_singapore <- calculate_travel_time(
   friction_surface = friction_singapore,
@@
 )
-travel_time_sin
+
+trave_time_singapore
 ```
 
-Et voi*lah* --- a raster of walking time from the nearest station.
+##### Plot results
+
+We present the resulting calculated travel times below where (as expected) the travel times are lowest near station exits and progressively higher further away.
+Note that the results in `trave_time_singapore` include infinite values (`Inf`  above).
+The islands to the south and north-east are shown as filled cells in the figure above, i.e., they are not masked out by `singapore_shapefile`.
+But because those islands they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in the figure below.
+
 ```{r}
-contour(
-  x = travel_time_sin,
-  filled = TRUE,
-  nlevels = 20,
-  col = viridis::magma(19, direction = -1)
+#| label: fig-result
+#| echo: true
+#| fig-cap: "Map of walking travel time in Singapore, in minutes from 
+#|   nearest MRT or LRT station."
+ggplot() +
+  geom_spatraster(
+    data = trave_time_singapore
+  ) +
+  scale_fill_viridis_c(
+    option = "A",
+    direction = -1,
+    na.value = "transparent"
+  ) +
+  theme_minimal() +
+  labs(fill = "Minutes") +
+  geom_spatvector(
+    data = singapore_shapefile,
+    fill = "transparent",
+    col = "grey20"
   )
@@
 
-[^1]: Land Transport Authority. (2019). LTA MRT Station Exit (GEOJSON) (2024)
+[^1]: D.
+J.
+Weiss, A.
+Nelson, C.
+A.
+Vargas-Ruiz, K.
+Gligoric, S., Bavadekar, E.
+Gabrilovich, A.
+Bertozzi-Villa, J.
+Rozier, H.
+S.
+Gibson, T., Shekel, C.
+Kamath, A.
+Lieber, K.
+Schulman, Y.
+Shao, V.
+Qarkaxhija, A.
+K.
+Nandi, S.
+H.
+Keddie, S.
+Rumisha, P.
+Amratia, R.
+Arambepola, E.
+G.
+Chestnutt, J.
+J.
+Millar, T.
+L.
+Symons, E.
+Cameron, K.
+E.
+Battle, S.
+Bhatt, and P.
+W.
+Gething.
+Global maps of travel time to healthcare facilities.
+(2020) Nature Medicine.
+<https://doi.org/10.1038/s41591-020-1059-1>
+
+[^2]: Land Transport Authority. (2019). LTA MRT Station Exit (GEOJSON) (2024)
 [Dataset]. data.gov.sg. Retrieved December 10, 2024 from
-https://data.gov.sg/datasets/d_b39d3a0871985372d7e1637193335da5/view
+<https://data.gov.sg/datasets/d_b39d3a0871985372d7e1637193335da5/view>.

### 2025-03-03 README.Rmd
typos and missing footnote in vignette
add contributor code of conduct *not knitted to md yet because bloody aws behind malariaAtlas is down again*

@@
 [![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
+[![Contributor Covenant](https://img.shields.io/badge/Contributor%20Covenant-2.1-4baaaa.svg)](code_of_conduct.md)
 <!-- badges: end -->
@@
 
-We will use the friction surface from Weiss et al (2020) that can be downloaded by `traveltime` with the function `get_friction_surface()`.
+We will use the friction surface from Weiss et al.
+(2020) that can be downloaded by `traveltime` with the function `get_friction_surface()`.
 This function takes extents in a variety of formats and returns the surface for that extent only.
@@
 
+## Code of Conduct
+
+Please note that the `traveltime` project is released with a [Contributor Code of Conduct](https://idem-lab.github.io/traveltime/CODE_OF_CONDUCT.html).
+By contributing to this project, you agree to abide by its terms.
+
 [^1]: D.

### 2025-03-05 R/utils.R
typo

@@ -1,6 +1,6 @@
 
 # warn_and_return checks if a file exists already and returns the raster
-# from the exsting file with a warning that no calculation is being performed
+# from the existing file with a warning that no calculation is being performed
 warn_and_return <- function(
   filename,
   overwrite

### 2025-03-05 tests/testthat/test-get_friction_surface.R
upgrade testing infrastructure for errors

@@ -25,3 +25,15 @@ test_that("Are SpatRasters", {
   expect_s4_class(wfs, "SpatRaster")
 
 })
+
+
+test_that(
+  desc = "throws error",
+  code = {
+    expect_error(
+      get_friction_surface(
+        extent = NULL
+      )
+    )
+  }
+)

### 2025-03-05 tests/testthat/test-calculate_travel_time.R
more testing stuff

@@ -36,4 +36,46 @@ test_that("Is SpatRaster", {
   )
 
   expect_s4_class(tt2, "SpatRaster")
+
+
+  # check throws error if points is junk
+  expect_error(
+    calculate_travel_time(
+      friction_surface = wfs,
+      points = 1
+    )
+  )
+
+  # check parses points as SpatVector OK
+  from_here_SpatVector <- terra::vect(
+    x = from_here,
+    geom = c(
+      "x",
+      "y"
+    )
+  )
+
+  tt3 <- calculate_travel_time(
+    friction_surface = wfs,
+    points = from_here_SpatVector,
+    filename = tempfile(
+      fileext = ".tif"
+    )
+  )
+
+  expect_s4_class(tt3, "SpatRaster")
+
+
+
 })
+
+test_that(
+  desc = "aborts if friction surface is not a SpatRaster",
+  code = {
+    expect_error(
+      calculate_travel_time(
+        friction_surface = TRUE
+      )
+    )
+  }
+)

### 2025-03-05 tests/testthat/test-ext_matrix.R

@@ -1,0 +1,20 @@
+test_that(
+  desc = "takes SpatExtent",
+  code = {
+    x <- c(111, 112, 0, 1)
+
+    y <-  terra::ext(x)
+
+    z <- ext_matrix(y)
+
+    expect_type(
+      object = z,
+      type = "double"
+    )
+
+    expect_equal(
+      object = dim(z),
+      expected = c(2, 2)
+    )
+  }
+)

### 2025-03-05 tests/testthat/test-utils.R

@@ -1,3 +1,44 @@
-test_that("multiplication works", {
-  expect_equal(2 * 2, 4)
-})
+test_that(
+  desc = "Warn and return throws error",
+  code = {
+
+    temptif <- tempfile(
+      fileext = ".tif"
+    )
+
+    temprast <- terra::rast()
+
+    temprast[] <- 1
+
+    terra::writeRaster(
+      x = temprast,
+      filename = temptif
+    )
+
+    temprast <- terra::rast(
+      x = temptif
+    )
+
+
+    expect_warning(
+      warn_and_return(
+        filename = temptif,
+        overwrite = FALSE
+      )
+    )
+
+    # returned_rast <- warn_and_return(
+    #   filename = temptif,
+    #   overwrite = FALSE
+    # )
+#
+#     expect_s4_class(
+#       object = warn_and_return(
+#         filename = temptif,
+#         overwrite = FALSE
+#       ),
+#       class = "SpatRaster"
+#     )
+
+  }
+)

### 2025-03-05 tests/testthat/test-ext_matrix.R
100% test coverage mfers

@@ -18,3 +18,84 @@ test_that(
     )
   }
 )
+
+test_that(
+  desc = "ext_matrix fails if too long",
+  code = {
+
+    x <- c(1,2:10)
+
+    expect_error(
+      object = ext_matrix(x)
+    )
+  }
+)
+
+test_that(
+  desc = "ext_matrix fails if double too short",
+  code = {
+
+    x <- 1
+
+    expect_error(
+      object = ext_matrix(x)
+    )
+  }
+)
+
+test_that(
+  desc = "ext_matrix fails if integer incorrect length",
+  code = {
+
+    x <- c(1,2)
+
+    expect_error(
+      object = ext_matrix(x)
+    )
+  }
+)
+
+
+
+test_that(
+  desc = "ext_matrix fails if dim incorrect",
+  code = {
+
+    x <- matrix(1)
+
+    expect_error(
+      object = ext_matrix(x)
+    )
+  }
+)
+
+
+test_that(
+  desc = "ext_matrix fails if inappropriate class",
+  code = {
+
+    x <- "junk"
+
+    expect_error(
+      object = ext_matrix(x)
+    )
+  }
+)
+
+test_that(
+  desc = "ext_matrix parses double",
+  code = {
+
+    x <- c(111, 112, 0, 1)
+
+    y <- ext_matrix(extent = x)
+
+    expect_true(
+      object = inherits(
+        x = y,
+        what = "matrix"
+      )
+    )
+  }
+)
+

### 2025-03-11 paper/paper.qmd

--- 2025-02-12 paper/paper.qmd
@@
 format:
-  #md: default
-  html:
+  # md: default
+  wordcount-html:
+  # html:
     keep-md: true
@@
 Here we present a software package --- `traveltime` --- written in and for the language R [@Rref].
-`traveltime` enables a user to create a map of the motorised or walking travel time over an area of interest from a user-specified set of geographic coordinates.
-The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations.
-We envisage this software having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias; @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], allocating public defibrillators [@tierney2018novel], setting rehabilitation districts for stroke patients [@padgham2019introduction], or understanding access to agricultural processing facilities [@zhao2023replanting].
+`traveltime` enables a user to create a raster map of the travel time over an area of interest from a user-specified set of geographic coordinates.
+Applications include: estimating sampling bias [@dennis2000bias], allocating defibrillators [@tierney2018novel], setting health districts [@padgham2019introduction], or mapping access to vehicle chargers [@falchetta2021electric] and agricultural facilities [@zhao2023replanting].
 
-The work-flow requires two steps:
+The work-flow requires two key steps:
 
--   preparing a 'friction surface' for the area of interest, and then
+-   preparing a friction surface for the area of interest, and then
 -   calculating travel time over that surface for the points of interest.
@@
 `traveltime` provides a spatial interface using object classes from the `terra` package [@terra].
-It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` class object; and returns the result as a `SpatRaster` class object.
-The travel time is calculated as movement over a resistance 'friction surface' [@gdistance2017].
-To provide easy access to the existing friction surfaces generated by @weiss2020global, `traveltime` uses the package `malariaAtlas` to allow users to access friction surfaces for the area of interest; though users can also supply any other friction surfaces to `traveltime`.
-Functionality from `gdistance` [@gdistance2017] is also used internally to calculate the minimum least-cost-distance for each cell from the points of interest.
+Travel time is calculated as movement over a resistance 'friction surface' [@gdistance2017].
+To provide easy access to the existing friction surfaces generated by @weiss2020global, `traveltime` internally uses `malariaAtlas` [@pfeffer2018malariaatlas] to download surfaces for the area of interest; though users can also supply any other friction surface.
 
-`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and has reference documents at <https://idem-lab.github.io/traveltime/>.
-Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying methodological work [@weiss2018global; @weiss2020global] as well.
+`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and documented at <https://idem-lab.github.io/traveltime/>.
+Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package are accompanied by citing the underlying methodological work as well [@weiss2018global; @weiss2020global].
 
@@
 
-Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] have generated significant interest and use[^1], and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata].
-There is clear demand for these type of products.
+Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care [@hulland2019travel; @weiss2020global] have generated much interest and use[^1], and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata].
+Here we extend that work to enable travel time calculations from any arbitrary set of locations and friction surface.
 
@@
 
-@weiss2020global made their code available as an R script to allow for reproduction and extension of their analyses (<https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt> ).
-To further enable extension of this work, here we have developed an R package based on that code to seamlessly calculate the travel time from any arbitrary set of locations.
+A gaggle of R packages provide superficially similar though fundamentally different functionality via the [TravelTime.com](https://www.TravelTime.com) API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh].
+Their 'Isochron' polygons --- areas reachable within a given time from a given location --- are most comparable to what `traveltime::calculate_travel_time()` calculates.
+However, each isochron is a single polygon calculated is for a single point location and specified maximum travel time, rather than a raster of temporal gradation across a landscape, jointly for an arbitrary number of points, as in `traveltime`.
 
-Other R packages provide superficially similar though fundamentally different functionality.
-A gaggle of R packages provide interfaces to the [TravelTime.com](https://www.TravelTime.com) API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh].
-The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons --- areas reachable within a given time from a given location.
-The isochron polygons are most comparable to what `traveltime::calculate_travel_time()` calculates, though each isochron polygon is a single polygon calculated is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than a raster of gradation across a landscape for an arbitrary number of points as in `traveltime`.
-TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, plus additional calculation of the minimum value for each cell from all points.
-Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use of this resource.
-
-With `traveltime`, we provide free and open source software to estimate travel time from any number of user-supplied locations, across a complete area of interest, and with convenient access to motorised or walking friction surfaces with global coverage.
+With `traveltime`, we provide free and open source software to estimate travel time from any number of user-supplied locations, across a complete area of interest, and with convenient access to motorised transport or walking friction surfaces with global coverage.
 
@@
 
-In this example we will calculate the walking travel time from the nearest mass transit station across the island nation of Singapore --- specifically Mass Rapid Transit (MRT) and Light Rail Transit (LRT) stations --- and create a map of this.
+Here we provide a very brief example to calculate and map the walking travel time from rail transport stations across Singapore.
+[Complete code for this example is available as a vignettte in package documentation](https://idem-lab.github.io/traveltime/articles/singapore.html).
 
-## Prepare the data and friction surface
+## Prepare data and friction surface
 
-For this exercise, we need two items of data:
+We need two items of data:
 
--   our area of interest --- in this case a map of Singapore, and
--   our points to calculate travel time from --- here the locations of Singapore's MRT and LRT stations.
+-   our area of interest --- Singapore, and
+-   our points to calculate travel time from --- Singapore's Mass Rapid Transit (MRT) and Light Rail Transit (LRT) stations.
 
-We can download a national-level polygon boundary of Singapore from the GADM [@gadm] database using the `geodata` package [@geodata].
-Here we download only the national boundary (`level = 0`) and at a low resolution (`resolution = 2`).
-Our boundary `singapore_shapefile` is a `SpatVector` class object.
+We download `singapore_shapefile`, a polygon boundary of Singapore from the GADM [@gadm] database using the `geodata` package [@geodata] as our area of interest.
+Our points are the `traveltime::stations` data, containing coordinates of all LRT and MRT station exits in Singapore [@singdata].
+We then download a walking travel friction surface for Singapore:
 
 ```{r}
+#| echo: false
+#| output: false
 #| label: get-basemap
@@
 
-The `stations` data set included in the `traveltime` package is a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from @singdata:
-
 ```{r}
+#| echo: false
+#| output: false
 #| label: show-stations
@@
 
-Now that we have the two items of data that we require, the next step is to prepare a friction surface for our area of interest.
-We will use the friction surface from @weiss2020global that can be downloaded by `traveltime` with the function `get_friction_surface()`.
-We can pass in our basemap `singapore_shapefile`, a `SpatVector`, directly as the `extent`.
-We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`.
-(Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we then mask out areas outside of the land boundary in `singapore_shapefile`:
-
 ```{r}
@@
 #| output: false
-friction_singapore <- get_friction_surface(
+friction_singapore <- traveltime::get_friction_surface(
     surface = "walk2020",
@@
   )|> 
-  mask(singapore_shapefile)
+  terra::mask(singapore_shapefile)
 ```
 
-Thus we have our friction surface as a `SpatRaster`:
+(Alternatively, we could use `surface = "motor2020"` for motorised travel. We're also only interested in walking *on land* so we `mask` out areas outside of the land boundary in `singapore_shapefile`)
 
 ```{r }
+#| echo: false
+#| output: false
 #| label: show-friction-surface-raster
@@
 
-Below we plot the friction surface raster `friction_singapore`, with the vector boundary `singapore_shapefile` as a grey line, and `stations` as grey points (Figure \ref{fig-data}).
-`traveltime` takes resistance values of friction [@gdistance2017], so higher values of friction indicate more time travelling across a given cell.
-In this case walking friction is fairly uniform across the islands barring a few small inaccesible locations.
-More heterogeneity would be expected in less developed locations with fewer transport corridors.
+We plot these data below. `traveltime` takes resistance values of friction [@gdistance2017], so higher values of friction indicate more time travelling across a given cell.
 
@@
 
-We present the resulting calculated travel times in Figure \ref{fig-result} where, as expected, the travel times are lowest near station exits (per Figure \ref{fig-data}) and progressively higher further away.
-Note that the results in `trave_time_singapore` include infinite values (`Inf`  above).
-In Figure \ref{fig-data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
-The raster cells for these islands appear absent in Figure \ref{fig-result}.
-Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig-result}.
+Resulting travel times are Figure \ref{fig-result}[^2].
 
@@
 
+[^2]: In Figure \ref{fig-data}, the islands to the south and north-east are filled, but unconnected with the mainland.
+These cells appear absent in Figure \ref{fig-result}; because they are not connected to any cells with a station, the calculated travel time is infinite, and so are not plotted in Figure \ref{fig-result}.
+
 # Opportunities for future development
 
-The `traveltime` package is immediately suitable to a range of applications where travel to custom locations is of interest.
-Nonetheless, we see opportunities to build the package utility into the future through two mechanisms: (1) capability to better distribute a wider range friction surfaces, and (2) additional methods to handle large spatial extents.
+The `traveltime` package is immediately suitable to a range of application. Nonetheless, we see opportunities to build the package utility through:
 
-Firstly, `traveltime` currently has access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution[^2].
-Although the user can supply their own friction surface, we expect most applications will use these existing surfaces given the extensive work needed in creating new ones [@weiss2018global; @weiss2020global].
-As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available, though this is likely to occur first through `malariaAtlas`.
-Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses.
-For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km^2^ cell size is a relatively large area to walk across, and thus actual waking times may vary significantly within each cell.
-We underline however that a user can provide their own higher resolution friction surface at present if desired.
-
-[^2]: Approximately 0.008333 decimal degrees, or just below 1 km^2^ at the equator
-
-At the other end of the scale, the calculations can require relatively large amounts of onboard memory for analyses over large landscapes (e.g.
-analyses over Africa required \~ 72 GB RAM).
-Developing methods to handle large landscapes either with less memory or via cloud resources would be helpful to make such analyses accessible to those without access to larger computing resources.
+  - capability to better distribute a wider range friction surfaces, and
+  - additional methods to efficiently compute results over large spatial extents.
 

### 2025-04-07 paper/paper.qmd
switch to quarto render with line numbers

@@
   # md: default
-  wordcount-html:
+  # wordcount-html:
   # html:
+  #   keep-md: true
+  #   fig-height: 4
+  #   fig-align: center
+  #   fig-format: png
+  #   dpi: 300
+  pdf:
     keep-md: true
@@
     fig-align: center
+    dpi: 300
     fig-format: png
-    dpi: 300
+header-includes:
+  \usepackage{lineno} \linenumbers
+  \usepackage{hanging}
 tags: 
@@
    index: 4
-date: In prep 2025
+date: 2025-04-07
 bibliography: paper.bib

### 2025-04-09 paper/paper.qmd
flesh out again, render to docx

@@
   #   dpi: 300
-  pdf:
+  docx:
     keep-md: true
@@
 Here we present a software package --- `traveltime` --- written in and for the language R [@Rref].
-`traveltime` enables a user to create a raster map of the travel time over an area of interest from a user-specified set of geographic coordinates.
-Applications include: estimating sampling bias [@dennis2000bias], allocating defibrillators [@tierney2018novel], setting health districts [@padgham2019introduction], or mapping access to vehicle chargers [@falchetta2021electric] and agricultural facilities [@zhao2023replanting].
+`traveltime` enables a user to create a raster map of the travel time over an area of interest from a user-specified set of locations (geographic coordinates).
+The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to the nearest of the supplied locations.
+We envisage this software having diverse applications including: estimating sampling bias [@dennis2000bias], allocating defibrillators [@tierney2018novel], setting health districts [@padgham2019introduction], or mapping access to vehicle chargers [@falchetta2021electric] and agricultural facilities [@zhao2023replanting].
 
@@
 
-# Statement of need
+# Context
 
@@
 However, each isochron is a single polygon calculated is for a single point location and specified maximum travel time, rather than a raster of temporal gradation across a landscape, jointly for an arbitrary number of points, as in `traveltime`.
+TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, plus additional calculation of the minimum value for each cell from all points.
+Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use of this resource.
 
@@
 
-Resulting travel times are Figure \ref{fig-result}[^2].
+We present the resulting calculated travel times in Figure \ref{fig-result} where, as expected, the travel times are lowest near station exits (per Figure \ref{fig-data}) and progressively higher further away.
+Note that the results in `trave_time_singapore` include infinite values (`Inf` above).
+In Figure \ref{fig-data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
+The raster cells for these islands appear absent in Figure \ref{fig-result}.
+Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig-result}.
 
@@
 
-[^2]: In Figure \ref{fig-data}, the islands to the south and north-east are filled, but unconnected with the mainland.
-These cells appear absent in Figure \ref{fig-result}; because they are not connected to any cells with a station, the calculated travel time is infinite, and so are not plotted in Figure \ref{fig-result}.
-
 # Opportunities for future development
@@
   - additional methods to efficiently compute results over large spatial extents.
+
+Firstly, `traveltime` currently facilitates access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution[^2].
+Although the user can presently supply their own friction surface, we expect most applications will use these existing surfaces given the extensive work needed in creating new ones [@weiss2018global; @weiss2020global].
+As landscapes are dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available.
+Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses.
+For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km^2^ cell size is a relatively large area to walk across, and thus actual waking times may vary significantly within each cell.
+We underline however that a user can provide their own higher resolution friction surface at present if desired.
+
+[^2]: Approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator
 

### 2025-04-09 paper/paper.qmd
render unto docx that which is docxs's

@@
   #   dpi: 300
+  #pdf:
   docx:
@@
     fig-format: png
-header-includes:
-  \usepackage{lineno} \linenumbers
-  \usepackage{hanging}
+# header-includes:
+#   \usepackage{lineno} \linenumbers
+#   \usepackage{hanging}
 tags: 
@@
   - spatial analysis
-authors:
+author:
   - name: Gerard E. Ryan

### 2025-04-09 README.Rmd
RENDER
Include Singapore open data licence acknowledgement

@@
 
-# traveltime
+# `traveltime`
 
@@
 
+## Imported data
+This repository contains information from the dataset "LTA MRT Station Exit (GEOJSON)" accessed on the 10th of December 2024 from  [data.gov.sg](https://data.gov.sg/datasets/d_b39d3a0871985372d7e1637193335da5/view), which is made available under the terms of the Singapore Open Data Licence version 1.0 <https://data.gov.sg/open-data-licence>
+
 [^1]: D.

### 2025-05-02 paper/paper.qmd
Include Singapore open data licence acknowledgement
replace latex fig refs with rmardown ones AGAIN NB: this adds in the word Figure to the numeric reference so this word is deleted again. Bah

@@
 
-We present the resulting calculated travel times in Figure \ref{fig-result} where, as expected, the travel times are lowest near station exits (per Figure \ref{fig-data}) and progressively higher further away.
+We present the resulting calculated travel times in @fig-result where, as expected, the travel times are lowest near station exits (per @fig-data) and progressively higher further away.
 Note that the results in `trave_time_singapore` include infinite values (`Inf` above).
-In Figure \ref{fig-data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
-The raster cells for these islands appear absent in Figure \ref{fig-result}.
-Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig-result}.
+In @fig-data, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
+The raster cells for these islands appear absent in @fig-result.
+Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in @fig-result.
 
@@
 
+The package associated with this paper contains information from the dataset "LTA MRT Station Exit (GEOJSON)" accessed on the 10th of December 2024 from  data.gov.sg, which is made available under the terms of the Singapore Open Data Licence version 1.0 https://data.gov.sg/open-data-licence.
+
 # References

### 2025-05-02 paper/paper.qmd
update existing material into verixiv strucutre

@@
    index: 4
-date: 2025-04-07
+date: 2025-05-02
 bibliography: paper.bib
@@
 
-# Summary
+# Abstract
+
+Understanding and mapping the time to travel among locations is useful for many activities from urban planning to public health and myriad others.
+Here we present a software package --- `traveltime` --- written in and for the language R.
+`traveltime` enables a user to create a raster map of the travel time over an area of interest from a user-specified set of locations defined by geographic coordinates.
+The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to the nearest of the supplied locations.
+We envisage this software having diverse applications including: estimating sampling bias, allocating defibrillators, setting health districts, or mapping access to vehicle chargers and agricultural facilities.
+The work-flow requires two key steps: preparing a friction surface for the area of interest, and then calculating travel time over that surface for the points of interest.
+`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and documented at <https://idem-lab.github.io/traveltime/>.
+
+# Introduction
 
 Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite].
-Here we present a software package --- `traveltime` --- written in and for the language R [@Rref].
-`traveltime` enables a user to create a raster map of the travel time over an area of interest from a user-specified set of locations (geographic coordinates).
-The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to the nearest of the supplied locations.
-We envisage this software having diverse applications including: estimating sampling bias [@dennis2000bias], allocating defibrillators [@tierney2018novel], setting health districts [@padgham2019introduction], or mapping access to vehicle chargers [@falchetta2021electric] and agricultural facilities [@zhao2023replanting].
-
-The work-flow requires two key steps:
-
--   preparing a friction surface for the area of interest, and then
--   calculating travel time over that surface for the points of interest.
-
-`traveltime` provides a spatial interface using object classes from the `terra` package [@terra].
-Travel time is calculated as movement over a resistance 'friction surface' [@gdistance2017].
-To provide easy access to the existing friction surfaces generated by @weiss2020global, `traveltime` internally uses `malariaAtlas` [@pfeffer2018malariaatlas] to download surfaces for the area of interest; though users can also supply any other friction surface.
-
-`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and documented at <https://idem-lab.github.io/traveltime/>.
-Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package are accompanied by citing the underlying methodological work as well [@weiss2018global; @weiss2020global].
-
-# Context
-
 Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care [@hulland2019travel; @weiss2020global] have generated much interest and use[^1], and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata].
@@
 
-[^1]: Collectively >1600 citations per Google Scholar at the 28th of January 2025.
+We present a software package --- `traveltime` --- written in and for the language R [@Rref].
+`traveltime` enables a user to create a raster map of the travel time over an area of interest from a user-specified set of locations defined by geographic coordinates.
+The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to the nearest of the supplied locations.
 
@@
 
-# Example: walking from public transport in Singapore
+[^1]: Collectively >1600 citations per Google Scholar at the 28th of January 2025.
 
-Here we provide a very brief example to calculate and map the walking travel time from rail transport stations across Singapore.
+# Methods
+
+## Implementation
+
+`traveltime` is an R [@Rref] package and requires installation of R version 4.1 or a more recent version `traveltime` provides a spatial interface using object classes from the `terra` package [@terra].
+Travel time is calculated as movement over a resistance 'friction surface' [@gdistance2017].
+To provide easy access to the existing friction surfaces generated by @weiss2020global, `traveltime` internally uses the R package `malariaAtlas` [@pfeffer2018malariaatlas] to download surfaces for the area of interest; though users can also supply any other friction surface raster.
+
+## Operation
+
+The work-flow requires two key steps:
+
+-   preparing a friction surface for the area of interest, and then
+-   calculating travel time over that surface for the points of interest.
+
+### Installation
+
+`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and documented at <https://idem-lab.github.io/traveltime/>.
+It can be installed in R as follows:
+
+``` r
+install.packages("traveltime", repos = c("https://idem-lab.r-universe.dev"))
+```
+
+### Example: walking from public transport in Singapore
+
+Here we provide an example to calculate and map the walking travel time from rail transport stations across Singapore.
 [Complete code for this example is available as a vignettte in package documentation](https://idem-lab.github.io/traveltime/articles/singapore.html).
 
-## Prepare data and friction surface
+#### Prepare data and friction surface
 
@@
 
-We download `singapore_shapefile`, a polygon boundary of Singapore from the GADM [@gadm] database using the `geodata` package [@geodata] as our area of interest.
-Our points are the `traveltime::stations` data, containing coordinates of all LRT and MRT station exits in Singapore [@singdata].
-We then download a walking travel friction surface for Singapore:
+We download `singapore_shapefile`, a polygon boundary of Singapore from the GADM [@gadm] database using the `geodata` package [@geodata] as our area of interest:
 
 ```{r}
-#| echo: false
-#| output: false
+#| echo: true
+#| output: true
 #| label: get-basemap
@@
 
-```{r}
-#| echo: false
-#| output: false
-#| label: show-stations
-library(traveltime)
-head(stations)
-```
+Next we use the function `traveltime::get_friction_surface` to retrieve a walking friction surface for our area of interest.
+Alternatively, we could use `surface = "motor2020"` for motorised travel.
+We're also only interested in walking *on land* so we `mask` out areas outside of the land boundary in `singapore_shapefile`.
+Users supplying their own friction surfaces do not need to download one in this fashion, only ensure that it is in `SpatRaster` format.
 
@@
 #| echo: true
-#| output: false
+#| output: true
 friction_singapore <- traveltime::get_friction_surface(
@@
   terra::mask(singapore_shapefile)
+
+friction_singapore
 ```
 
-(Alternatively, we could use `surface = "motor2020"` for motorised travel. We're also only interested in walking *on land* so we `mask` out areas outside of the land boundary in `singapore_shapefile`)
+Our points are the `traveltime::stations` data, containing coordinates of all LRT and MRT station exits in Singapore [@singdata]:
 
-```{r }
-#| echo: false
-#| output: false
-#| label: show-friction-surface-raster
-friction_singapore
+```{r}
+#| echo: true
+#| output: true
+#| label: show-stations
+library(traveltime)
+head(stations)
 ```
@@
 
-## Calculate and plot the travel time
+#### Calculate and plot the travel time
 
@@
 
-# Opportunities for future development
+# Discussion
 
-The `traveltime` package is immediately suitable to a range of application. Nonetheless, we see opportunities to build the package utility through:
+The `traveltime` package is immediately suitable to be used 'out-of-the-box' with any set of coordinates, in any part of the globe.
+We envisage this software having diverse applications including: estimating sampling bias [@dennis2000bias], allocating defibrillators [@tierney2018novel], setting health districts [@padgham2019introduction], or mapping access to vehicle chargers [@falchetta2021electric] and agricultural facilities [@zhao2023replanting].
+Nonetheless, we see opportunities to build the package utility through:
 
@@
 
-Firstly, `traveltime` currently facilitates access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution[^2].
+Firstly, `traveltime` currently facilitates access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator).
 Although the user can presently supply their own friction surface, we expect most applications will use these existing surfaces given the extensive work needed in creating new ones [@weiss2018global; @weiss2020global].
@@
 
-[^2]: Approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator
+At the other end of the scale, as the area of interest increases, the size of the matrix of calculations necessary increases exponentially, making significant memory demands for analyses over large landscapes (e.g.
+analyses over Africa required \~ 72 GB RAM to run successfully).
+Developing methods to handle large landscapes either with less memory or via cloud resources would be helpful to make such analyses accessible to those without access to larger computing resources.
+
+Lastly, although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package are accompanied by citing the underlying methodological work as well [@weiss2018global; @weiss2020global].
 

### 2025-05-02 paper/paper.qmd
add / reformat section headers for verixiv
still need to get images spat out at better resolution this is rubbish.

over this nonsense.

@@
     keep-md: true
-    fig-height: 4
+    fig-height: 6
     fig-align: center
-    dpi: 300
-    fig-format: png
+    dpi: 600
+    fig-format: jpeg
 # header-includes:
@@
 
-# Acknowledgements
+# Competing interests
+
+No competing interests were disclosed.
+
+# Grant Information
 
@@
 
+# Figure Permissions
+
+The authors confirm ownership of the figures used in this manuscript.
+
+# Software availability
+
+All software described here is available from R-Universe <https://idem-lab.r-universe.dev/traveltime> and GitHub <https://github.com/idem-lab/traveltime>, and documented at <https://idem-lab.github.io/traveltime/>.
+
 The package associated with this paper contains information from the dataset "LTA MRT Station Exit (GEOJSON)" accessed on the 10th of December 2024 from  data.gov.sg, which is made available under the terms of the Singapore Open Data Licence version 1.0 https://data.gov.sg/open-data-licence.

### 2025-05-05 paper/paper.qmd
use author block <https://github.com/kapsner/authors-block> extension to add author addresses and corresponding author to header in docx

@@
     fig-align: center
-    dpi: 600
-    fig-format: jpeg
+    dpi: 300
+    fig-format: png
 # header-includes:
@@
     corresponding: true
-    affiliation: "1, 2"
+    email: gerry.ryan@thekids.org.au
+    #affiliation: "1, 2"
+    affiliations:
+      - ref: TKRIA
+      - ref: MSPGH
   - name: Nicholas Tierney
     orcid: 0000-0003-1460-8722
-    affiliation: "1, 3"
+    #affiliation: "1, 3"
+    affiliations:
+      - ref: TKRIA
+      - ref: CU
   - name: Nick Golding
     orcid: 0000-0001-8916-5570
-    affiliation: "1, 4"
+    #affiliation: "1, 4"
+    affiliations:
+      - ref: TKRIA
+      - ref: UWA
   - name: Daniel J. Weiss
     orcid: 0000-0002-6175-5648
-    affiliation: "1, 3"
+    #affiliation: "1, 3"
+    affiliations:
+      - ref: TKRIA
+      - ref: CU
 affiliations:
  - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
+   id: TKRIA
    index: 1
@@
    index: 2
+   id: MSPGH
  - name: Curtin University, Bentley, WA, Australia
    index: 3
+   id: CU
  - name: University of Western Australia, WA, Australia
    index: 4
+   id: UWA
 date: 2025-05-02
 bibliography: paper.bib
+filters:
+  - authors-block
 ---

### 2025-05-19 README.Rmd
fix typo `trave_time_singapore` in README, paper, and vignette Also update acknowledgements and endmatter order per VeriXiv

@@
 #| echo: true
-trave_time_singapore <- calculate_travel_time(
+travel_time_singapore <- calculate_travel_time(
   friction_surface = friction_singapore,
@@
 
-trave_time_singapore
+travel_time_singapore
 ```
@@
 We present the resulting calculated travel times below where (as expected) the travel times are lowest near station exits and progressively higher further away.
-Note that the results in `trave_time_singapore` include infinite values (`Inf`  above).
+Note that the results in `travel_time_singapore` include infinite values (`Inf`  above).
 The islands to the south and north-east are shown as filled cells in the figure above, i.e., they are not masked out by `singapore_shapefile`.
@@
   geom_spatraster(
-    data = trave_time_singapore
+    data = travel_time_singapore
   ) +

### 2025-05-19 paper/paper.qmd
add code doi from zenodo to paper
fix typo `trave_time_singapore` in README, paper, and vignette Also update acknowledgements and endmatter order per VeriXiv

@@
 #| label: calculate-result
-trave_time_singapore <- calculate_travel_time(
+travel_time_singapore <- calculate_travel_time(
   friction_surface = friction_singapore,
@@
 We present the resulting calculated travel times in @fig-result where, as expected, the travel times are lowest near station exits (per @fig-data) and progressively higher further away.
-Note that the results in `trave_time_singapore` include infinite values (`Inf` above).
+Note that the results in `travel_time_singapore` include infinite values (`Inf` above).
 In @fig-data, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland.
@@
 
+# Figure Permissions
+
+The authors confirm ownership of the figures used in this manuscript.
+
+# Data and Software availability
+
+All software described here is available from R-Universe <https://idem-lab.r-universe.dev/traveltime> and GitHub <https://github.com/idem-lab/traveltime>, and documented at <https://idem-lab.github.io/traveltime/>.
+Code used in this paper is released via GitHub as [v4](https://github.com/idem-lab/traveltime/releases/tag/v3), and archived on Zenodo under [DOI 10.5281/zenodo.15347016](https://doi.org/10.5281/zenodo.15347016).
+
+The package associated with this paper contains information from the dataset "LTA MRT Station Exit (GEOJSON)" accessed on the 10th of December 2024 from  data.gov.sg, which is made available under the terms of the Singapore Open Data Licence version 1.0 https://data.gov.sg/open-data-licence.
+
 # Competing interests
@@
 No competing interests were disclosed.
+
+# Acknowledgements
+
+An earlier version of this article has been archived on EarthArXiv <https://doi.org/10.31223/X56M74>
 
@@
 
-# Figure Permissions
-
-The authors confirm ownership of the figures used in this manuscript.
-
-# Software availability
-
-All software described here is available from R-Universe <https://idem-lab.r-universe.dev/traveltime> and GitHub <https://github.com/idem-lab/traveltime>, and documented at <https://idem-lab.github.io/traveltime/>.
-
-The package associated with this paper contains information from the dataset "LTA MRT Station Exit (GEOJSON)" accessed on the 10th of December 2024 from  data.gov.sg, which is made available under the terms of the Singapore Open Data Licence version 1.0 https://data.gov.sg/open-data-licence.
-
 # References

### 2025-05-22 README.Rmd
update paper link to VeriXiv preprint

@@
 
-Ryan, G.E., Tierney, N., Golding, N., and Weiss, D.J (2025).
-[traveltime: an R package to calculate travel time across a landscape from user-specified locations](https://doi.org/10.31223/X56M74){target="_blank"}.
-**EarthArXiv** 10.31223/X56M74
+**Ryan, G.E.**, Tierney, N., Golding, N., and Weiss, D.J (2025).
+[traveltime: an R package to calculate travel time across a landscape from user-specified locations](https://doi.org/10.12688/verixiv.989.1){target="_blank"}.
+**VeriXiv** 2:99 DOI 10.12688/verixiv.989.1
 

### 2025-07-24 README.Rmd
update ref to published version

--- 2025-05-19 README.Rmd
@@
 Ryan, G.E., Tierney, N., Golding, N., and Weiss, D.J (2025).
-[traveltime: an R package to calculate travel time across a landscape from user-specified locations](https://doi.org/10.31223/X56M74){target="_blank"}.
-**EarthArXiv** 10.31223/X56M74
+[traveltime: an R package to calculate travel time across a landscape from user-specified locations](https://doi.org/10.12688/gatesopenres.16356.1){target="_blank"}.
+Gates Open Research 9:50
 

### 2025-08-10 README.Rmd
gh dl monitor

@@
 [![Contributor Covenant](https://img.shields.io/badge/Contributor%20Covenant-2.1-4baaaa.svg)](code_of_conduct.md)
+![GitHub Downloads (all assets, latest release)](https://img.shields.io/github/downloads/idem-lab/traveltime/latest/total)
 <!-- badges: end -->
